### 2021-05-27 article/manuscript.tex
Writing.

\RequirePackage[l2tabu,orthodox]{nag} % turn on warnings because of bad style
\documentclass[a4paper,11pt,bibtotoc]{scrartcl}
\usepackage{natbib}

\usepackage[utf8]{inputenc}

\usepackage[T1]{fontenc}        % Tries to use Postscript Type 1 Fonts for better rendering
\usepackage{lmodern}            % Provides the Latin Modern Font which offers more glyphs than the default Computer Modern
\usepackage{amsmath} % Provides all mathematical commands
\usepackage{amsfonts} % Provides all mathematical commands

\usepackage{hyperref}           % Provides clickable links in the PDF-document for \ref
\usepackage{grffile}            % Allow you to include images (like graphicx). Usage: \includegraphics{path/to/file}

% Allows to set units
\usepackage[ugly]{units}        % Allows you to type units with correct spacing and font style. Usage: $\unit[100]{m}$ or $\unitfrac[100]{m}{s}$

% Additional packages
\usepackage{url}                % Lets you typeset urls. Usage: \url{http://...}
\usepackage{breakurl}           % Enables linebreaks for urls
\usepackage{xspace}             % Use \xpsace in macros to automatically insert space based on context. Usage: \newcommand{\es}{ESPResSo\xspace}
\usepackage{xcolor}             % Obviously colors. Usage: \color{red} Red text
\usepackage{booktabs}           % Nice rules for tables. Usage \begin{tabular}\toprule ... \midrule ... \bottomrule

% Source code listings
\usepackage{listings}           % Source Code Listings. Usage: \begin{lstlisting}...\end{lstlisting}
\lstloadlanguages{python}       % Default highlighting set to "python"

\begin{document}

\title{GPROF-NN: A neural-network based implementation of the Goddard Profiling Algorithm}
\date{\today}

\maketitle

\section{Introduction}
\section{Methods and data}

This section describes the original GPROF algorithm as well as the two newly
developed neural-network-based versions.

\subsection{The original GPROF algorithm}

The original GPROF algorithm, here referred to simply as GPROF, is based on a
Bayesian formulation of the retrieval problem. Given an observation
$\mathbf{y}$, the result of the retrieval is defined as the posterior
distribution $P(\mathbf{x} | \mathbf{y})$ over the values $\mathbf{x}$ of a
numerical representation of the atmospheric state.

 The posterior distribution $P(\mathbf{x} | \mathbf{y})$ is calculated using
a database consisting of pairs
$(\mathbf{y}_1, \mathbf{x}_1), \ldots (\mathbf{y}_N, \mathbf{x}_N)$
of observations $\mathbf{y}_i$ and corresponding atmospheric states
$\mathbf{x}_i$. Assuming that the database describes the true a priori
distribution $P(\mathbf{x})$ sufficiently well, the expected
value w.r.t to the posterior distribution $P(\mathbf{x} | \mathbf{x})$ of the
atmospheric state $\mathbf{x}$ can be approximated using
\begin{align}\label{eq:gprof_retrieval}
  \int_{\mathbf{x} } \mathbf{x}\: dP(\mathbf{x} | \mathbf{y}) =
  \int_{\mathbf{x} } \mathbf{x}\: \frac{P(\mathbf{y} | \mathbf{x})}{P(\mathbf{y})} dP(\mathbf{x}) \approx
  \frac{\sum_i \mathbf{x} P(\mathbf{y}|\mathbf{x}_i)}{\sum_i P(\mathbf{y}|\mathbf{x})}
\end{align}
The conditional probability $P(\mathbf{y} | \mathbf{x}_i)$ of the observation
$\mathbf{y}$ given an atmospheric state from the database is modeled assuming
  unbiased  Gaussian errors in the observations:
  %
\begin{align}\label{eq:gprof_error}
  P(\mathbf{y}|\mathbf{x}_i) = \frac{1}{\sqrt{2\pi\text{det}(\mathbf{S}^{-1})}}
  \exp \left \{
  - \frac{1}{2}
  (\mathbf{y} - \mathbf{y}_i)^T \mathbf{S}^{-1}  (\mathbf{y} - \mathbf{y}_i)
  \right \}
\end{align}
for a certain covariance matrix $\mathbf{S}$ describing the errors due to noise
as well as other potential error sources that affect the comparison of a real
observation $\mathbf{y}$ and the observations in the database.
Eq.~(\ref{eq:gprof_retrieval}) corresponds to a resampling of the
states in the database with case-specific weights calculated using
Eq.~\ref{eq:gprof_error}. This can be extended to obtain specific quantiles of
the a posteriori distribution (described for example in \citet{pfreundschuh18})
or to derive probabilities of certain characteristics of the a posteriori state
such as the presence of precipitation in a given observation.

A difficulty of the formulation in Eq.~(\ref{eq:gprof_retrieval}) is that
calculating the retrieval result requires traversing a large database. To
speed up the retrieval, observations in the GPROF database are clustered
with respect to the database observations $\mathbf{y}_i$, so that for each
cluster of observations only the central observations, the corresponding
mean state and the number of observations in the cluster need to be stored.

Since the retrieval of complex atmospheric states from the limited information
content of a multi-spectral microwave observation is underconstrained, it is
desirable to incorporate ancillary information to increase its accuracy. In
GPROF the database is split into strata with respect to two-meter temperature
($T_{2m}$), total-column water vapor ($\text{TCWV}$) surface type ($\text{ST}$)
as well as an additional airlifting index for observations in mountain regions.

The basic functioning of the GPROF algorithm is illustrated in
Fig.~\ref{fig:gprof_legacy}. Panel (a) illustrates the reweighting of the
database entries according to Eq.~(\ref{eq:gprof_retrieval}) and
(\ref{eq:gprof_error}). The clustering of the observations to improve retrieval
performance and the stratification of the database to incorporate acillary data
is illustrated in panel (b) and (c), respectively.

\begin{figure}[hbpt]
  \centering
    \includegraphics[width=\textwidth]{../figures/gprof_legacy}
    \caption{The GPROF retrieval algorithm. Panel (a) illustrates the
      resampling of the observation based in the database to approximate the a
      posteriori distribution. Panel (b) exemplifies the clustering of
      observations that is performed to speed up the calculation of retrieval
      results. Panel (c) illustrates the incorporation of ancillary data in
      through application of a stratified retrieval database.}
  \label{fig:gprof_legacy}
\end{figure}

\subsection{Training and test data}

The neural-network-based implementations of the GPROF algorithm that will be
presented below use the same data for training and evaluation as the original
GPROF algorithm. This is done to ensure a fair comparison comparability of the
algorithms against each other as well as functional equivalence.

The main part of the GPROF retrieval database consists of atmospheric profiles
and corresponding observations derived from a year of combined radar-radiometer
observations from the Dual-frequency Precipitation Radar (DPR) and the GPM
Microwave Imager (GMI). These profiles are complemented with gauge-corrected
precipitation estimates derived from ground radar over the US for snow surfaces
and precipitation derived from ERA5 over sea ice.

The available data has been split into three parts: Data from day 6 to the end
of each month are used for training and data from day 4 and 5 are used to
monitor retrieval performance during training and to tune model hyperparameters.
The data from day 1 to 3 of each month is set aside and used for the final
comparison of the three algorithms presented in Sect.~\ref{sec:results}.

The various retrieval variables that are present in the database are listed
in Tab.~\ref{tab:variables}. While surface precipitation is certainly the
primary target of the GPROF algorithm, the observations are sensitive to
a range of additional quantities. Profiles are represented at 28 levels
with a resolution of $500\ \unit{m}$ between $0$ and $10\ \unit{km}$ and
$1 \ \unit{km}$ above that.

\begin{table}
  \caption{Retrieval quantities in the retrieval database}
  \label{tab:variables}
  \centering
  \begin{tabular}{l|r|r}
    Retrieval variable & Unit & Type\\
    \hline
    \hline

    Surface precipitation & $\unit{mm\ h^{-1}}$ & Scalar \\
    Convective precipitation & $\unit{mm\ h^{-1}}$ & Scalar \\
    Cloud water path & $\unit{kg\ m^{-2}}$ & Scalar \\
    Rain water path & $\unit{kg\ m^{-2}}$ & Scalar \\
    Ice water path & $\unit{kg\ m^{-2}}$ & Scalar \\
    Cloud water content & $\unit{g\ m^{-3}}$ & Profile \\
    Rain water content & $\unit{g\ m^{-3}}$ & Profile \\
    Snow water content & $\unit{g\ m^{-3}}$ & Profile \\
    Latent heat & $\unit{K \ h^{-1}}$ & Profile \\
  \end{tabular}
  \end{table}

\subsection{Probabilistic regression with neural networks}

The ill-posed nature of the retrieval of many atmospheric variables from
satellite observations requires a probabilistic approach in order to quantify
the inherent uncertainty of the retrieval results. In many retrievals these
uncertainties are handled by employing Bayesian methods. As shown in a previous
study \cite{pfreundschuh18}, instead of a direct application of Bayes theorem
the posterior distribution may also be predicted directly using neural networks
in a probabilistic regression framework.

The approach proposed in \citet{pfreundschuh18} uses a quantile regression
neural network (QRNN) to predict a sequence of quantiles of the posterior
distribution and has been show to yield equally reliable results as other
explicitly Bayesian retrieval methods that are commonly used to retrieve
atmospheric quantities. For each scalar retrieval variable $x$, the QRNN is
trained to predict a vector of values corresponding to the quantiles of the a
posteriori distribution by minimizing the sum of the quantile losses
\begin{align}
  \mathcal{L}_\tau(\hat{x}, x) &= (\tau - \mathbb{I}_{x < \hat{x}})(x - \hat{x})
\end{align}
for the selected quantile levels $\tau = \tau_1, \ldots, \tau_n$.

In addition to QRNNs, another approach for non-parametric probabilistic
regression has been considered for the GPROF-NN algorithm. This approach
consists of predicting a binned approximation of the probability density
function (PDF) of the posterior distribution. A prior application of this
approach can be found in \cite{metnet}. In lack of a better name, we will refer
to this approach as density regression neural network (DRNN). For each scalar
retrieval variable $x$, the DRNN is trained to classify the retrieval input into
a discrete sequence of bins covering the range of values of $x$. This is done
by converting values of the scalar retrieval variable $x$ into a categorical
variable and training the network using a categorical cross entropy loss:
\begin{align}
  \mathcal{L}(\hat{x}, x) &= \log(p_i) - \sum_{i \neq j}  \log(p_j)
\end{align}

Figure~\ref{fig:qrnn_drnn} illustrates these two approaches when applied to
retrieve surface precipitation along the central pixels of the GMI swath.
While the QRNN provides a direct estimate of the posterior CDF, the DRNN
provides an estimate of the corresponding PDF. Since both approaches provide
a direct estimate of the a posteriori distribution, the two approaches can
be used interchangeably.

\subsection{GPROF-NN 0D}

The GPROF-NN 0D algorithm retrieves all cloud- and precipitation-related
retrieval variables from a single observation and the ancillary data, which
makes it functionally equivalent to that algorithm.

\subsubsection{Network architecture}

\begin{figure}[hbpt]
  \centering
    \includegraphics[width=\textwidth]{../figures/gprof_nn_0d}
    \caption{Illustration of the neural network architecture used in the
      GPROF-NN 0D algorithm. The network consists of a common body and
      one head for each retrieval variable. Each block in body and head
      consists of fully-connected layer, layer norm and GELU activation function.
    }
  \label{fig:gprof_nn_0d}
\end{figure}

The network used in the GPROF-NN 0D algorithm is a feed-forward multi-layer
perceptron a  separate head for each of the different retrieval variables.
Profile variables share a single head and the expansion of the profile levels
is performed in the last layer of the network. The basic building block of
the network consists of a fully-connected layer followed by layer normalization
and a GELU activation function. The body of the network consists of
$N_b$ of those blocks and each head of $N_h$. Note that the case $N_b = 0$ is
equivalent to predicting each retrieval target with a separate neural network.
A shared body in the neural network allows computations to be shared between
different retrieval variables and therefore makes inference and training more
efficient.

\subsubsection{Training}

The GPROF-NN 0D algorithm is trained to predict all retrieval variable in
parallel by minimizing the sum of all losses. Each network is trained for
for 70 epochs with the Adam optimizer and a cosine annealing learning rate
schedule. Warm restarts are performed after 10, 20 and 40 epochs. The initial
learning rates for the first and second part of the training set to $5\cdot 10^{-4}$
and reduced to $10^{-4}$ for its last part.

learning rate is

\subsubsection{Validation-set performance}

\subsection{GPROF-NN 2D}

\subsubsection{Architecture}

\subsubsection{Validation-set performance}

\section{Results}

\bibliographystyle{alpha}
\bibliography{research}

\end{document}

### 2021-05-28 article/manuscript.tex

@@
-\RequirePackage[l2tabu,orthodox]{nag} % turn on warnings because of bad style
-\documentclass[a4paper,11pt,bibtotoc]{scrartcl}
-\usepackage{natbib}
+\title{GPROF-NN: A neural-network based implementation of the Goddard Profiling Algorithm}
 
-\usepackage[utf8]{inputenc}
+\documentclass[12pt]{article}
 
-\usepackage[T1]{fontenc}        % Tries to use Postscript Type 1 Fonts for better rendering
-\usepackage{lmodern}            % Provides the Latin Modern Font which offers more glyphs than the default Computer Modern
-\usepackage{amsmath} % Provides all mathematical commands
-\usepackage{amsfonts} % Provides all mathematical commands
-
-\usepackage{hyperref}           % Provides clickable links in the PDF-document for \ref
-\usepackage{grffile}            % Allow you to include images (like graphicx). Usage: \includegraphics{path/to/file}
-
-% Allows to set units
-\usepackage[ugly]{units}        % Allows you to type units with correct spacing and font style. Usage: $\unit[100]{m}$ or $\unitfrac[100]{m}{s}$
-
-% Additional packages
-\usepackage{url}                % Lets you typeset urls. Usage: \url{http://...}
-\usepackage{breakurl}           % Enables linebreaks for urls
-\usepackage{xspace}             % Use \xpsace in macros to automatically insert space based on context. Usage: \newcommand{\es}{ESPResSo\xspace}
-\usepackage{xcolor}             % Obviously colors. Usage: \color{red} Red text
-\usepackage{booktabs}           % Nice rules for tables. Usage \begin{tabular}\toprule ... \midrule ... \bottomrule
-
-% Source code listings
-\usepackage{listings}           % Source Code Listings. Usage: \begin{lstlisting}...\end{lstlisting}
-\lstloadlanguages{python}       % Default highlighting set to "python"
-
+\usepackage{amsmath}
 \begin{document}
-
-\title{GPROF-NN: A neural-network based implementation of the Goddard Profiling Algorithm}
-\date{\today}
-
 \maketitle
@@
 \section{Introduction}
-\section{Methods and data}
+\section{Methods}
 
@@
 
-The original GPROF algorithm, here referred to simply as GPROF, is based on a
-Bayesian formulation of the retrieval problem. Given an observation
-$\mathbf{y}$, the result of the retrieval is defined as the posterior
-distribution $P(\mathbf{x} | \mathbf{y})$ over the values $\mathbf{x}$ of a
-numerical representation of the atmospheric state.
+The original GPROF algorithm is based on a Bayesian formulation of the retrieval
+problem. Given an observation $\mathbf{y}$, the result of the retrieval is
+defined as the posterior distribution $P(\mathbf{x} | \mathbf{y})$ of the
+atmospheric state $\mathbf{x}$.
 
- The posterior distribution $P(\mathbf{x} | \mathbf{y})$ is calculated using
-a database consisting of pairs
+To calculate the posterior $P(\mathbf{x} | \mathbf{y})$ the algorithm uses a
+database consisting of pairs
 $(\mathbf{y}_1, \mathbf{x}_1), \ldots (\mathbf{y}_N, \mathbf{x}_N)$
-of observations $\mathbf{y}_i$ and corresponding atmospheric states
+of observations $\mathbf{y}_i$ and corresponding states
 $\mathbf{x}_i$. Assuming that the database describes the true a priori
-distribution $P(\mathbf{x})$ sufficiently well, the expected
+distribution of atmospheric states $\mathbf{x}$ sufficiently well, the expected
 value w.r.t to the posterior distribution $P(\mathbf{x} | \mathbf{x})$ of the
 atmospheric state $\mathbf{x}$ can be approximated using
-\begin{align}\label{eq:gprof_retrieval}
-  \int_{\mathbf{x} } \mathbf{x}\: dP(\mathbf{x} | \mathbf{y}) =
-  \int_{\mathbf{x} } \mathbf{x}\: \frac{P(\mathbf{y} | \mathbf{x})}{P(\mathbf{y})} dP(\mathbf{x}) \approx
-  \frac{\sum_i \mathbf{x} P(\mathbf{y}|\mathbf{x}_i)}{\sum_i P(\mathbf{y}|\mathbf{x})}
+\begin{align}
+  \int_{\mathbf{x} } f(\mathbf{x})\: d\mathbf{x} \approx
+  \frac{\sum_i f(\mathbf{x}) P(\mathbf{y}|\mathbf{x}_i)}{\sum_i P(\mathbf{y}|\mathbf{x})}
 \end{align}
-The conditional probability $P(\mathbf{y} | \mathbf{x}_i)$ of the observation
-$\mathbf{y}$ given an atmospheric state from the database is modeled assuming
-  unbiased  Gaussian errors in the observations:
-  %
-\begin{align}\label{eq:gprof_error}
-  P(\mathbf{y}|\mathbf{x}_i) = \frac{1}{\sqrt{2\pi\text{det}(\mathbf{S}^{-1})}}
+The conditional probability of the observation given an atmospheric state
+from the database, $P(\mathbf{y}|\mathbf{x}_i)$, is approximated using a
+Gaussian error model
+\begin{align}
+  P(\mathbf{y}|\mathbf{x}_i) = \frac{1}{\sqrt(2\pi\text{det}(\mathbf{S}^{-1}))}
   \exp \left \{
   - \frac{1}{2}
-  (\mathbf{y} - \mathbf{y}_i)^T \mathbf{S}^{-1}  (\mathbf{y} - \mathbf{y}_i)
-  \right \}
-\end{align}
-for a certain covariance matrix $\mathbf{S}$ describing the errors due to noise
-as well as other potential error sources that affect the comparison of a real
-observation $\mathbf{y}$ and the observations in the database.
-Eq.~(\ref{eq:gprof_retrieval}) corresponds to a resampling of the
-states in the database with case-specific weights calculated using
-Eq.~\ref{eq:gprof_error}. This can be extended to obtain specific quantiles of
-the a posteriori distribution (described for example in \citet{pfreundschuh18})
-or to derive probabilities of certain characteristics of the a posteriori state
-such as the presence of precipitation in a given observation.
-
-A difficulty of the formulation in Eq.~(\ref{eq:gprof_retrieval}) is that
-calculating the retrieval result requires traversing a large database. To
-speed up the retrieval, observations in the GPROF database are clustered
-with respect to the database observations $\mathbf{y}_i$, so that for each
-cluster of observations only the central observations, the corresponding
-mean state and the number of observations in the cluster need to be stored.
-
-Since the retrieval of complex atmospheric states from the limited information
-content of a multi-spectral microwave observation is underconstrained, it is
-desirable to incorporate ancillary information to increase its accuracy. In
-GPROF the database is split into strata with respect to two-meter temperature
-($T_{2m}$), total-column water vapor ($\text{TCWV}$) surface type ($\text{ST}$)
-as well as an additional airlifting index for observations in mountain regions.
-
-The basic functioning of the GPROF algorithm is illustrated in
-Fig.~\ref{fig:gprof_legacy}. Panel (a) illustrates the reweighting of the
-database entries according to Eq.~(\ref{eq:gprof_retrieval}) and
-(\ref{eq:gprof_error}). The clustering of the observations to improve retrieval
-performance and the stratification of the database to incorporate acillary data
-is illustrated in panel (b) and (c), respectively.
-
-\begin{figure}[hbpt]
-  \centering
-    \includegraphics[width=\textwidth]{../figures/gprof_legacy}
-    \caption{The GPROF retrieval algorithm. Panel (a) illustrates the
-      resampling of the observation based in the database to approximate the a
-      posteriori distribution. Panel (b) exemplifies the clustering of
-      observations that is performed to speed up the calculation of retrieval
-      results. Panel (c) illustrates the incorporation of ancillary data in
-      through application of a stratified retrieval database.}
-  \label{fig:gprof_legacy}
-\end{figure}
-
-\subsection{Training and test data}
-
-The neural-network-based implementations of the GPROF algorithm that will be
-presented below use the same data for training and evaluation as the original
-GPROF algorithm. This is done to ensure a fair comparison comparability of the
-algorithms against each other as well as functional equivalence.
-
-The main part of the GPROF retrieval database consists of atmospheric profiles
-and corresponding observations derived from a year of combined radar-radiometer
-observations from the Dual-frequency Precipitation Radar (DPR) and the GPM
-Microwave Imager (GMI). These profiles are complemented with gauge-corrected
-precipitation estimates derived from ground radar over the US for snow surfaces
-and precipitation derived from ERA5 over sea ice.
-
-The available data has been split into three parts: Data from day 6 to the end
-of each month are used for training and data from day 4 and 5 are used to
-monitor retrieval performance during training and to tune model hyperparameters.
-The data from day 1 to 3 of each month is set aside and used for the final
-comparison of the three algorithms presented in Sect.~\ref{sec:results}.
-
-The various retrieval variables that are present in the database are listed
-in Tab.~\ref{tab:variables}. While surface precipitation is certainly the
-primary target of the GPROF algorithm, the observations are sensitive to
-a range of additional quantities. Profiles are represented at 28 levels
-with a resolution of $500\ \unit{m}$ between $0$ and $10\ \unit{km}$ and
-$1 \ \unit{km}$ above that.
-
-\begin{table}
-  \caption{Retrieval quantities in the retrieval database}
-  \label{tab:variables}
-  \centering
-  \begin{tabular}{l|r|r}
-    Retrieval variable & Unit & Type\\
-    \hline
-    \hline
-
-    Surface precipitation & $\unit{mm\ h^{-1}}$ & Scalar \\
-    Convective precipitation & $\unit{mm\ h^{-1}}$ & Scalar \\
-    Cloud water path & $\unit{kg\ m^{-2}}$ & Scalar \\
-    Rain water path & $\unit{kg\ m^{-2}}$ & Scalar \\
-    Ice water path & $\unit{kg\ m^{-2}}$ & Scalar \\
-    Cloud water content & $\unit{g\ m^{-3}}$ & Profile \\
-    Rain water content & $\unit{g\ m^{-3}}$ & Profile \\
-    Snow water content & $\unit{g\ m^{-3}}$ & Profile \\
-    Latent heat & $\unit{K \ h^{-1}}$ & Profile \\
-  \end{tabular}
-  \end{table}
-
-\subsection{Probabilistic regression with neural networks}
-
-The ill-posed nature of the retrieval of many atmospheric variables from
-satellite observations requires a probabilistic approach in order to quantify
-the inherent uncertainty of the retrieval results. In many retrievals these
-uncertainties are handled by employing Bayesian methods. As shown in a previous
-study \cite{pfreundschuh18}, instead of a direct application of Bayes theorem
-the posterior distribution may also be predicted directly using neural networks
-in a probabilistic regression framework.
-
-The approach proposed in \citet{pfreundschuh18} uses a quantile regression
-neural network (QRNN) to predict a sequence of quantiles of the posterior
-distribution and has been show to yield equally reliable results as other
-explicitly Bayesian retrieval methods that are commonly used to retrieve
-atmospheric quantities. For each scalar retrieval variable $x$, the QRNN is
-trained to predict a vector of values corresponding to the quantiles of the a
-posteriori distribution by minimizing the sum of the quantile losses
-\begin{align}
-  \mathcal{L}_\tau(\hat{x}, x) &= (\tau - \mathbb{I}_{x < \hat{x}})(x - \hat{x})
-\end{align}
-for the selected quantile levels $\tau = \tau_1, \ldots, \tau_n$.
-
-In addition to QRNNs, another approach for non-parametric probabilistic
-regression has been considered for the GPROF-NN algorithm. This approach
-consists of predicting a binned approximation of the probability density
-function (PDF) of the posterior distribution. A prior application of this
-approach can be found in \cite{metnet}. In lack of a better name, we will refer
-to this approach as density regression neural network (DRNN). For each scalar
-retrieval variable $x$, the DRNN is trained to classify the retrieval input into
-a discrete sequence of bins covering the range of values of $x$. This is done
-by converting values of the scalar retrieval variable $x$ into a categorical
-variable and training the network using a categorical cross entropy loss:
-\begin{align}
-  \mathcal{L}(\hat{x}, x) &= \log(p_i) - \sum_{i \neq j}  \log(p_j)
+  (\mathbf{y} - \mathbf{y}_i})^T \mathbf{S}^{-1}  (\mathbf{y} - \mathbf{y}_i)}{2})
+  \right }
 \end{align}
 
-Figure~\ref{fig:qrnn_drnn} illustrates these two approaches when applied to
-retrieve surface precipitation along the central pixels of the GMI swath.
-While the QRNN provides a direct estimate of the posterior CDF, the DRNN
-provides an estimate of the corresponding PDF. Since both approaches provide
-a direct estimate of the a posteriori distribution, the two approaches can
-be used interchangeably.
-
-\subsection{GPROF-NN 0D}
-
-The GPROF-NN 0D algorithm retrieves all cloud- and precipitation-related
-retrieval variables from a single observation and the ancillary data, which
-makes it functionally equivalent to that algorithm.
-
-\subsubsection{Network architecture}
-
-\begin{figure}[hbpt]
-  \centering
-    \includegraphics[width=\textwidth]{../figures/gprof_nn_0d}
-    \caption{Illustration of the neural network architecture used in the
-      GPROF-NN 0D algorithm. The network consists of a common body and
-      one head for each retrieval variable. Each block in body and head
-      consists of fully-connected layer, layer norm and GELU activation function.
-    }
-  \label{fig:gprof_nn_0d}
-\end{figure}
-
-The network used in the GPROF-NN 0D algorithm is a feed-forward multi-layer
-perceptron a  separate head for each of the different retrieval variables.
-Profile variables share a single head and the expansion of the profile levels
-is performed in the last layer of the network. The basic building block of
-the network consists of a fully-connected layer followed by layer normalization
-and a GELU activation function. The body of the network consists of
-$N_b$ of those blocks and each head of $N_h$. Note that the case $N_b = 0$ is
-equivalent to predicting each retrieval target with a separate neural network.
-A shared body in the neural network allows computations to be shared between
-different retrieval variables and therefore makes inference and training more
-efficient.
-
-\subsubsection{Training}
-
-The GPROF-NN 0D algorithm is trained to predict all retrieval variable in
-parallel by minimizing the sum of all losses. Each network is trained for
-for 70 epochs with the Adam optimizer and a cosine annealing learning rate
-schedule. Warm restarts are performed after 10, 20 and 40 epochs. The initial
-learning rates for the first and second part of the training set to $5\cdot 10^{-4}$
-and reduced to $10^{-4}$ for its last part.
-
-learning rate is
-
-\subsubsection{Validation-set performance}
-
-\subsection{GPROF-NN 2D}
-
-\subsubsection{Architecture}
-
-\subsubsection{Validation-set performance}
-
-\section{Results}
-
-\bibliographystyle{alpha}
-\bibliography{research}
+This approach can be extended to also estimate other statistics of the a
+posteriori distribution such as its quantiles.
 

### 2021-05-28 article/manuscript.tex
Writing.

--- 2021-05-27 article/manuscript.tex
@@
 
-The GPROF-NN 0D algorithm is trained to predict all retrieval variable in
-parallel by minimizing the sum of all losses. Each network is trained for
-for 70 epochs with the Adam optimizer and a cosine annealing learning rate
+The GPROF-NN 0D algorithm is trained to predict all retrieval variables in
+parallel by simply minimizing the sum of all losses. Each network is trained for
+for 70 epochs using the Adam optimizer and a cosine annealing learning rate
 schedule. Warm restarts are performed after 10, 20 and 40 epochs. The initial
-learning rates for the first and second part of the training set to $5\cdot 10^{-4}$
-and reduced to $10^{-4}$ for its last part.
+learning rates for the first and second part of the training set to $5\cdot
+10^{-4}$ and reduced to $10^{-4}$ for its last part.
 
-learning rate is
+\subsubsection{Hyperparameter tuning}
 
-\subsubsection{Validation-set performance}
+To find a well-performing model, a range of configurations of the network
+architecture described in Fig.~\ref{fig:gprof_nn_0d} has been trained on roughly
+$10\ \unit{\%}$ of the training data and evaluated on a similar fraction of the
+validation data. Each model configuration has been trained once as a QRNN and
+once as a DRNN. The resulting mean percentage (MPE) error and mean absolute
+percentage errors (MAPE) for a subset of the retrieval variables are displayed
+in Fig.~\ref{fig:gprof_nn_0d_tuning}.
 
@@
 \subsubsection{Architecture}
+
+The network architecture for the GPROF-NN 2D algorithm is a fully-convolutional
+asymmetric encoder-decoder architecture followed by a separate network head for
+each retrieval variable. Each stage of the encoder consists of a downsampling
+layer followed by $N_B$ convolution blocks, each of which consists of two
+exception layers followed by layer activation and GELU activation functions. The
+input and outputs of each of those blocks are connected by a residual
+connection. The decoder stages contain only a single of those blocks per layer.
+Additional skip connections are inserted to provide a short cut between
+respective stages of the encoder and decoder.
+
+\subsubsection{Data augmentation}
+
+Two properties make the application of a convolutional neural network to the GMI
+observations peculiar: First, ground truth data is available only at the center
+of the input images where the GMI observations overlap with DPR observations.
+Secondly, the GMI observations are, due to the conical scanning of the sensor,
+not invariant with respect to translations. This means that when the GMI
+observations are interpreted as a rectilinear image any observed structure will
+look different at the center of the swath compared to the edges. We found that
+naive training of a convolutional neural network with this data can lead to
+artifacts outside the swath center. To avoid these, the original images in the
+training data are augmented on-the-fly with remapped versions that simulate the
+effect of the relative placement of the observations in the swath. In addition
+to this, observations from channels 10 and up are randomly set to missing in the
+outer edges of the swath to simulate the effect of the missing observations that
+effects the outer edges of the GMI swath.
 

### 2021-05-30 gprof_nn/models.py
Training updates.

@@ -28,7 +28,7 @@ for k in BINS:
     if k != "latent_heat":
         BINS[k][0] == 0.0
 
-QUANTILES = np.linspace(1e-3, 1.0 - 1e-3, 128)
+QUANTILES = np.linspace(1e-3, 1.0 - 1e-3, 64)
 
 
 class ClampedExp(nn.Module):

### 2021-05-30 scripts/train_gprof_nn_0d.py

@@ -13,7 +13,7 @@ from quantnn.data import DataFolder
 from quantnn.normalizer import Normalizer
 from quantnn.models.pytorch.logging import TensorBoardLogger
 from quantnn.metrics import ScatterPlot
-from quantnn.transformations import Log
+from quantnn.transformations import LogLinear
 
 from gprof_nn.data.training_data import GPROF0DDataset
 from gprof_nn.models import GPROFNN0D, BINS, QUANTILES
@@ -107,13 +107,17 @@ if network_type == "drnn":
                       exp_activation=False)
     xrnn = DRNN(BINS, model=model)
 elif network_type == "qrnn_exp":
+    transformation = {t: LogLinear() for t in targets}
+    transformation["latent_heat"] = None
     model = GPROFNN0D(n_layers_body,
                       n_layers_head,
                       n_neurons,
-                      128,
+                      64,
                       target=targets,
-                      exp_activation=True)
-    xrnn = QRNN(QUANTILES, model=model)
+                      exp_activation=False)
+    xrnn = QRNN(QUANTILES,
+                model=model,
+                transformation=transformation)
 else:
     model = GPROFNN0D(n_layers_body,
                       n_layers_head,
@@ -142,8 +146,9 @@ metrics = ["MeanSquaredError", "Bias", "CalibrationPlot", "CRPS"]
 scatter_plot = ScatterPlot(log_scale=True)
 metrics.append(scatter_plot)
 
+n_epochs = 10
 optimizer = optim.Adam(model.parameters(), lr=0.0005)
-scheduler = optim.lr_scheduler.ReduceLROnPlateau(optimizer, factor=0.5, patience=4)
+scheduler = optim.lr_scheduler.CosineAnnealingLR(optimizer, n_epochs)
 xrnn.train(training_data=training_data,
            validation_data=validation_data,
            n_epochs=n_epochs,
@@ -153,3 +158,30 @@ xrnn.train(training_data=training_data,
            metrics=metrics,
            device=device,
            mask=-9999)
+xrnn.save(model_path / network_name)
+n_epochs = 20
+optimizer = optim.Adam(model.parameters(), lr=0.001)
+scheduler = optim.lr_scheduler.CosineAnnealingLR(optimizer, n_epochs)
+xrnn.train(training_data=training_data,
+           validation_data=validation_data,
+           n_epochs=n_epochs,
+           optimizer=optimizer,
+           scheduler=scheduler,
+           logger=logger,
[…]

### 2021-05-30 scripts/train_gprof_nn_0d.sh

@@ -1,10 +1,11 @@
 #TRAINING_DATA=/gdata/simon/gprof_0d/era5/gmi/training_data_small
-TRAINING_DATA=/home/simonpf/src/regn/data/training_data_small
+TRAINING_DATA=/home/simonpf/src/gprof_nn/data/training_data/gmi/era5
 #VALIDATION_DATA=/gdata/simon/gprof_0d/era5/gmi/validation_data_small
-VALIDATION_DATA=/home/simonpf/src/regn/data/validation_data_small
+VALIDATION_DATA=/home/simonpf/src/gprof_nn/data/validation_data/gmi/era5
 
 MODEL_PATH=${HOME}/src/gprof_nn/models/
-TARGETS="surface_precip convective_precip rain_water_path ice_water_path cloud_water_path cloud_water_content snow_water_content rain_water_content latent_heat"
+TARGETS="surface_precip convective_precip rain_water_path ice_water_path cloud_water_path cloud_water_content snow_water_content rain_water_content"
+TARGETS="surface_precip rain_water_content"
 
 export QUANTNN_LOG_LEVEL=INFO
 export OMP_NUM_THREADS=4

### 2021-05-31 gprof_nn/models.py
Fixed docstring.

@@ -1,9 +1,8 @@
 """
-regn.models
+gprof_nn.models
 ===========
 
-
-
+Neural network models used for the GPROF-NN algorithms.
 """
 import numpy as np
 import torch

### 2021-05-31 gprof_nn/models.py
Added default argument to forward method so that model can be used in stand alone mode.

@@ -99,7 +99,7 @@ class HyperResNetFC(nn.Module):
             else:
                 self.output_activation = None
 
-    def forward(self, x, acc_in, li=1):
+    def forward(self, x, acc_in=None, li=1):
         """
         Forward input through network.
 

### 2021-05-31 gprof_nn/data/mrms.py
Working version of data extraction from sim files.

@@ -172,14 +172,6 @@ class MRMSMatchFile:
         coords_sim = latlon_to_ecef(lons, lats)
         coords_sim = np.concatenate(coords_sim, 1)
 
-        surface_types = get_surface_type_map(start_time)
-        surface_types = surface_types.interp(
-            latitude=input_data["latitude"],
-            longitude=input_data["longitude"],
-            method="nearest"
-        )
-        input_data["surface_type"].data[:] = surface_types.data
-
         kdtree = KDTree(coords_1c)
         dists, indices = kdtree.query(coords_sim)
 

### 2021-05-31 test/test_preprocessor.py
Adapt training_data classes to new sparse data format.

@@ -4,10 +4,13 @@ Tests for reading the preprocessor format.
 from pathlib import Path
 
 import numpy as np
+from quantnn.qrnn import QRNN
+from quantnn.normalizer import Normalizer
 
 from gprof_nn.data.preprocessor import PreprocessorFile
 from gprof_nn.data.training_data import (GPROF0DDataset,
-                                         write_preprocessor_file)
+                                         write_preprocessor_file,
+                                         run_retrieval_0d)
 
 def test_preprocessor_file(tmp_path):
     """
@@ -16,7 +19,7 @@ def test_preprocessor_file(tmp_path):
     dataset.
     """
     path = Path(__file__).parent
-    input_file = path / "data" / "gprof_gmi_era5.nc"
+    input_file = path / "data" / "training_data.nc"
 
     targets = ["surface_precip",
                "ice_water_path",

### 2021-05-31 gprof_nn/models.py
Changed quantiles for training.

@@ -32,7 +32,7 @@ for k in BINS:
     if k != "latent_heat":
         BINS[k][0] == 0.0
 
-QUANTILES = np.linspace(1e-3, 1.0 - 1e-3, 64)
+QUANTILES = np.linspace(0.0, 1.0, 66)[1:-1]
 
 
 class ClampedExp(nn.Module):

### 2021-05-31 scripts/train_gprof_nn_0d.py

@@ -122,7 +122,7 @@ else:
     model = GPROFNN0D(n_layers_body,
                       n_layers_head,
                       n_neurons,
-                      128,
+                      64,
                       target=targets,
                       exp_activation=False)
     xrnn = QRNN(QUANTILES, model=model)

### 2021-05-31 scripts/train_gprof_nn_0d.sh

@@ -1,13 +1,12 @@
-#TRAINING_DATA=/gdata/simon/gprof_0d/era5/gmi/training_data_small
-TRAINING_DATA=/home/simonpf/src/gprof_nn/data/training_data/gmi/era5
-#VALIDATION_DATA=/gdata/simon/gprof_0d/era5/gmi/validation_data_small
-VALIDATION_DATA=/home/simonpf/src/gprof_nn/data/validation_data/gmi/era5
+TRAINING_DATA=/gdata/simon/gprof_nn/training_data/gmi/era5
+#TRAINING_DATA=/home/simonpf/src/gprof_nn/data/training_data/gmi/era5
+VALIDATION_DATA=/gdata/simon/gprof_nn/validation_data/gmi/era5
+#VALIDATION_DATA=/home/simonpf/src/gprof_nn/data/validation_data/gmi/era5
 
 MODEL_PATH=${HOME}/src/gprof_nn/models/
 TARGETS="surface_precip convective_precip rain_water_path ice_water_path cloud_water_path cloud_water_content snow_water_content rain_water_content"
-TARGETS="surface_precip rain_water_content"
 
 export QUANTNN_LOG_LEVEL=INFO
 export OMP_NUM_THREADS=4
 
-python train_gprof_nn_0d.py  ${TRAINING_DATA} ${VALIDATION_DATA} ${MODEL_PATH} --n_neurons 512 --n_layers_body 6 --n_layers_head 2 --device cuda:0 --targets ${TARGETS} --type qrnn --batch_size 4096
+python train_gprof_nn_0d.py  ${TRAINING_DATA} ${VALIDATION_DATA} ${MODEL_PATH} --n_neurons 256 --n_layers_body 6 --n_layers_head 2 --device cuda:0 --targets ${TARGETS} --type qrnn --batch_size 4096

### 2021-09-22 article/manuscript.tex
Writing.

@@
 \section{Introduction}
+
+The Goddard Profiling Algorithm (GPROF, \authorcite{kummerow14},
+\yearcite{kummerow14}) is the operational precipitation retrieval algorithm for
+the passive microwave (PMW) observations from the satellites of the Global
+Precipitation Measurement (GPM, \authorcite{hou14}, \yearcite{hou14}) mission.
+The aim of GPM is to provide global measurements of precipitation at a temporal
+resolution of 3 hours. It is thus an essential component of the system of global
+observations that enables monitoring of the hydrological cycle for the benefit
+of science and society.
+
+The development of GPROF was originally motivated by the Tropical Rainfall
+Measurement Mission (TRMM,), the precursor of the GPM mission, and thus dates
+back almost 30 years \citep{kummerow94, kummerow96}. Since then, the algorithm
+has undergone several improvements. Methodologically, the most fundamental
+modification has been presented already in \citet{kummerow96}, which introduced
+the Bayesian retrieval scheme that is based on a database of observations and
+known, corresponding precipitation rates and forms the basis of the algorithm
+until today. Following that, improvements were mostly related to and driven by
+improvements of the database. While the first version of GPROF still used hand
+crafted hydrometeor profiles to generate the database, these were replaced by
+profiles from a meso-scale weather model \citep{kummerow96}. An important
+improvement was the replacement of the model-based database by an
+observationally-generated database for the GPROF 2010 algorithm
+\citep{kummerow11, kummerow15}, which helped reduce errors caused by
+misrepresentation of atmospheric states in the database. The 2014 version
+of GPROF \citet{kummerow15} introduced the first fully-parametric version
+of the algorithm that was designed to be applicable to all sensors of
+the GPM constellation and became the first version of the operational
+PMW precipitation retrieval.
+
+This study revisits the actual retrieval method that forms the basis of the
+current GPROF algorithm, which was first introduced in \citet{kummerow96}.
+Throughout its evolution, the retrieval database of GPROF has grown to a
+considerable size and currently contains around $10^9$ observations. This makes
+it ideally suited as training data for a machine-learning based retrieval.
+However, most common machine-learning regression techniques are incompatible
+with the Bayesian formulation of inverse problems \citep{rodgers00} upon which
+the current algorithm is based. \citet{pfreundschuh18} proposed quantile
+regression neural networks (QRNNs) as a simple approach to reconcile
+inverse-problem theory with neural-network retrievals.
+
+This study presents two novel, neural-network-based implementations of the GPROF
+algorithm. The first algorithm, referred to as GPROF-NN 0D, is designed to be
+functionally identical to the original GPROF algorithm. Similar to GPROF,
+it processes observations pixel-by-pixel and  uses the same inputs and
+produces the same outputs as GPROF. The second algorithm, referred GPROF-NN 2D,
+also produces the same outputs as GPROF, but processes all observations
+simultaneously, thus allowing it to make use of structural information in
+the observations. The aim of this study is thus to investigate (1) to
+what extent a switching to a functionally equivalent neural-network based
+retrieval (GPROF-NN 0D) can help to improve retrievals of precipitation
+and the 3D structure of hydrometeors and (2) to what extent structural
+information in the retrieval can help improve retrieval accuracy.
+
 \section{Methods and data}
 
-This section describes the original GPROF algorithm as well as the two newly
-developed neural-network-based versions.
+Since its conception, the GPROF algorithm has made use of a retrieval
+database consisting of satellite observations and corresponding known
+precipitation rates. In this sense, the retrieval method alone may be
+viewed as a machine learning method. However, since the distribution
+and structure of precipitation was unknown and remains affected by
+significant uncertainties, the focus of improvements of the GPROF
+algorithm has been generation and representativeness of the database.
+
+This section presents the retrieval database that forms the basis of the
+original and the novel GPROF algorithms as well as the data that is used
+to test the methods. This is followed by a description of the original
+algorithm and the two novel GPROF-NN 0D and GPROF-NN 2D algorithms.
 
@@
 
-The original GPROF algorithm, here referred to simply as GPROF, is based on a
-Bayesian formulation of the retrieval problem. Given an observation
-$\mathbf{y}$, the result of the retrieval is defined as the posterior
-distribution $P(\mathbf{x} | \mathbf{y})$ over the values $\mathbf{x}$ of a
-numerical representation of the atmospheric state.
-
- The posterior distribution $P(\mathbf{x} | \mathbf{y})$ is calculated using
-a database consisting of pairs
-$(\mathbf{y}_1, \mathbf{x}_1), \ldots (\mathbf{y}_N, \mathbf{x}_N)$
-of observations $\mathbf{y}_i$ and corresponding atmospheric states
-$\mathbf{x}_i$. Assuming that the database describes the true a priori
-distribution $P(\mathbf{x})$ sufficiently well, the expected
-value w.r.t to the posterior distribution $P(\mathbf{x} | \mathbf{x})$ of the
-atmospheric state $\mathbf{x}$ can be approximated using
+The original GPROF algorithm, here referred to simply as GPROF, is a Bayesian
+method in the sense that it formulates the retrieval as the task of finding the
+posterior distribution $P(\mathbf{x} | \mathbf{y})$ of the precipitation given
+an observation vector $\mathbf{y}$. As mentioned above, the posterior
+distribution $P(\mathbf{x} | \mathbf{y})$ is calculated using a database
+consisting of pairs $(\mathbf{y}_1, \mathbf{x}_1), \ldots (\mathbf{y}_N,
+\mathbf{x}_N)$ of observations $\mathbf{y}_i$ and corresponding atmospheric
+states $\mathbf{x}_i$. Assuming that the database describes the true a priori
+$P(\mathbf{x})$, i.e. the climatological distribution of surface precipitation,
+sufficiently well, the expected value w.r.t to the posterior distribution
+$P(\mathbf{x} | \mathbf{x})$ of the atmospheric state $\mathbf{x}$ can be
+approximated using
 \begin{align}\label{eq:gprof_retrieval}
@@
   \centering
-    \includegraphics[width=\textwidth]{../figures/gprof_nn_0d}
+    \includegraphics[width=\textwidth]{../figures/gprof_nn_0d_cdf}
     \caption{Illustration of the neural network architecture used in the
@@
 
+\begin{figure}[hbpt]
+  \centering
+  \includegraphics[width=0.8\textwidth]{../plots/gprof_nn_2d_arch}
+  \caption{
+    The neural network architecture employed by the GPROF-NN 2D retrieval.
+  }
+  \label{fig:data_augmenation}
+\end{figure}
+
 \subsubsection{Data augmentation}
+
+\begin{figure}[hbpt]
+  \centering
+    \includegraphics[width=\textwidth]{../figures/data_augmentation}
+    \caption{
+      Illustration of the data augmentation applied for the training of the GPROF-NN 2D algorithm.
+    }
+  \label{fig:data_augmenation}
+\end{figure}
 
@@
 
+This section presents the results of the experiments that were performed to
+asses the characteristics of the GPROF, GPROF-NN 0D and GPROF-NN 2D retrieval
+algorithms.
+
+\subsection{Results on test dataset}
+
+We begin be assessing the retrieval performance on the test data set, which
+consists of the database observations from the first three days of every month.
+Since the training data used to train the neural-network retrievals stems from
+the same time period, this allows assessing the inherent capabilities of the
+retrieval algorithms while minimizing potential aliasing effects caused by the
+inter-annual variability of precipitation.
+
+\subsubsection{Scalar retrieval variables}
+
+Retrieval results for scalar retrieval targets are scattered against their
+reference values in Fig~\ref{fig:results_scalar}. It should be noted that
+frequencies have been normalized column-wise since results for high rain rates
+would otherwise not be visible in the plots. As these results show, there is a
+consistent, gradual improvement between the GPROF, GPROF-NN 0D and GPROF-NN 2D
+algorithms for all retrieval targets.
+
+The neural-network algorithms generally show the largest improvements for small
+values where for surface precipitation and convective precipitation systematic
+deviations from the reference distribution are observed. Although improvements
+at larger values are generally smaller, they are still consistent across the
+examined ranges. The most significant improvement is observed for convective
+precipitation. Here the distributions of the GPROF and GPROF-NN 0D algorithms
+deviate notably from the diagonal but that of GPROF-NN 2D is located correctly
+over the diagonal.
+
+\begin{figure}[hbpt]
+  \centering
+  \includegraphics[width=\textwidth]{../plots/results_scatter}
+  \caption{
+    Scatter plots of scalar retrieval targets for the
+    three retrieval algorithms. The columns display the results
+    for the GPROF, GPROF-NN 0D and GPROF-NN 2D algoriths, while
+    rows display the results for different retrieval targets. Frequencies
+    in the plots have been normalized column-wise, i.e. per bin of
+    the reference value.
+  }
+  \label{fig:data_augmenation}
+\end{figure}
+
+Since the surface type has a strong effect on the lower-frequency observations
+used in the retrieval, its impact on the retrieval of surface precipitation is
+assessed in Fig.~\ref{fig:results_surface}. The figure displays the Bias, RMSE
+and MAPE for the different surface classes that are distinguished in the
+retrieval for each of the three algorithms.
+
+Also when considered separately for different surface types the results of the
+surface precipitation retrieval show the same overall pattern as observed in
+the scatter plots in Fig.~\ref{fig:results_scalar}: The results of the GPROF-NN 0D
+retrieval are more accurate than those of the GPROF algorithm and the results
+of the GPROF-NN 2D algorithm are in-turn better than those of the GPROF-NN 0D
+algorithm. These results are observed consistently across all surface types
+and considered metric.
+
+\begin{figure}[hbpt]
+  \centering
+  \includegraphics[width=\textwidth]{../plots/results_surface_types}
+  \caption{
+    Bias, root-mean squared error (RMSE), and maximum absolute percentage error (MAPE) for
+    the three retrieval algorithms and different surface types.
+  }
+  \label{fig:data_augmenation}
+\end{figure}
+
+In addition to quantitative precipitation estimates, the GPROF algorithm also provides a
+probabilistic classification of pixels into raining and non-raining pixels by predicting
+a probability of precipitation. As described in ~\ref{sec:method}, a similar quantity can
+be derived from the probabilistic prediction of the QRNN-NN algorithms. To assess the quality
+of these predictions, the calibration of the predicted probability and the receiver-operating
+characteristic curve are displayed in Fig.~\ref{fig:pop}.
+
+The calibration curves show, that predictions from all algorithms are fairly
+well calibrated, i.e. the predicted probabilities agree well with those observed
+on the independent test data. Although all algorithms exhibit small deviations from the 
+diagonal, the GPROF-NN algorithms provide slightly better calibrated predictions.
+
+For the ROC curve, the results are more conclusive and show a clear improvement in classification
+skill for the GPROF-NN 0D algorithm compared to GPROF. The GPROF-NN 2D algorithm further improves
+the classification slightly compared to the GPROF-NN 0D algorithm.
+
+\begin{figure}[hbpt]
+  \centering
+  \includegraphics[width=\textwidth]{../plots/results_pop}
+  \caption{
+    Calibration (Panel (a)) and receiver-operating characteristic (ROC, Panel(b)) for
+    the predicted probability of precipitation.
+  }
+  \label{fig:roc}
+\end{figure}
+
+\subsubsection{Profile retrieval variables}
+
+Finally, we also assess the retrieval performance of profile retrievals.
+Profiles of Bias and RMSE for the three retrieved profile quantities are
+displayed in Fig.~\ref{fig:profiles}. The results show a significant improvement
+in bias and RMSE for the GPROF-NN retrievals for all three quantities.
+However, there is only a small difference between the GPROF-NN 0D and
+GPROF-NN 2D algorithms.
+
+\subsubsection{Probabilistic precipitation retrievals}
+
+The error statistics for the quantitative retrieval targets presented above all
+used the mean of the a posteriori distribution to compare against the GPROF
+algorithm. Traditionally, GPROF produces only 4 output values that make use of
+information from the posterior distribution exceeding the posterior mean: The
+estimated probability of precipitation, the precipitation flag, the 1st and 3rd
+terciles of the surface precipitation (provided as a measure of uncertainty) and
+the most likely precipitation. The neural-network version of GPROF predict the
+full posterior distribution for the scalar retrieval variables, and so storing
+only the posterior mean discards a large part of this information.
+
+Since storing the full posterior distribution would require excessive amounts
+of storage space, we here aim to explore the benefits of providing a random
+sample from the posterior distribution as retrieval result. Although the
+sampling neglects spatial correlation results, the results should be
+able to truthfully reproduce the distribution of precipitation values in the
+test data when averaged over multiple orbits.
+
+To illustrate the statistical benefits of randomly-sampled retrieval estimates
+we begin by examining the zonal distributions of surface precipitation on
+$1^\circ$-degree bins. Fig.~\ref{fig:results_zonal} displays the reference
+distribution of the test data (filled contours) together with the quantiles of
+the retrieved distributions (shaded lines) for the case when the posterior mean
+is used as the retrieval estimate (1st row) and for the case when a random
+sample is used as retrieval estimate (2nd row). Note that the color scaling
+has been chosen so that if the retrieved and reference distribution were similar,
+the shaded lines would lie on top of the boundaries between the filled contours
+in the background.
+
+As the first row of panels shows, the retrieved distribution does not match the
+corresponding distribution of the test data. Especially the high quantiles of
+the reference distribution are strongly underestimated by the retrieved
+distributions. However, when instead of the posterior mean random samples are
+used as precipitation estimates, the agreement between the retrieved and the
+test distribution is significantly improved. Although some discrepancies remain
+for the lower quantiles of the distribution, very good agreement is found between
+the high quantiles of the distribution.
+
+  \begin{figure}[hbpt]
+    \centering
+    \includegraphics[width=\textwidth]{../plots/zonal_distributions}
+    \caption{
+      Zonal distributions of test data and corresponding retrieved values. Colored, filled contours
+      in the background show the CDFs of the zonal distribution of precipitation in the test data.
+      Grey-shaded line display the quantiles of the corresponding retrieved distributions based on
+      the posterior mean (Row 1) and random samples from the posterior distribution (Row 2).
+    }
+    \label{fig:results_zonal}
+  \end{figure}
+
+  For a different perspective on these results, we have also considered the
+  ability of the retrieval to reproduce specific percentiles of the test data on
+  a spatial grid of size $\SI{5}{\degree}\times\SI{5}{\degree}$. When the
+  distribution of percentiles derived from the retrieved values are plotted
+  against the percentiles of the test data (Fig.\ \ref{fig:scatter_percentiles})
+  it becomes evident that using the posterior mean (grey-shaded, filled
+  contours) leads to systematic underestimation of the true value. However, when
+  random samples are used these systematic deviations are corrected and the
+  percentiles of the retrieved distributions scatter symmetrically around the
+  line of equality.
+
+  \begin{figure}[hbpt]
+    \centering
+    \includegraphics[width=\textwidth]{../plots/scatter_percentiles}
+    \caption{
+      95th (Row 1) and 99th (Row 2) percentiles of the distribution
+      of surface precipitation in $5^\circ \times 5^\circ$ boxes in the
+      test data scattered against the corresponding retrieved values. Filled
+      contours in the background show the distribution obtained by using the
+      posterior mean as retrieved value and colored contours the distribution
+      that is obtained when a random sample from the posterior distribution
+      is used as retrieval value. The columns show the results for the
+      GPROF-NN 0D (Column 1) and the GPROF-NN 2D algorithm  (Column 2).
+    }
+    \label{fig:scatter_percentiles}
+  \end{figure}
+
+\begin{figure}[hbpt]
+  \centering
+  \includegraphics[width=\textwidth]{../plots/results_spatial}
+  \caption{
+    Spatial distributions of Bias (Column 1), RMSE (Column 2) and MAPE (Column 3) error statistics for the
+    GPROF (Row 1), GPROF-NN 0D (Row 2), GPROF-NN 2D (Row 3) algorithms.
+  }
+    \label{fig:results_spatial}
+\end{figure}
+
+\subsection{Case study: Hurricane Dorian}
+
+To conclude this investigation into the retrieval performance of the newly
+developed GPROF retrievals, we consider a overpass of the GPM Core Observatory
+over Hurricane Dorian that occurred on September, 1st 2019 at 20:20 UTC.
+Since. This allows us to compare the results of the GPROF retrieval to the
+GPM combined product at the center of the swath, which has been averaged using
+a Gaussian kernel corresponding to a footprint with a full-width at half-maximum
+of $\SI{18}{\kilo \meter}$ in along-track and $\SI{10}{\kilo \meter}$ in
+across-track direction.
+
+To allow for a detailed comparison, Fig.~\ref{fig:hurricane_precip} displays the
+retrieved surface precipitation on a log-scale over a region of $\SI{2000}{\kilo
+  \meter}$ as well as on a linear scale on a smaller region close to the
+hurricane. When considered at log-scale, obvious structural differences in the
+retrieved precipitation are visible: The GPROF retrieval produces large swaths
+of low precipitation that cover most of the scene, which are not present in the
+combined retrieval. They are clearly reduced in the results of the GPROF-NN 0D
+retrieval and even more so in the results of the GPROF-NN 2D retrieval which
+agrees well with the results from the combined retrieval.
+
+The two rightmost panels of Fig.~\ref{fig:hurricane_precip} display the the
+regions of high precipitation around the eye of the hurricane in more detail.
+The most salient features in the surface precipitation from the combined
+algorithm is the eye wall and the concentric rain band situated south-west of
+the eye. Both the high precipitation withing the eyewall as well as the rainband
+are visible in the results of the neural-network versions of GPROF. While the
+surface precipitation in the outer rain band is retrieved accurately both
+retrievals overestimate the precipitation in the righter flank of the eyewall.
+However, compared to the standard GPROF algorithm, the results agree
+considerably better with those of the combined algorithm.
+
+\begin{figure}[hbpt]
+  \centering
+  \includegraphics[width=\textwidth]{../plots/hurricane_surface_precip}
+  \caption{
+    Retrieved surface precipitation from the GPM combined product (Panels (a),
+    (c)), the GPROF algorithm (Panels (b), (d)), and the neural network versions
+    GPROF-NN 0D (Panels (e), (g)) and GPROF-NN 2D (Panels (f), (h)) around
+    Hurricane Dorian on 2019-09-01 21:23:00 UTC. Panels (a), (b), (e), (f)
+    display the precipitation on a logarithmic scale over an area
+    spanning $20\ 000\ \unit{km}$ in zonal and meridional directions. Panels (c),
+    (d), (g), (f) display the surface precipitation on a linear color scale in
+    an enlarged area spanning $5\ 000\ \unit{km}$ around the Hurricane. In the
+    background the closest true-color composite from GOES 16 is shown.
+  }
+  }
+  \label{fig:hurricane_precip}
+\end{figure}
+
+Figure~\ref{fig:hurricane_3d} displays the retrieved RWC and SWC for the same
+observation of Hurricane Dorian. The four panels show isosurfaces the retrieved
+water contents from the combined retrieval, GPROF and the GPROF-NN 0D and 2D
+algorithms. Also here, the agreement with the results from the combined retrieval
+is better for the GPROF-NN retrievals than for GPROF. Especially, the GPROF-NN 2D
+agrees well with the combined results.
+
+\begin{figure}[hbpt]
+  \centeringn
+  \includegraphics[width=\textwidth]{../plots/dorian.r4.0548.png}
+  \caption{
+    3D isosurfaces of retrieved rain water content (RWC, red) and snow water
+    content (SWC, blue) for the same scene as depicted in
+    Fig.~\ref{fig:hurricane_precip}. Different panels display the results from
+    the GPM combined product, GPROF and the GPROF-NN 0D and 2D algorithms.
+  }
+    \label{fig:hurricane_3d}
+\end{figure}
+
+\section{Discussion}
+
+Above we have presented two neural network based implementations of the GPROF
+algorithm. Both algorithms were designed to be functionally equivalent to the
+current implementation which means that they can potentially replace the current
+method in an upcoming version. We have assessed the performance of the three
+algorithms on test data that was not seen by any of the neural network based
+retrieval methods before the evaluation. Here we have seen a significant and
+consistent improvements of the retrieval accuracy across all considered metrics
+and outputs. Similar tendencies were found when comparing retrieval results from
+a Hurricane overpass outside the database period to the GPM combined retrieval.
+
+\subsection{Retrieval performance}
+
+Our results show a consistent improvement of the GPROF-NN retrievals over the
+original GPROF algorithm. This is of course an encouraging result considering
+that all three retrievals are based on exactly the same retrieval database,
+indicating that the GPROF retrievals can be improved simply by switching to a
+more powerful method.
+
+But why do the neural-network-based retrievals perform better than the
+traditional approach? To investigate this, we have calculated the gradients of
+the retrieved surface precipitation with respect to the brightness temperatures
+for the GPROF and GPROF-NN 0D retrievals. Comparison of the mean gradients and
+their component-wise standard deviation reveals that the GPROF-NN 0D retrieval
+exhibits larger gradients with more variation than the GPROF retrieval. This
+indicates that the retrieval manifold that describes the relation between input
+observations and retrieved surface precipitation exhibits more structure for the
+neural network retrieval than it does for standard GPROF.
+
+Based on the formulation of the standard GPROF retrieval, the retrieval may be
+viewed as nearest-neighbor regression with a Gaussian kernel and a diagonal
+covariance matrix. Although the theoretical formulation of the retrieval assumes
+the elements of the covariance matrix to correspond to the standard deviations
+of the random errors in the observations, these errors are in practice inflated
+to account for sparsity of the retrieval database. Our results suggest that this
+may limit the ability of the method to reproduce structure of the retrieval
+manifold leading to a loss of sensitivity.
+
+We find additional improvements in retrieval accuracy when structural
+information is included in the retrieval by means of processing scenes of
+observations simultaneously instead of pixel-by-pixel. Intuitively, this makes
+sense since a human observer with some experience with microwave imagery can
+identify characteristic rain structures. An interesting continuation of
+this work would be to investigate, which structural patterns the neural
+network has learned, which we aim to pursue in a follow-up study.
+
+\subsection{The importance of ancillary data}
+
+An operational aspect that considerably complicates the GPROF algorithm is
+the required ancillary data. Currently two version of GPROF exist: The
+near real time version, which uses forecasts from the GANAL model as source
+for the ancillary data, and a climate version, which uses ERA5 data from
+\citep{era5}. For the development of the GPROF-NN algorithms, we have
+investigated how much the inclusion of ancillary actually improves the
+retrieval. The results are presented in the appendix. Although omitting
+the ancillary data has a negative impact on the retrieval performance,
+both retrievals still perform better than the original GPROF
+implementation. Moreover, the GPROF-NN 2D retrieval without ancillary
+data performs only marginally worse than the retrieval with ancillary
+data and better than both variants of the GPROF-NN 0D retrieval.
+
+Although this may not have any direct consequences on the operational
+application of GPROF, it still has interesting consequences for the development
+of future retrievals. For one, this obviously means that incorporating spatial
+information into a MW precipitation retrieval improves the retrieval to a larger
+extent than adding ancillary information from secondary sources without adding
+latency due to data availability constraints. But, this also opens up the
+possibility of distributing the retrieval algorithm instead of the retrieval.
+The neural network model for the GPROF-NN 2D algorithm has a size of about
+$\SI{70}{\mega \byte}$, which makes it roughly as large as the level 1C file
+required to run the retrieval. This use case may be interesting for the
+development of specialized retrievals that have a lower general interest but
+could in this way be made available to the community without the need to set up
+an operational processing pipline. An example would be a high-resolution
+hydrometeor profile, which may be of interest for case studies but whose output
+would be too large to archive operationally.
+
+\subsection{Limitations}
+
+Our main finding is that the accuracy of the GPROF retrieval can be
+significantly improved by replacing the retrieval method with a more expressive
+one. However, The majority of the experiments on which this finding is based use
+data from the time period covered by the retrieval database. Based on these
+results alone it is therefor not possible conclude that these models will
+generalize to other time periods. This is why, as a next step, we will test the
+both neural-network based retrieval algorithms in an operational setting to see
+how they perform along side the standard GPROF retrieval.
+
+Another aspect that has been neglected in this study is the accuracy of the
+retrieval database that is used both in standard GPROF and the neural-network
+version. The database is derived from the GPM combined product combined with
+MIRS, ERA5 and MRMS. Its accuracy will thus always be limited by the accuracy of
+those products. To what degree the improvements in the GPROF retrieval translate
+to independent validation data thus remains to be analyzed in a future study.
+
+Finally, another point that should be mentioned here is that although both
+GPROF-NN are porbabilistic in the sense that they provide an estimate of
+the postior distribution, they model the distributions of the output values
+as independent distributions. This means that, although uncertainties are
+well calibrated on average, they may not be on the scale of single
+precipitation systems. However, the is also the case for the standard
+GPROF retrieval and the authors are note aware of any method that could solve
+this issue in ana operational context.
+
+\section{Conclusions}
+
+The GPROF-NN algorithms and their performance assessments that were presented in
+this study constitute a first step towards a neural network based implementation
+of the passive precipitation retrievals for the GPM constellation. Although it
+has to be taken into account that our experiments neglect relevant errors
+sources that will ultimately add to the full retrieval uncertainty, our results
+remain encouraging: Simply by replacing the retrieval with a neural network
+based method the accuracy of the retrieval can be increased at the same time as
+the computational complexity of the retrieval is decreased. This potentially
+makes the upgrade of the retrieval implementation a very simple and cheap way
+of improving global precipitation measurements.
+
+As a next step we will run extend the implementation of the GPROF-NN 0D and
+GPROF-NN 2D algorithms to the other sensors of the GPM constellation and run
+them operationally along side standard GPROF. The three methods will then
+be validated and assessed using gauge-corrected gound-based radar measurements
+in a follow up study. This is to ensure that the upgrade of the retrieval method
+doesn't degrade the quality of the operational GPM PMW retrievals.
+
+In addition to potential improvements in retrieval accuracy through an upgrade
+of the retrieval method, we presented two novel applications that would be
+enabled by the neural-network based retrievals:
+
+\begin{enumerate}
+\item The inclusion of random samples from the posterior distribution in the
+  retrieval output: We have presented this as an efficient way of representing
+  retrieval uncertainties that helps to improve the representation of extreme
+  events in the retrieval results.
+\item Running retrieval 'on the edge': Our results show that incorporating
+  structural information into the retrieval yields larger performance benefits
+  than incorporating ancillary information. This allows simplifying the
+  retrieval pipeline to the point that only observations are required as input,
+  which, together with the availability of high-performance machine learning
+  software, opens up the possibility letting the user run the retrieval on his
+  own computer, i.e. to perform the inference 'on the edge'. This may be
+  beneficial for retrievals with very high resolution, i.e. whose storage
+  requirement are much larger than those of the L1C data, and which may be
+  interesting to users on a case-by-case basis.
+\end{enumerate}
+
+Building on the approach proposed in \citet{pfreundschuh18}, this work presents
+a way of integrating and reconciling machine learning methods with retrieval
+pipelines built on Bayesian retrieval algorithms. We find that the machine
+learning based retrieval, which is built on readily available, high-performance
+machine learning libraries, is easiert to implement, runs faster and performs
+better than the hand-crafted retrieval algorithm. Although it remains to be seen
+to which extent this will improve retrievals with respect to independent
+validation data, it seems reasonable to expect at least incremental improvements
+in global precipitation measurements as well as interesting opportunities for
+the novel retrieval applications.
+
 \bibliographystyle{alpha}

### 2021-11-29 gprof_nn/data/bin.py
Refactored reading routines for bin files.

@@ -3,10 +3,8 @@
 gprof_nn.data.bin
 =================
 
-This module contains interfaces to read the binned database database
-format of GPROF V7.
+This module contains interfaces to read the GPROF V7 bin files.
 """
-import asyncio
 import logging
 from pathlib import Path
 from concurrent.futures import ProcessPoolExecutor
@@ -14,11 +12,10 @@ import re
 
 import numpy as np
 import pandas as pd
+from rich.progress import track
 import xarray as xr
-import tqdm.asyncio
 
 from gprof_nn import sensors
-from gprof_nn.definitions import PROFILE_NAMES
 
 LOGGER = logging.getLogger(__name__)
 N_LAYERS = 28
@@ -146,9 +143,9 @@ class BinFile:
             dim_dict[self.sensor.n_angles] = "angles"
 
         record_type = self.sensor.get_bin_file_record(self.surface_type)
-        for k, t, *shape in record_type.descr:
-            if k == "scan_time":
-                data = self.handle[k]
+        for key, _, *shape in record_type.descr:
+            if key == "scan_time":
+                data = self.handle[key]
                 date = pd.DataFrame(
                     {
                         "year": data[:, 0],
@@ -159,20 +156,20 @@ class BinFile:
                         "second": data[:, 5],
                     }
                 )
-                results[k] = (("samples",), pd.to_datetime(date, errors="coerce"))
+                results[key] = (("samples",), pd.to_datetime(date, errors="coerce"))
             else:
                 dims = ("samples",)
                 if shape:
                     dims = dims + tuple([dim_dict[s] for s in shape[0]])
-                results[k] = dims, self.handle[k][indices]
+                results[key] = dims, self.handle[key][indices]
 
         results["surface_type"] = (
             ("samples",),
-            self.surface_type * np.ones(n_samples, dtype=np.int),
+            self.surface_type * np.ones(n_samples, dtype=np.int32),
         )
         results["airmass_type"] = (
             ("samples",),
-            self.airmass_type * np.ones(n_samples, dtype=np.int),
+            self.airmass_type * np.ones(n_samples, dtype=np.int32),
         )
         source = 0
         if self.surface_type in [8, 9, 10, 11]:
@@ -181,11 +178,14 @@ class BinFile:
             source = 2
         results["source"] = (
             ("samples",),
-            source * np.ones(n_samples, dtype=np.int),
+            source * np.ones(n_samples, dtype=np.int32),
         )
 
[…]

### 2021-11-29 test/test_bin.py

@@ -1,6 +1,6 @@
 """
-Tests for the extraction of training samples for tehe GPROF-NN 0D
-algorithm from .bin files.
+This file tests the 'gprof_nn.data.bin' module which provides functionality
+to read and extract training data from the 'bin' files used by GPROF.
 """
 from pathlib import Path
 
@@ -8,28 +8,79 @@ import numpy as np
 import xarray as xr
 
 from gprof_nn import sensors
-from gprof_nn.data.bin import (FileProcessor,
-                               BinFile)
-from gprof_nn.data.training_data import GPROF_NN_0D_Dataset
+from gprof_nn.data.bin import FileProcessor, BinFile
+from gprof_nn.data.training_data import GPROF_NN_1D_Dataset
+
+
+DATA_PATH = Path(__file__).parent / "data"
 
 
 def test_bin_file_gmi():
     """
-    Test reading of GMI bin files and ensure all values are physical.
+    Test reading the different types of bin files for standard surface types,
+    for sea ice and snow.
     """
-    path = Path(__file__).parent
-    input_file = BinFile(path / "data" / "gmi" / "gpm_291_55_04.bin").to_xarray_dataset()
+    #
+    # Simulator-derived bin files.
+    #
 
-    assert np.all(input_file["surface_precip"] >= 0)
-    assert np.all(input_file["convective_precip"] >= 0)
-    assert np.all(input_file["rain_water_path"] >= 0)
-    assert np.all(input_file["two_meter_temperature"] > 291 - 0.5)
-    assert np.all(input_file["two_meter_temperature"] < 291 + 0.5)
-    assert np.all(input_file["total_column_water_vapor"] > 55 - 0.5)
-    assert np.all(input_file["total_column_water_vapor"] < 55 + 0.5)
-    assert np.all(input_file["surface_type"] == 4)
-    assert np.all(input_file["airmass_type"] == 0)
-    tbs = input_file.brightness_temperatures.data
+    input_file = DATA_PATH / "gmi" / "bin" / "gpm_275_14_03_17.bin"
+    input_data = BinFile(input_file).to_xarray_dataset()
+
+    assert np.all(input_data["surface_precip"] >= 0)
+    assert np.all(input_data["convective_precip"] >= 0)
+    assert np.all(input_data["rain_water_path"] >= 0)
+    assert np.all(input_data["two_meter_temperature"] > 275 - 0.5)
[…]

### 2021-11-29 gprof_nn/data/sim.py
Refactored gprof_nn.data.sim module.

@@ -10,7 +10,7 @@ temperatures.
 The module also provides functionality to extract the training data for the
 GPROF-NN algorithm from these files.
 """
-from concurrent.futures import ProcessPoolExecutor, TimeoutError
+from concurrent import futures
 from datetime import datetime
 import logging
 import os
@@ -66,13 +66,37 @@ GENERIC_HEADER = np.dtype(
 class SimFile:
     """
     Interface class to read GPROF .sim files.
+
+    The main purpose of this class is to provide and interface to read
+    .sim files and convert them to 'xarray.Dataset' objects via the
+    'to_xarray_dataset' method.
+
+    Attributes:
+        path: The path of the file
+        granule: The GPM CO granule number to which this file corresponds.
+        date: Date object specifying the day of the corresponding GPM orbit.
+        sensor: Sensor object representing the sensor corresponding to
+            this file.
+        header: Numpy structured array containing the header data of the
+            file.
+        data: Numpy structured array containing raw data of the file.
     """
 
     @classmethod
     def find_files(cls, path, sensor=sensors.GMI, day=None):
         """
-        Find all files that match the standard filename pattern for GMI
-        sim files.
+        Find all files that match the standard filename pattern for
+        sim files for the given sensor.
+
+        Args:
+            path: Root of the directory tree in which to look for .sim
+                files.
+            sensor: The sensor for which to find .sim files.
+            day: If given search is restricted to the given day within
+                each month.
+
+        Return:
+            A list containing the found .sim files.
         """
         if day is None:
             pattern = sensor.sim_file_pattern.format(day="??")
@@ -86,7 +110,7 @@ class SimFile:
 
     def __init__(self, path):
         """
-        Open .sim file.
+        Open a .sim file.
 
         Args:
             path: Path to the .sim file to open.
@@ -94,16 +118,18 @@ class SimFile:
         self.path = path
         parts = str(path).split(".")
         self.granule = int(parts[-2])
-        self.year = int(parts[-3][:4])
-        self.month = int(parts[-3][4:6])
-        self.day = int(parts[-3][6:])
+
+        year = int(parts[-3][:4])
+        month = int(parts[-3][4:6])
+        day = int(parts[-3][6:])
+        self.date = datetime(year, month, day)
 
[…]

### 2021-11-29 test/test_sim.py

@@ -1,5 +1,6 @@
 """
-Tests for the reading and processing of .sim files.
+This file contains tests for the reading and processing of .sim files
+defined in 'gprof_nn.data.sim.py'.
 """
 from pathlib import Path
 
@@ -20,6 +21,7 @@ from gprof_nn.data.sim import (SimFile,
 from gprof_nn.data.preprocessor import PreprocessorFile
 
 
+DATA_PATH = Path(__file__).parent / "data"
 HAS_ARCHIVES = Path(sensors.GMI.l1c_file_path).exists()
 
 
@@ -28,13 +30,17 @@ def test_open_sim_file():
     Tests reading a GMI L1C file and matching it to data in
     a GMI .sim file.
     """
-    data_path = Path(__file__).parent / "data"
-    sim_file = SimFile(data_path / "GMI.dbsatTb.20190101.027510.sim")
+    input_file = DATA_PATH / "gmi" / "sim" / "GMI.dbsatTb.20190101.027510.sim"
+    sim_file = SimFile(input_file)
     data = sim_file.to_xarray_dataset()
 
+    assert "surface_precip" in data.variables.keys()
     assert "latent_heat" in data.variables.keys()
     assert "snow_water_content" in data.variables.keys()
     assert "rain_water_content" in data.variables.keys()
+
+    valid = data.surface_precip > -9999
+    assert np.all(data.surface_precip[valid] >= 0.0)
 
 
 def test_match_l1c_gmi():
@@ -42,8 +48,8 @@ def test_match_l1c_gmi():
     Tests reading a GMI L1C file and matching it to data in
     a GMI .sim file.
     """
-    data_path = Path(__file__).parent / "data"
-    l1c_file = L1CFile.open_granule(27510, data_path, sensors.GMI)
+    l1c_path = DATA_PATH / "gmi" / "l1c"
+    l1c_file = L1CFile.open_granule(27510, l1c_path, sensors.GMI)
     l1c_data = l1c_file.to_xarray_dataset()
     tbs = l1c_data.brightness_temperatures.data
     tbs_new = np.zeros(tbs.shape[:2] + (15,))
@@ -53,7 +59,8 @@ def test_match_l1c_gmi():
     l1c_data["brightness_temperatures"] = (("scans", "pixels", "channels"),
                                            tbs_new)
 
-    sim_file = SimFile(data_path / "GMI.dbsatTb.20190101.027510.sim")
+    bin_path = DATA_PATH / "gmi" / "sim"
[…]

### 2021-11-29 gprof_nn/bin/retrieve.py
Some fixes to gprof_nn commands.

@@ -152,7 +152,7 @@ def run(args):
         output_files = []
         for f in input_files:
             of = f.relative_to(input)
-            if of.suffix in [".nc", ".HDF5"]:
+            if of.suffix in [".nc", ".nc.gz", ".HDF5"]:
                 of = of.with_suffix(".nc")
             else:
                 of = of.with_suffix(".bin")

### 2021-11-29 gprof_nn/bin/run_preprocessor.py

@@ -90,7 +90,10 @@ def process_file(
             sensor,
             configuration=configuration
     )
-    
+
+    if not output.parent.exists():
+        output.parent.mkdir(parents=True, exist_ok=True)
+
     data.to_netcdf(output)
 
 

### 2021-11-29 gprof_nn/bin/statistics.py

@@ -60,8 +60,13 @@ def add_parser(subparsers):
 
 
 STATS = {
-    "training": [
+    "training_1d": [
         statistics.TrainingDataStatistics(kind="1d"),
+        statistics.ZonalDistribution(),
+        statistics.GlobalDistribution()
+    ],
+    "training_3d": [
+        statistics.TrainingDataStatistics(kind="3d"),
         statistics.ZonalDistribution(),
         statistics.GlobalDistribution()
     ],
@@ -80,9 +85,10 @@ STATS = {
 }
 
 PATTERNS = {
-    "training": "**/*.nc*",
+    "training_1d": "**/*.nc*",
+    "training_3d": "**/*.nc*",
     "bin": "**/*.bin",
-    "observations": "**/*.HDF5",
+    "observations": "**/*.nc",
     "retrieval": "*.nc",
     "combined": "**/2B.GPM*.HDF5"
 }
@@ -108,7 +114,10 @@ def run(args):
         )
         return 1
 
-    kind = args.kind.lower()
+    kind = args.kind.lower().strip()
+    if kind == "training":
+        kind = "training_1d"
+
     if not kind in STATS.keys():
         LOGGER.error(
             f"'kind' argument must be one of {list(STATS.keys())}."

### 2021-11-29 gprof_nn/bin/train.py

@@ -277,8 +277,8 @@ def run(args):
                         validation_data,
                         output,
                         args)
-    elif variant.upper() == "2D":
-        run_training_2d(sensor,
+    elif variant.upper() == "3D":
+        run_training_3d(sensor,
                         configuration,
                         training_data,
                         validation_data,
@@ -293,7 +293,7 @@ def run(args):
                          args)
     else:
         raise ValueError(
-                "'variant' should be one of '1D', '2D', 'SIM'."
+                "'variant' should be one of '1D', '3D', 'SIM'."
                 )
 
 
@@ -373,7 +373,7 @@ def run_training_1d(sensor,
         dataset_factory,
         kwargs=kwargs,
         queue_size=128,
-        n_workers=4
+        n_workers=6
     )
 
     kwargs = {
@@ -389,7 +389,7 @@ def run_training_1d(sensor,
         dataset_factory,
         kwargs=kwargs,
         queue_size=128,
-        n_workers=2
+        n_workers=4
     )
 
     #
@@ -495,7 +495,7 @@ def run_training_1d(sensor,
         xrnn.save(output)
 
 
-def run_training_2d(sensor,
+def run_training_3d(sensor,
                     configuration,
                     training_data,
                     validation_data,
@@ -521,6 +521,8 @@ def run_training_2d(sensor,
     from quantnn.metrics import ScatterPlot
     import torch
     from torch import optim
+    torch.multiprocessing.set_sharing_strategy('file_system')
+    torch.set_num_threads(1)
 
     n_blocks = args.n_blocks[0]
     n_neurons_body = args.n_neurons_body
@@ -597,10 +599,10 @@ def run_training_2d(sensor,
         xrnn = None
 
     if xrnn is None:
+        LOGGER.info(
+            f"Creating new model of type {network_type}."
+        )
         if network_type == "drnn":
-            LOGGER.info(
-                f"Creating new model of type {network_type}."
-            )
             xrnn = GPROF_NN_3D_DRNN(sensor,
                                     n_blocks,
                                     n_neurons_body,
@@ -706,6 +708,9 @@ def run_training_sim(sensor,
     import torch
     from torch import optim
 
+    torch.multiprocessing.set_sharing_strategy('file_system')
+    torch.set_num_threads(1)
+
     n_blocks = args.n_blocks[0]
[…]

### 2021-11-29 gprof_nn/data/bin.py
Various fixes.

@@ -38,7 +38,7 @@ class BinFile:
     Attributes:
         temperature(``float``): The surface temperature corresponding to the
             bin.
-        tpw(``float``): The total precipitable water corresponding to the bin.
+        tcwv(``float``): The total precipitable water corresponding to the bin.
         surface_type(``int``): The surface type corresponding to the bin.
         airmass_type(``int``): The airmass type corresponding to the bin.
         header: Structured numpy array containing header data of the file.
@@ -60,7 +60,7 @@ class BinFile:
 
         parts = Path(filename).name[:-4].split("_")
         self.temperature = float(parts[1])
-        self.tpw = float(parts[2])
+        self.tcwv = float(parts[2])
         self.surface_type = int(parts[-1])
 
         if len(parts) == 4:
@@ -87,7 +87,7 @@ class BinFile:
 
         np.random.seed(
             np.array(
-                [(self.temperature), self.tpw, self.surface_type, self.airmass_type]
+                [(self.temperature), self.tcwv, self.surface_type, self.airmass_type]
             ).astype(np.int64)
         )
         self.indices = np.random.permutation(self.n_profiles)
@@ -181,7 +181,7 @@ class BinFile:
             source * np.ones(n_samples, dtype=np.int32),
         )
 
-        results["tpw"] = (("samples"), self.tpw * np.ones(n_samples, dtype=np.float32))
+        results["tcwv_bin"] = (("samples"), self.tcwv * np.ones(n_samples, dtype=np.float32))
         results["temperature"] = (
             ("samples",),
             self.temperature * np.ones(n_samples, np.float32),
@@ -235,8 +235,8 @@ class FileProcessor:
         path,
         t2m_min=227.0,
         t2m_max=307.0,
-        tpw_min=0.0,
-        tpw_max=76.0,
+        tcwv_min=0.0,
+        tcwv_max=76.0,
         include_profiles=False,
     ):
         """
@@ -248,8 +248,8 @@ class FileProcessor:
                 bins.
             t2m_max: The maximum bin surface temperature for which to consider
                 bins.
-            tpw_min: The minimum bin-tpw value to consider.
-            tpw_max: The maximum bin-tpw value to consider.
+            tcwv_min: The minimum bin-tcwv value to consider.
+            tcwv_max: The maximum bin-tcwv value to consider.
 
         """
         self.path = path
@@ -261,10 +261,10 @@ class FileProcessor:
             if match:
                 groups = match.groups()
[…]

### 2021-11-29 gprof_nn/data/cdf.py

@@ -100,6 +100,7 @@ class CdfCorrection:
         tcwv = self.corrections.total_column_water_vapor.data
         self.tcwv_min = tcwv.min()
         self.tcwv_max = tcwv.max()
+        self.surface_types = self.corrections.surface_type.size
 
     def _apply_correction_cross_track(self,
                                       sensor,
@@ -113,13 +114,16 @@ class CdfCorrection:
         tcwv = total_column_water_vapor
         tbs = brightness_temperatures.copy()
 
-        # CDF uses 3 surface type:
-        # - 1: Ocean
-        # - 2: Land
-        # - 3: Coast
-        st_inds = st.astype(np.int32)
-        st_inds[st_inds > 1] = 2
-        st_inds[st == 13] = 1
+        if self.surface_types == 3:
+            # CDF uses 3 surface type:
+            # - 1: Ocean
+            # - 2: Land
+            # - 3: Coast
+            st_inds = st.astype(np.int32)
+            st_inds[st_inds > 1] = 2
+            st_inds[st == 13] = 1
+        else:
+            st_inds = st.astype(np.int32)
 
         n_bins = sensor.n_angles
         eia_inds = np.digitize(np.abs(eia), sensor.angle_bins)
@@ -138,12 +142,15 @@ class CdfCorrection:
 
             if augment:
                 quantiles = self.corrections.cdf.data[
-                    st_inds - 1, :, eia_inds, tcwv_inds, tbs_inds
+                    st_inds - 1, i, eia_inds, tcwv_inds, tbs_inds
                 ]
                 err_lo, err_hi = np.random.normal(size=2)
                 err = (1.0 - quantiles) * err_lo + quantiles * err_hi
-                err = 0.2 * corrections * err
+                err = 0.05 * corrections * err
+                shape = corrections.shape
                 corrections += err
+
+            #corrections[st_inds > 1] = 0.0
 
             tbs[..., i] += corrections
 

### 2021-11-29 gprof_nn/data/l1c.py

@@ -192,7 +192,7 @@ class L1CFile:
             header = data.attrs["FileHeader"].decode().split()[6:8]
             satellite = header[0].split("=")[1][:-1]
             sensor = header[1].split("=")[1][:-1]
-            self.sensor = sensors.get_sensor(satellite, sensor)
+            self.sensor = sensors.get_sensor(sensor, platform=satellite)
 
 
     @property

### 2021-11-29 gprof_nn/data/preprocessor.py

@@ -12,6 +12,7 @@ Additionally, it defines functions to run the preprocessor on the CSU
 from datetime import datetime
 import logging
 import os
+import pickle
 import shutil
 import subprocess
 import tempfile
@@ -895,3 +896,11 @@ TWB_INTERP_LAND = sp.interpolate.interp1d(
 TWB_INTERP_OCEAN = sp.interpolate.interp1d(
     TWB_TABLE[:, 0] + 273.15, TWB_TABLE[:, 2], assume_sorted=True, kind="linear"
 )
+
+def get_orbit_header(sensor):
+    sensor_name = sensor.name
+    platform = sensor.platform.name
+    key = (sensor_name.upper(), platform.upper())
+    if key in ORBIT_HEADERS:
+        return pickle.loads(ORBIT_HEADERS[key])
+    return pickle.loads(DEFAULT_HEADER)

### 2021-11-29 gprof_nn/data/training_data.py

@@ -97,6 +97,113 @@ def decompress_and_load(filename):
     return data
 
 
+def write_preprocessor_file_xtrack(input_data, output_file):
+    if not isinstance(input_data, xr.Dataset):
+        data = xr.open_dataset(input_data)
+    else:
+        data = input_data
+
+    sensor = data.attrs["sensor"]
+    platform = data.attrs["platform"].replace("-", "")
+    sensor = sensors.get_sensor(sensor, platform)
+    sensor_name = sensor.name
+    platform_name = sensor.platform.name
+
+    eia = input_data.earth_incidence_angle.data
+    bins = sensor.viewing_geometry.get_earth_incidence_angles()
+    bins = 0.5 * (bins[1:] + bins[:-1])
+    indices = np.digitize(eia, bins)
+    cts, _ = np.histogram(indices, bins = np.arange(bins.size + 1) - 0.5)
+
+    n_scans = cts.max()
+    n_pixels = sensor.viewing_geometry.pixels_per_scan
+    n_chans = sensor.n_chans
+
+    if "pixels" not in data.dims or "scans" not in data.dims:
+        dim_offset = -1
+    else:
+        if hasattr(data, "samples"):
+            dim_offset = 1
+        else:
+            dim_offset = 0
+
+    new_dataset = {
+        "scans": np.arange(n_scans),
+        "pixels": np.arange(n_pixels),
+        "channels": np.arange(n_chans),
+    }
+
+    dims = ("scans", "pixels", "channels")
+    for k in data:
+        da = data[k]
+        if k == "scan_time":
+            t = da.data.ravel()[0]
+            new_dataset[k] = (("scans",), np.repeat(t, n_scans))
+        else:
+            new_shape = (n_scans, n_pixels) + da.shape[(2 + dim_offset) :]
+            new_shape = new_shape[: len(da.data.shape) - dim_offset]
+            dims = ("scans", "pixels") + da.dims[2 + dim_offset :]
+            if "pixels_center" in dims:
+                continue
+
+            new_data = -9999.9 * np.ones(new_shape, da.data.dtype)
+            for i in range(n_pixels):
+                mask = indices == i
+                n_elems = mask.sum()
+                new_data[:n_elems, i] = da.data[mask]
+
+            if new_data.dtype in [np.float32, np.float64]:
+                new_data = np.nan_to_num(new_data, nan=-9999.9)
+
+            if k == "airimass_type":
+                new_data[new_data <= 0] = 1
+            new_dataset[k] = (dims, new_data)
+
[…]

### 2021-11-29 gprof_nn/legacy.py

@@ -15,12 +15,13 @@ import shutil
 import numpy as np
 import xarray as xr
 
-import gprof_nn.logging
 from gprof_nn import sensors
 from gprof_nn.definitions import ALL_TARGETS
 from gprof_nn.data.preprocessor import PreprocessorFile, run_preprocessor
 from gprof_nn.data.retrieval import RetrievalFile
-from gprof_nn.data.training_data import GPROF_NN_1D_Dataset, write_preprocessor_file
+from gprof_nn.data.training_data import (write_preprocessor_file,
+                                         GPROF_NN_1D_Dataset,
+                                         GPROF_NN_3D_Dataset)
 from gprof_nn.sensors import CrossTrackScanner, ConicalScanner
 
 
@@ -64,7 +65,7 @@ DEFAULT_SENSITIVITIES = np.load(
 
 def has_gprof():
     """
-    Determin whether the legacy GPROF algorithm is available
+    Determine whether the legacy GPROF algorithm is available
     on the system.
     """
     return shutil.which(EXECUTABLES["STANDARD"]) is not None
@@ -181,8 +182,7 @@ def execute_gprof(
                 error.stderr,
             )
             return None
-        else:
-            raise error
+        raise error
     results = RetrievalFile(
         output_file, has_profiles=has_profiles, has_sensitivity=has_sensitivity
     )
@@ -190,7 +190,13 @@ def execute_gprof(
 
 
 def run_gprof_training_data(
-    sensor, configuration, input_file, mode, profiles, nedts=None
+        sensor,
+        configuration,
+        input_file,
+        mode,
+        profiles,
+        nedts=None,
+        preserve_structure=False
 ):
     """
     Runs GPROF algorithm on training data in GPROF-NN format and includes
@@ -198,7 +204,8 @@ def run_gprof_training_data(
 
     Args:
         sensor: The sensor for which to run GPROF.
-        configuration: The configuration with which to run GPROF.
+        configuration: The configuration with which to run GPROF ('ERA5'
+             or 'GANAL'.
         input_file: Path to the NetCDF file containing the validation
             data.
         mode: The mode in which to run GPROF ('STANDARD', 'SENSITIVITY'
@@ -211,25 +218,41 @@ def run_gprof_training_data(
         'xarray.Dataset' containing the retrieval results.
     """
     targets = ALL_TARGETS + ["latitude", "longitude"]
-    input_data = GPROF_NN_1D_Dataset(
-        input_file,
[…]

### 2021-11-29 gprof_nn/models.py

@@ -769,20 +769,20 @@ class XceptionBlock(nn.Module):
 
             self.block_1 = nn.Sequential(
             SeparableConv3x3(channels_in, channels_out),
-            nn.GroupNorm(1, channels_out),
+            nn.GroupNorm(32, channels_out),
             SymmetricPadding(1),
             nn.MaxPool2d(kernel_size=3, stride=stride),
             nn.GELU())
         else:
             self.block_1 = nn.Sequential(
                 SeparableConv3x3(channels_in, channels_out),
-                nn.GroupNorm(1, channels_out),
+                nn.GroupNorm(32, channels_out),
                 nn.GELU(),
             )
 
         self.block_2 = nn.Sequential(
             SeparableConv3x3(channels_out, channels_out),
-            nn.GroupNorm(1, channels_out),
+            nn.GroupNorm(32, channels_out),
             nn.GELU(),
         )
 
@@ -853,7 +853,7 @@ class UpsamplingStage(nn.Module):
                                     align_corners=False)
         self.block = nn.Sequential(
             SeparableConv3x3(n_channels * 2, n_channels),
-            nn.GroupNorm(1, n_channels),
+            nn.GroupNorm(32, n_channels),
             nn.GELU(),
         )
 
@@ -894,7 +894,7 @@ class MLPHead(nn.Module):
             self.layers.append(
                 nn.Sequential(
                     nn.Conv2d(n_inputs, n_hidden, 1),
-                    nn.GroupNorm(1, n_hidden),
+                    nn.GroupNorm(n_hidden, n_hidden),
                     nn.GELU(),
                 )
             )
@@ -1089,6 +1089,86 @@ class XceptionFPN(nn.Module):
             y = self.heads[k](x)
             results[k] = y
         return results
+
+    def copy_model(self, sensor):
+        """
+        Create a new model for a different model and copy  weights
+        from this model.
+
+        Note: Only GMI based convolutional model can be used to derive
+        models for other sensors.
+
+        Args:
+            sensor: The sensor for which to create a new model.
+
+        Return:
+            A new model for the given sensor.
+        """
+        n_blocks = [
+            len(self.down_block_2) - 1,
+            len(self.down_block_4) - 1,
+            len(self.down_block_8) - 1,
+            len(self.down_block_16) - 1,
+            len(self.down_block_32) - 1
+        ]
+
+        n_features_body = self.in_block.out_channels
[…]

### 2021-11-29 gprof_nn/retrieval.py

@@ -256,7 +256,6 @@ class RetrievalDriver:
     retrieval using different neural network models and writing output to
     different formats.
     """
-
     def __init__(
         self,
         input_file,
@@ -265,6 +264,7 @@ class RetrievalDriver:
         output_file=None,
         device="cpu",
         compress=True,
+        preserve_structure=False
     ):
         """
         Create retrieval driver.
@@ -280,6 +280,8 @@ class RetrievalDriver:
             device: String identifying the device to run the retrieval on.
                 "cpu" to run on CPU or "cuda:i" to run on the ith GPU.
             compress: If set to ``True`` NetCDF output will be gzipped.
+            preserve_structure: Special option that will ensure that the
+                spatial structure is conserved even for the 1D retrieval.
         """
         input_file = Path(input_file)
         self.input_file = input_file
@@ -324,6 +326,8 @@ class RetrievalDriver:
         else:
             self.output_file = output_file
 
+        self.preserve_structure = preserve_structure
+
     def _load_input_data(self):
         """
         Load retrieval input data.
@@ -338,7 +342,10 @@ class RetrievalDriver:
                 self.input_file, self.model.normalizer
             )
         else:
-            input_data = self.model.netcdf_class(
+            loader_class = self.model.netcdf_class
+            if loader_class == NetcdfLoader1D and self.preserve_structure:
+                loader_class = NetcdfLoader1DFull
+            input_data = loader_class(
                 self.input_file, normalizer=self.model.normalizer
             )
         return input_data
@@ -455,7 +462,10 @@ class RetrievalDriver:
         results.to_netcdf(self.output_file)
         if self.compress:
             subprocess.run(["gzip", "-f", self.output_file], check=True)
-        return self.output_file
+            output_file = self.output_file.with_suffix(".nc.gz")
+        else:
+            output_file = self.output_file
+        return output_file
 
 
 class RetrievalGradientDriver(RetrievalDriver):
@@ -637,6 +647,7 @@ class NetcdfLoader1D(GPROF_NN_1D_Dataset):
         i_start = i * self.batch_size
         i_end = i_start + self.batch_size
         x = torch.tensor(self.x[i_start:i_end])
+
         return x
 
     def finalize(self, data):
@@ -747,6 +758,78 @@ class NetcdfLoader3D(GPROF_NN_3D_Dataset):
 # Kept for backwards compatibility.
[…]

### 2021-11-29 gprof_nn/sensors.py

@@ -528,6 +528,7 @@ class ConicalScanner(Sensor):
         super().__init__(
             types.CONICAL, name, channels, angles, platform, viewing_geometry
         )
+        self._n_angles = 1
         n_chans = len(channels)
 
         self._mrms_file_path = mrms_file_path
@@ -862,6 +863,9 @@ class CrossTrackScanner(Sensor):
             self.correction = None
         else:
             self.correction = CdfCorrection(correction)
+        # Convert modeling error to list of channel-wise modeling errors.
+        if isinstance(modeling_error, int):
+            modeling_error = [modelling_error] * n_chans
         self.modeling_error = modeling_error
 
     def __repr__(self):
@@ -1075,7 +1079,13 @@ class CrossTrackScanner(Sensor):
                     * np.all(mask_tbs, axis=(-2, -1))
                     * np.all(mask_biases, axis=-1)
                 )
-                eia = rng.uniform(angles_min, angles_max, size=mask.sum())
+                if augment:
+                    eia = rng.uniform(angles_min, angles_max, size=mask.sum())
+                else:
+                    angles = self.viewing_geometry.get_earth_incidence_angles()
+                    indices = np.arange(mask.sum()) % angles.size
+                    eia = angles[indices]
+
                 weights = calculate_interpolation_weights(np.abs(eia), self.angles)
             else:
                 weights = None
@@ -1095,7 +1105,7 @@ class CrossTrackScanner(Sensor):
             st = scene.surface_type.data[mask]
 
             if self.correction is not None:
-                tbs = self.correction(self, st, eia, tcwv, tbs)
+                tbs = self.correction(self, st, eia, tcwv, tbs, augment=augment)
 
             if augment and self.nedt is not None:
                 noise = rng.normal(size=tbs.shape)
@@ -1175,7 +1185,7 @@ class CrossTrackScanner(Sensor):
         j_end = int(center[1, 0, 0] + width // 2)
         eia = self.viewing_geometry.get_earth_incidence_angles()
         eia = eia[j_start:j_end]
-        weights = calculate_interpolation_weights(eia, self.angles)
[…]

### 2021-11-29 gprof_nn/statistics.py

@@ -14,6 +14,7 @@ class.
 """
 from abc import ABC, abstractmethod
 from concurrent.futures import ProcessPoolExecutor
+import gc
 import logging
 import multiprocessing
 from pathlib import Path
@@ -1184,11 +1185,10 @@ class TrainingDataStatistics(Statistic):
             self._initialize_data(sensor, dataset)
 
         st = dataset["surface_type"]
-        sp = dataset["surface_precip"]
 
         for i in range(18):
             # Select only TBs that are actually used for training.
-            i_st = ((st == i + 1) * (sp >= 0)).data
+            i_st = (st == i + 1).data
             tcwv = dataset["total_column_water_vapor"].data[i_st]
             tbs = dataset["brightness_temperatures"].data[i_st]
 
@@ -1896,10 +1896,13 @@ class ObservationStatistics(Statistic):
         bins = np.arange(0, 19) + 0.5
         cs, _ = np.histogram(st, bins=bins)
         self.st += cs
-        at = dataset["airmass_type"]
+        at = dataset["airmass_type"].data
         bins = np.arange(-1, 4) + 0.5
         cs, _ = np.histogram(at, bins=bins)
         self.at += cs
+
+        dataset.close()
+        del dataset
 
     def merge(self, other):
         """
@@ -1920,7 +1923,6 @@ class ObservationStatistics(Statistic):
             self.tcwv += other.tcwv
             self.st += other.st
             self.at += other.at
-
 
     def save(self, destination):
         """
@@ -2309,7 +2311,7 @@ class StatisticsProcessor:
         LOGGER.info("Starting processing of %s files.", len(self.files))
 
         pool = ProcessPoolExecutor(n_workers)
-        batches = [[f] for f in np.random.permutation(self.files)]
+        batches = _split_files(self.files, 100) #[[f] for f in np.random.permutation(self.files)]
 
         log_queue = gprof_nn.logging.get_log_queue()
         tasks = []
@@ -2326,6 +2328,8 @@ class StatisticsProcessor:
             gprof_nn.logging.log_messages()
             for s_old, s_new in zip(stats, t.result()):
                 s_old.merge(s_new)
+                del s_new
+                gc.collect()
 
         for s in stats:
             s.save(output_path)

### 2021-11-29 scripts/train_gprof_nn_0d.sh

@@ -4,7 +4,7 @@ CONFIG=era5
 TRAINING_DATA=/gdata1/simon/gprof_nn/training_data/${SENSOR}/${CONFIG}
 VALIDATION_DATA=/gdata1/simon/gprof_nn/validation_data/${SENSOR}/${CONFIG}
 
-MODEL_PATH=${HOME}/src/gprof_nn/models/gprof_nn_1d_mhs_noaa19_era5.pckl
+MODEL_PATH=${HOME}/src/gprof_nn/models/gprof_nn_1d_mhs_noaa19_era5_corrected_full.pckl
 TARGETS="surface_precip convective_precip rain_water_path ice_water_path cloud_water_path cloud_water_content snow_water_content rain_water_content latent_heat"
 #TARGETS="surface_precip rain_water_content"
 #TARGETS="surface_precip"
@@ -12,4 +12,4 @@ TARGETS="surface_precip convective_precip rain_water_path ice_water_path cloud_w
 export QUANTNN_LOG_LEVEL=INFO
 export OMP_NUM_THREADS=2
  
-gprof_nn train 1D MHS_NOAA19 ERA5 ${TRAINING_DATA} ${VALIDATION_DATA} ${MODEL_PATH} --n_neurons_body 512 --n_layers_body 8 --n_neurons_head 256 --n_layers_head 2 --activation GELU --residuals hyper --device cuda:1 --targets ${TARGETS} --type qrnn_exp --batch_size 2048 --n_epochs 10 --learning_rate 0.0005 --no_ancillary
+gprof_nn train 1D MHS_NOAA19_FULL ERA5 ${TRAINING_DATA} ${VALIDATION_DATA} ${MODEL_PATH} --n_neurons_body 512 --n_layers_body 8 --n_neurons_head 256 --n_layers_head 2 --activation GELU --residuals hyper --device cuda:1 --targets ${TARGETS} --type qrnn_exp --batch_size 2048 --n_epochs 1 --learning_rate 0.0005 --no_ancillary

### 2021-11-29 scripts/train_gprof_nn_2d.sh

@@ -1,12 +1,14 @@
 TRAINING_DATA=/gdata1/simon/gprof_nn/training_data/gmi/era5
+TRAINING_DATA=/gdata1/simon/gprof_nn/training_data/mhs_sim/era5
 #TRAINING_DATA=/home/simonpf/src/gprof_nn/data/training_data/gmi/era5
 VALIDATION_DATA=/gdata1/simon/gprof_nn/validation_data/gmi/era5
+VALIDATION_DATA=/gdata1/simon/gprof_nn/validation_data/mhs_sim/era5
 #VALIDATION_DATA=/home/simonpf/src/gprof_nn/data/validation_data/gmi/era5
 
-MODEL_PATH=${HOME}/src/gprof_nn/models/gprof_nn_2d_gmi_era5_na.pckl
+MODEL_PATH=${HOME}/src/gprof_nn/models/gprof_nn_3d_mhs_era5_new.pckl
 TARGETS="surface_precip convective_precip rain_water_path ice_water_path cloud_water_path cloud_water_content snow_water_content rain_water_content latent_heat"
 
 export QUANTNN_LOG_LEVEL=INFO
 export OMP_NUM_THREADS=2
 
-gprof_nn train 2D  GMI ERA5 ${TRAINING_DATA} ${VALIDATION_DATA} ${MODEL_PATH} --n_neurons_head 256 --n_neurons_body  512 --n_blocks 6 --n_layers_head 4 --device cuda:1 --targets ${TARGETS} --type drnn --batch_size 4 --n_epochs 30 --learning_rate 0.0005 --activation GELU
+gprof_nn train 3D  MHS_NOAA19_FULL ERA5 ${TRAINING_DATA} ${VALIDATION_DATA} ${MODEL_PATH} --n_neurons_head 256 --n_neurons_body  512 --n_blocks 4 --n_layers_head 4 --device cuda:1 --targets ${TARGETS} --type qrnn_exp --batch_size 8 --n_epochs 10 --learning_rate 0.0005 --activation GELU

### 2021-11-29 scripts/train_simulator_mhs.sh

@@ -8,4 +8,4 @@ MODEL_PATH=${HOME}/src/gprof_nn/models/
 export QUANTNN_LOG_LEVEL=INFO
 export OMP_NUM_THREADS=1
  
-gprof_nn train SIM MHS ERA5 ${TRAINING_DATA} ${VALIDATION_DATA} ${MODEL_PATH} --n_neurons_body 256 --n_neurons_head 256 --n_layers_head 4  --device cuda:1 --batch_size 32
+gprof_nn train SIM MHS ERA5 ${TRAINING_DATA} ${VALIDATION_DATA} ${MODEL_PATH} --n_neurons_body 256 --n_neurons_head 128 --n_layers_head 4  --device cuda:0 --batch_size 4 --n_epochs 20 --learning_rate 0.0005

### 2021-11-29 test/test_legacy.py

@@ -6,7 +6,8 @@ from pathlib import Path
 import numpy as np
 import pytest
 
-from gprof_nn.data.training_data import (GPROF_NN_0D_Dataset,
+from gprof_nn import sensors
+from gprof_nn.data.training_data import (GPROF_NN_1D_Dataset,
                                          write_preprocessor_file)
 from gprof_nn.legacy import (has_gprof,
                              write_sensitivity_file,
@@ -26,17 +27,22 @@ def test_write_sensitivity_file(tmp_path):
     """
     nedts_ref = DEFAULT_SENSITIVITIES
     sensitivity_file = tmp_path / "sensitivities.txt"
-    write_sensitivity_file(sensitivity_file, nedts=nedts_ref)
+    write_sensitivity_file(sensors.GMI, sensitivity_file, nedts=nedts_ref)
     nedts = np.loadtxt(sensitivity_file)
     assert np.all(np.isclose(nedts_ref, nedts))
 
 
 @pytest.mark.skipif(not HAS_GPROF, reason="GPROF executable missing.")
 def test_run_gprof_training_data():
+    """
+    Test running the legacy GPROF algorithm on training data.
+    """
     path = Path(__file__).parent
     input_file = path / "data" / "gmi" / "gprof_nn_gmi_era5.nc"
 
-    results = run_gprof_training_data(input_file,
+    results = run_gprof_training_data(sensors.GMI,
+                                      "ERA5",
+                                      input_file,
                                       "STANDARD",
                                       False)
     assert "surface_precip" in results.variables
@@ -44,10 +50,34 @@ def test_run_gprof_training_data():
 
 
 @pytest.mark.skipif(not HAS_GPROF, reason="GPROF executable missing.")
+def test_run_gprof_training_data_preserve_structure():
+    """
+    Test running the legacy GPROF algorithm on training data while
+    preserving the spatial structure.
+    """
+    path = Path(__file__).parent
+    input_file = path / "data" / "gmi" / "gprof_nn_gmi_era5.nc"
+
+    results = run_gprof_training_data(sensors.GMI,
+                                      "ERA5",
+                                      input_file,
+                                      "STANDARD",
+                                      False,
+                                      presever_structure=True)
+    assert "surface_precip" in results.variables
+    assert "surface_precip_true" in results.variables
+
+
+@pytest.mark.skipif(not HAS_GPROF, reason="GPROF executable missing.")
 def test_run_gprof_standard():
+    """
+    Test running legacy GPROF on a preprocessor input file.
+    """
[…]

### 2021-11-29 test/test_retrieval.py

@@ -91,14 +91,14 @@ def test_retrieval_read_and_write(tmp_path):
                                  rtol=1e-2))
 
 
-def test_retrieval_preprocessor_0d_gmi(tmp_path):
+def test_retrieval_preprocessor_1d_gmi(tmp_path):
     """
     Ensure that GPROF-NN 1D retrieval works with preprocessor input.
     """
     data_path = Path(__file__).parent / "data"
     input_file = data_path / "gmi" / "GMIERA5_190101_027510.pp"
 
-    qrnn = QRNN.load(data_path / "gmi" / "gprof_nn_0d_gmi_era5.pckl")
+    qrnn = QRNN.load(data_path / "gmi" / "gprof_nn_1d_gmi_era5.pckl")
     driver = RetrievalDriver(input_file,
                              qrnn,
                              ancillary_data=data_path,
@@ -107,7 +107,7 @@ def test_retrieval_preprocessor_0d_gmi(tmp_path):
     data = RetrievalFile(output_file).to_xarray_dataset()
     assert "rain_water_content" in data.variables
 
-def test_retrieval_l1c_0d_gmi_na(tmp_path):
+def test_retrieval_l1c_1d_gmi_na(tmp_path):
     """
     Ensure that GPROF-NN 1D retrieval works with preprocessor input.
     """
@@ -116,7 +116,7 @@ def test_retrieval_l1c_0d_gmi_na(tmp_path):
         data_path / "gmi" /
         "1C-R.GPM.GMI.XCAL2016-C.20180124-S000358-E013632.022190.V05A.HDF5"
     )
-    qrnn = QRNN.load(data_path / "gmi" / "gprof_nn_0d_gmi_era5_na.pckl")
+    qrnn = QRNN.load(data_path / "gmi" / "gprof_nn_1d_gmi_era5_na.pckl")
     driver = RetrievalDriver(input_file,
                              qrnn,
                              ancillary_data=data_path,
@@ -126,14 +126,14 @@ def test_retrieval_l1c_0d_gmi_na(tmp_path):
     assert "rain_water_content" in data.variables
 
 
-def test_retrieval_preprocessor_0d_mhs(tmp_path):
+def test_retrieval_preprocessor_1d_mhs(tmp_path):
     """
     Ensure that GPROF-NN 1D retrieval works with preprocessor input.
     """
     data_path = Path(__file__).parent / "data"
[…]

### 2021-11-30 gprof_nn/bin/legacy.py
Some more fixes.

@@ -12,6 +12,7 @@ from concurrent.futures import ProcessPoolExecutor
 from pathlib import Path
 
 from rich.progress import track
+import pandas as pd
 
 import gprof_nn.logging
 from gprof_nn import sensors
@@ -61,6 +62,9 @@ def add_parser(subparsers):
     parser.add_argument('--full_profiles',
                         action='store_true',
                         help="Whether to include full profiles in the results.")
+    parser.add_argument('--preserve_structure',
+                        action='store_true',
+                        help="Whether or not to preserve the spatial structure.")
     parser.add_argument('--n_processes',
                         metavar="n",
                         type=int,
@@ -74,9 +78,10 @@ def process_file(sensor,
                  input_file,
                  output_file,
                  profiles,
-                 mode,
-                 nedts,
-                 log_queue):
+                 log_queue,
+                 mode="STANDARD",
+                 nedts=None,
+                 preserve_structure=False):
     """
     Helper function for distributed processing.
     """
@@ -85,12 +90,15 @@ def process_file(sensor,
     LOGGER.info("Processing file %s.", input_file)
 
     if input_file.suffix in [".gz", ".nc"]:
-        results = run_gprof_training_data(sensor,
-                                          configuration,
-                                          input_file,
-                                          mode,
-                                          profiles,
-                                          nedts=nedts)
+        results = run_gprof_training_data(
+            sensor,
+            configuration,
+            input_file,
+            mode,
+            profiles,
+            nedts=nedts,
+            preserve_structure=preserve_structure
+        )
     else:
         results = run_gprof_standard(sensor,
                                      configuration,
@@ -155,6 +163,8 @@ def run(args):
     else:
         mode = "STANDARD"
 
+    preserve_structure = args.preserve_structure
+
     profiles = args.profiles
     n_procs = args.n_processes
 
@@ -176,16 +186,19 @@ def run(args):
         output_files = []
         for f in input_files:
             of = f.relative_to(input)
-            of = of.with_suffix(".nc")
+            if of.suffix == ".gz":
+                of = of.with_suffix("")
             output_files.append(output / of)
-            output_files = [output]
+        print(output_files)
     else:
         input_files = [input]
         if output.is_dir():
-            output_files = [output / input.name]
+            filename = input.name
+            if filename.endswith(".gz"):
+                filename = filename[:-3]
+            output_files = [output / filename]
         else:
             output_files = [output]
-
 
     #
     # Run retrieval.
@@ -195,15 +208,11 @@ def run(args):
[…]

### 2021-11-30 gprof_nn/bin/train.py

@@ -563,7 +563,7 @@ def run_training_3d(sensor,
         dataset_factory,
         queue_size=256,
         kwargs=kwargs,
-        n_workers=6)
+        n_workers=4)
 
     kwargs = {
         "batch_size": 4 * batch_size,

### 2021-11-30 gprof_nn/data/training_data.py

@@ -1002,7 +1002,7 @@ class GPROF_NN_3D_Dataset:
             new_dataset["earth_incidence_angle"] = (dims[:-1], eia)
 
         dims = ("samples", "scans", "pixels", "layers")
-        for k, v in self.y.items():
+        for k, v in y.items():
             n_dims = v.ndim
             if n_dims > 3:
                 v = np.transpose(v, (0, 2, 3, 1))

### 2021-11-30 gprof_nn/data/types.py

@@ -290,6 +290,8 @@ def get_sim_file_record(n_chans, n_angles, n_layers, kind):
                 ("elevation", "f4"),
                 ("scan_time", date_type),
                 ("surface_type", "i4"),
+                ("surface_precip_combined", "f4"),
+                ("surface_precip_mirs", "f4"),
                 ("surface_precip", "f4"),
                 ("convective_precip", "f4"),
                 ("emissivity", "f4", (n_chans,)),

### 2021-11-30 gprof_nn/legacy.py

@@ -13,15 +13,17 @@ from pathlib import Path
 import shutil
 
 import numpy as np
+import pandas as pd
 import xarray as xr
 
-from gprof_nn import sensors
 from gprof_nn.definitions import ALL_TARGETS
-from gprof_nn.data.preprocessor import PreprocessorFile, run_preprocessor
+from gprof_nn.data.preprocessor import run_preprocessor
 from gprof_nn.data.retrieval import RetrievalFile
-from gprof_nn.data.training_data import (write_preprocessor_file,
-                                         GPROF_NN_1D_Dataset,
-                                         GPROF_NN_3D_Dataset)
+from gprof_nn.data.training_data import (
+    write_preprocessor_file,
+    GPROF_NN_1D_Dataset,
+    GPROF_NN_3D_Dataset,
+)
 from gprof_nn.sensors import CrossTrackScanner, ConicalScanner
 
 
@@ -103,8 +105,8 @@ def execute_gprof(
     sensor,
     configuration,
     input_file,
-    mode,
-    profiles,
+    mode="Standard",
+    profiles=False,
     nedts=None,
     robust=False,
 ):
@@ -114,6 +116,9 @@ def execute_gprof(
     Args:
         working_directory: The folder to use to store temporary files and
             execute the retrieval.
+        sensor: The sensor from which the input data originates.
+        configuration: Which configuration of the retrieval to run ('ERA5'
+            or 'GANAL')
         mode: Whether to include gradients or profiles in the
             retrieval.
         profiles: Whether profiles should be retrieved.
@@ -190,13 +195,13 @@ def execute_gprof(
 
 
 def run_gprof_training_data(
-        sensor,
-        configuration,
-        input_file,
-        mode,
-        profiles,
-        nedts=None,
-        preserve_structure=False
+    sensor,
+    configuration,
+    input_file,
+    mode,
+    profiles,
+    nedts=None,
+    preserve_structure=False,
 ):
     """
     Runs GPROF algorithm on training data in GPROF-NN format and includes
@@ -243,18 +248,14 @@ def run_gprof_training_data(
     with TemporaryDirectory() as tmp:
         tmp = Path(tmp)
 
-
         for batch in input_data:
 
             preprocessor_file = tmp / "input.pp"
             batch_input = input_data.to_xarray_dataset(batch=batch)
             new_dataset = write_preprocessor_file(batch_input, preprocessor_file)
             if new_dataset is not None:
[…]

### 2021-11-30 gprof_nn/models.py

@@ -1115,7 +1115,6 @@ class XceptionFPN(nn.Module):
         n_features_body = self.in_block.out_channels
         n_layers_head = len(self.heads["surface_precip"].layers)
         n_features_head = self.heads["surface_precip"].layers[0][0].out_channels
-        print(n_features_head)
         ancillary = self.ancillary
         targets = self.targets
 

### 2021-11-30 gprof_nn/bin/retrieve.py
Add 'preserve_structure' option to retrieve command.

@@ -55,6 +55,12 @@ def add_parser(subparsers):
     parser.add_argument('--gradients', action='store_true')
     parser.add_argument('--no_profiles', action='store_true')
     parser.add_argument(
+        '--preserve_structure',
+        action='store_true',
+        help=("Whether or not to preserve the spatial structure"
+              " of the 1D retrieval on training data.")
+    )
+    parser.add_argument(
         '--n_processes',
         metavar="n",
         type=int,
@@ -77,7 +83,8 @@ def process_file(input_file,
                  targets,
                  gradients,
                  device,
-                 log_queue):
+                 log_queue,
+                 preserve_structure=False):
     """
     Process input file.
 
@@ -102,7 +109,8 @@ def process_file(input_file,
     retrieval = driver(input_file,
                        xrnn,
                        output_file=output_file,
-                       device=device)
+                       device=device,
+                       preserve_structure=preserve_structure)
     retrieval.run()
 
 
@@ -132,6 +140,7 @@ def run(args):
     if not input.is_dir() and not output.exists():
         output.mkdir(parents=True, exist_ok=True)
 
+    preserve_structure = args.preserve_structure
     gradients = args.gradients
     n_procs = args.n_processes
     device = args.device
@@ -152,8 +161,10 @@ def run(args):
         output_files = []
         for f in input_files:
             of = f.relative_to(input)
-            if of.suffix in [".nc", ".nc.gz", ".HDF5"]:
+            if of.suffix in [".nc", ".HDF5"]:
                 of = of.with_suffix(".nc")
+            elif of.suffix in [".nc.gz"]:
+                of = of.with_suffix("")
             else:
                 of = of.with_suffix(".bin")
             output_files.append(output / of)
@@ -187,7 +198,8 @@ def run(args):
                               targets,
                               gradients,
                               device,
-                              log_queue)]
+                              log_queue,
+                              preserve_structure=preserve_structure)]
 
     for t in track(tasks, description="Processing files:"):
         gprof_nn.logging.log_messages()

### 2021-12-01 gprof_nn/definitions.py
Added method to calculate bootstrap mean and std. dev..

@@ -59,7 +59,7 @@ DATABASE_MONTHS = [
 
 TEST_DAYS = [1, 2, 3]
 VALIDATION_DAYS = [4, 5]
-TRAINING_DAYS = list(range(6, 32))
+TRAINING_DAYS = list(range(13, 32))
 
 LIMITS = {
     "brightness_temperatures": (0, 400),

### 2021-12-01 gprof_nn/utils.py

@@ -8,29 +8,10 @@ Collection of utility attributes and functions.
 import numpy as np
 import xarray as xr
 
+from gprof_nn.definitions import SURFACE_TYPE_NAMES
+
 # Rectangular bounding box around continental united states (CONUS).
 CONUS = (-130, 20, -60.0, 55)
-
-SURFACE_TYPE_NAMES = [
-    "Ocean",
-    "Sea-Ice",
-    "Vegetation 1",
-    "Vegetation 2",
-    "Vegetation 3",
-    "Vegetation 4",
-    "Vegetation 5",
-    "Snow 1",
-    "Snow 2",
-    "Snow 3",
-    "Snow 4",
-    "Standing Water",
-    "Land Coast",
-    "Mixed land/ocean o. water",
-    "Ocean or water Coast",
-    "Sea-ice edge",
-    "Mountain Rain",
-    "Mountain Snow",
-]
 
 
 def surface_type_to_name(surface_index):
@@ -196,3 +177,25 @@ def interpolate(variable, weights):
     shape = variable.shape[: weights.ndim] + (1,) * (variable.ndim - weights.ndim)
     weights_r = weights.reshape(shape)
     return np.sum(variable * weights_r, axis=weights.ndim - 1)
+
+def bootstrap_mean(data, n_samples=10):
+    """
+    Calculate mean and standard deviation using boostrapping.
+
+    Args:
+        data: 1D array containing the samples of which to calculate mean
+            and standard deviation.
+        n_samples: The number of bootstrap samples to perform.
+
+    Return:
+        Tuple ``(mu, std)`` containing the estimated mean ``mu`` and
+        corresponding standard deivation ``std``.
+    """
+    stats = []
+    for i in range(n_samples):
+        indices = np.random.randint(0, data.size, size=data.size)
+        stats.append(np.mean(data[indices]))
+    data_r = np.stack(stats)
+    mu = data_r.mean()
+    std = data_r.std()
+    return mu, std

### 2021-12-01 gprof_nn/bin/retrieve.py
Merge.

@@ -121,7 +121,6 @@ def run(args):
     Args:
         args: The namespace object provided by the top-level parser.
     """
-    mp.set_start_method("spawn")
 
     #
     # Check and load inputs.
@@ -144,6 +143,8 @@ def run(args):
     gradients = args.gradients
     n_procs = args.n_processes
     device = args.device
+    if device.startswith("cuda"):
+        mp.set_start_method("spawn")
 
     # Find files and determine output names.
     if input.is_dir():
@@ -161,7 +162,7 @@ def run(args):
         output_files = []
         for f in input_files:
             of = f.relative_to(input)
-            if of.suffix in [".nc", ".HDF5"]:
+            if of.suffix in [".gz", ".nc", ".HDF5"]:
                 of = of.with_suffix(".nc")
             elif of.suffix in [".nc.gz"]:
                 of = of.with_suffix("")

### 2021-12-01 gprof_nn/data/utils.py

@@ -111,7 +111,7 @@ def decompress_scene(scene, targets):
             dims = [d if d != "pixels_center" else "pixels" for d in dims]
             data[v] = (dims, data_r)
         else:
-            data[v] = (scene[v].dims, scene[v])
+            data[v] = (scene[v].dims, scene[v].data)
 
     return xr.Dataset(data)
 

### 2021-12-01 gprof_nn/retrieval.py

@@ -336,6 +336,9 @@ class RetrievalDriver:
             If the input data was successfully loaded the input data
             object is returned. ``None`` otherwise.
         """
+        sensor = self.model.sensor
+        platform = self.model.platform
+        sensor = sensors.get_sensor(sensor, platform)
         # Load input data.
         if self.input_format in [GPROF_BINARY, L1C]:
             input_data = self.model.preprocessor_class(
@@ -346,7 +349,7 @@ class RetrievalDriver:
             if loader_class == NetcdfLoader1D and self.preserve_structure:
                 loader_class = NetcdfLoader1DFull
             input_data = loader_class(
-                self.input_file, normalizer=self.model.normalizer
+                self.input_file, normalizer=self.model.normalizer, sensor=sensor
             )
         return input_data
 
@@ -603,7 +606,7 @@ class NetcdfLoader1D(GPROF_NN_1D_Dataset):
     in NetCDF data format.
     """
 
-    def __init__(self, filename, normalizer, batch_size=16 * 1024):
+    def __init__(self, filename, normalizer, sensor=None, batch_size=16 * 1024):
         """
         Create loader for input data in NetCDF format that provides input
         data for the GPROF-NN 1D retrieval.
@@ -624,6 +627,7 @@ class NetcdfLoader1D(GPROF_NN_1D_Dataset):
             batch_size=batch_size,
             shuffle=False,
             augment=False,
+            sensor=sensor
         )
         self.n_samples = len(self)
         self.scalar_dimensions = ("samples",)
@@ -678,7 +682,7 @@ class NetcdfLoader3D(GPROF_NN_3D_Dataset):
     in NetCDF data format.
     """
 
-    def __init__(self, filename, normalizer, batch_size=32):
+    def __init__(self, filename, normalizer, sensor=None, batch_size=32):
         """
         Create loader for input data in NetCDF format that provides input
         data for the GPROF-NN 3D retrieval.
@@ -698,6 +702,7 @@ class NetcdfLoader3D(GPROF_NN_3D_Dataset):
             batch_size=batch_size,
             shuffle=False,
             augment=False,
+            sensor=sensor,
             input_dimensions=(96, 192),
         )
         self.n_samples = len(self)

### 2021-12-03 gprof_nn/retrieval.py
Fix name of tercile.

@@ -89,7 +89,6 @@ def calculate_padding_dimensions(t):
     d_m = math.ceil(m / 32) * 32 - m
     p_l_m = d_m // 2
     p_r_m = d_m - p_l_m
-
     return (p_l_n, p_r_n, p_l_m, p_r_m)
 
 
@@ -203,6 +202,10 @@ def combine_input_data_2d(dataset, sensor, v_tbs="brightness_temperatures"):
 
         features += [t2m, tcwv, st_1h, am_1h]
 
+    if isinstance(sensor, sensors.CrossTrackScanner):
+        va = dataset["earth_incidence_angle"].data
+        features.insert(1, va[..., np.newaxis])
+
     input_data = np.concatenate(features, axis=-1)
     input_data = input_data.astype(np.float32)
     if input_data.ndim < 4:
@@ -264,7 +267,8 @@ class RetrievalDriver:
         output_file=None,
         device="cpu",
         compress=True,
-        preserve_structure=False
+        preserve_structure=False,
+        sensor=None
     ):
         """
         Create retrieval driver.
@@ -282,6 +286,7 @@ class RetrievalDriver:
             compress: If set to ``True`` NetCDF output will be gzipped.
             preserve_structure: Special option that will ensure that the
                 spatial structure is conserved even for the 1D retrieval.
+            sensor: Optional sensor argument to provide to the NetCDF loader.
         """
         input_file = Path(input_file)
         self.input_file = input_file
@@ -327,6 +332,7 @@ class RetrievalDriver:
             self.output_file = output_file
 
         self.preserve_structure = preserve_structure
+        self.sensor = sensor
 
     def _load_input_data(self):
         """
@@ -346,7 +352,9 @@ class RetrievalDriver:
             if loader_class == NetcdfLoader1D and self.preserve_structure:
                 loader_class = NetcdfLoader1DFull
             input_data = loader_class(
-                self.input_file, normalizer=self.model.normalizer
+                self.input_file,
+                normalizer=self.model.normalizer,
+                sensor=self.sensor
             )
         return input_data
 
@@ -365,7 +373,7 @@ class RetrievalDriver:
         """
         means = {}
         precip_1st_tercile = []
-        precip_3rd_tercile = []
+        precip_2nd_tercile = []
         pop = []
         samples = []
 
@@ -388,7 +396,7 @@ class RetrievalDriver:
                             y_pred=y, quantiles=[0.333, 0.667], key=k
                         )
[…]

### 2021-12-03 gprof_nn/augmentation.py
Change width of GMI input window.

@@ -594,9 +594,8 @@ def get_transformation_coordinates(lats,
     offset = coords_eucl_out[:, height // 2, width // 2]
     coords_rel_out = coords_eucl_out - offset[:, np.newaxis, np.newaxis]
 
-
     center_in = center_out.copy()
-    center_in[1, 0, 0] = get_center_pixel_input(x_i, 48)
+    center_in[1, 0, 0] = get_center_pixel_input(x_i, 56)
 
     input_geometry = Swath(lats, lons)
     coords_center_in = input_geometry.pixel_coordinates_to_euclidean(center_in)

### 2021-12-03 gprof_nn/data/retrieval.py
Changed name of tercile.

@@ -113,7 +113,7 @@ DATA_RECORD_TYPES = np.dtype(
         ("ice_water_path", "f4"),
         ("most_likely_precip", "f4"),
         ("precip_1st_tercile", "f4"),
-        ("precip_3rd_tercile", "f4"),
+        ("precip_2nd_tercile", "f4"),
         ("profile_t2m_index", "i2"),
         ("profile_index", f"{N_SPECIES}i2"),
         ("profile_scale", f"{N_SPECIES}f4"),
@@ -142,7 +142,7 @@ DATA_RECORD_TYPES_PROFILES = np.dtype(
         ("ice_water_path", "f4"),
         ("most_likely_precip", "f4"),
         ("precip_1st_tercile", "f4"),
-        ("precip_3rd_tercile", "f4"),
+        ("precip_2nd_tercile", "f4"),
         ("profile_t2m_index", "i2"),
         ("profile_index", f"{N_SPECIES}i2"),
         ("profile_scale", f"{N_SPECIES}f4"),
@@ -175,7 +175,7 @@ DATA_RECORD_TYPES_SENSITIVITY = np.dtype(
         ("ice_water_path", "f4"),
         ("most_likely_precip", "f4"),
         ("precip_1st_tercile", "f4"),
-        ("precip_3rd_tercile", "f4"),
+        ("precip_2nd_tercile", "f4"),
         ("profile_t2m_index", "i2"),
         ("profile_index", f"{N_SPECIES}i2"),
         ("profile_scale", f"{N_SPECIES}f4"),

### 2021-12-03 gprof_nn/data/training_data.py
Some more fixes.

@@ -52,7 +52,7 @@ _THRESHOLDS = {
 
 _INPUT_DIMENSIONS = {
     "GMI": (96, 128),
-    "MHS": (32, 128)
+    "MHS": (48, 128)
 }
 
 
@@ -155,8 +155,8 @@ def write_preprocessor_file_xtrack(input_data, output_file):
             if new_data.dtype in [np.float32, np.float64]:
                 new_data = np.nan_to_num(new_data, nan=-9999.9)
 
-            if k == "airimass_type":
-                new_data[new_data <= 0] = 1
+            #if k == "airimass_type":
+            #    new_data[new_data <= 0] = 0
             new_dataset[k] = (dims, new_data)
 
     if "nominal_eia" in data.attrs:
@@ -294,8 +294,8 @@ def write_preprocessor_file(input_data, output_file):
             elements = da.data.ravel()[:n_elems]
             if elements.dtype in [np.float32, np.float64]:
                 elements = np.nan_to_num(elements, nan=-9999.9)
-            if k == "airmass_type":
-                elements[elements <= 0] = 1
+            #if k == "airmass_type":
+            #    elements[elements <= 0] = 1
             new_dataset[k] = (dims, elements.reshape(new_shape))
 
     if "nominal_eia" in data.attrs:

### 2021-12-03 gprof_nn/bin/retrieve.py
Add option to pass sensor to netcdf retrieval.

@@ -16,6 +16,7 @@ from quantnn.qrnn import QRNN
 from rich.progress import track
 
 import gprof_nn.logging
+from gprof_nn import sensors
 from gprof_nn.retrieval import RetrievalDriver, RetrievalGradientDriver
 from gprof_nn.definitions import ALL_TARGETS, PROFILE_NAMES
 
@@ -55,6 +56,13 @@ def add_parser(subparsers):
     parser.add_argument('--gradients', action='store_true')
     parser.add_argument('--no_profiles', action='store_true')
     parser.add_argument(
+        '--sensor',
+        type=str,
+        metavar="sensor",
+        help="Name of a sensor object to use to load training data.",
+        default=None
+    )
+    parser.add_argument(
         '--preserve_structure',
         action='store_true',
         help=("Whether or not to preserve the spatial structure"
@@ -84,7 +92,8 @@ def process_file(input_file,
                  gradients,
                  device,
                  log_queue,
-                 preserve_structure=False):
+                 preserve_structure=False,
+                 sensor=None):
     """
     Process input file.
 
@@ -110,7 +119,8 @@ def process_file(input_file,
                        xrnn,
                        output_file=output_file,
                        device=device,
-                       preserve_structure=preserve_structure)
+                       preserve_structure=preserve_structure,
+                       sensor=sensor)
     retrieval.run()
 
 
@@ -130,17 +140,34 @@ def run(args):
     model = Path(args.model)
     if not model.exists():
         LOGGER.error("Given model is not an existing file.")
+        return 1
 
     input = Path(args.input)
     output = Path(args.output)
 
     if not input.exists():
         LOGGER.error("Input must be an existing file or folder.")
+        return 1
 
     if not input.is_dir() and not output.exists():
-        output.mkdir(parents=True, exist_ok=True)
+        if not output.suffix:
+            output.mkdir(parents=True, exist_ok=True)
 
     preserve_structure = args.preserve_structure
+
+    sensor_name = args.sensor
+    if sensor_name is not None:
+        try:
+            sensor = sensors.get_sensor(sensor_name)
+        except KeyError:
+            LOGGER.error(
+                "If provided, sensor must be a valid sensor name not '%s'.",
+                sensor_name
+            )
+            return 1
+    else:
+        sensor = None
+
     gradients = args.gradients
     n_procs = args.n_processes
     device = args.device
@@ -155,6 +182,7 @@ def run(args):
 
[…]

### 2021-12-04 gprof_nn/legacy.py
Added full-profile binary to legacy CLI.

@@ -33,11 +33,14 @@ LOGGER = logging.getLogger(__name__)
 EXECUTABLES = {
     "STANDARD": "GPROF_2021_V1",
     "SENSITIVITY": "GPROF_2020_V1_grads",
-    "PROFILES": "GPROF_2020_V1_prof",
+    "PROFILES": "GPROF_2021_V1_profs",
 }
 
 
-EXECUTABLES_X = {"STANDARD": "GPROF_2020_V1x"}
+EXECUTABLES_X = {
+    "STANDARD": "GPROF_2020_V1x",
+    "PROFILES": "GPROF_2020_V1x_profiles"
+}
 
 
 ANCILLARY_DATA = "/qdata1/pbrown/gpm/ancillary/"
@@ -244,6 +247,7 @@ def run_gprof_training_data(
             batch_size=256 * 2048,
         )
 
+    print("MODE: ", mode)
     results = []
     with TemporaryDirectory() as tmp:
         tmp = Path(tmp)
@@ -261,9 +265,9 @@ def run_gprof_training_data(
                 sensor,
                 configuration,
                 preprocessor_file,
-                mode,
-                profiles,
-                nedts,
+                mode=mode,
+                profiles=profiles,
+                nedts=nedts,
                 robust=True,
             )
 
@@ -329,6 +333,7 @@ def run_gprof_standard(sensor, configuration, input_file, mode, profiles, nedts=
             input_file = output_file
 
         results = execute_gprof(
-            tmp, sensor, configuration, input_file, mode, profiles, nedts, robust=True
+            tmp, sensor, configuration, input_file,
+            mode=mode, profiles=profiles, nedts=nedts, robust=True
         )
     return results

### 2021-12-08 gprof_nn/data/__init__.py
Added functionality to fetch models from a server.

@@ -1,0 +1,82 @@
+"""
+=============
+gprof_nn.data
+=============
+
+The data module provides functionality to dynamically download
+data needed by the module.
+
+Its sub-modules provide functionality to read and process various
+data formats required for the processing of training data.
+"""
+from pathlib import Path
+import shutil
+import urllib
+
+from appdirs import user_config_dir
+
+_DATA_DIR = Path(user_config_dir("pansat", "pansat"))
+_DATA_DIR.mkdir(parents=True, exist_ok=True)
+
+_MODEL_DIR = _DATA_DIR / "models"
+_MODEL_DIR.mkdir(parents=True, exist_ok=True)
+
+_DATA_URL = "http://0.0.0.0:8000/"
+
+
+def get_file(path):
+    """
+    Get file from user's data directory or download it if necessary.
+
+    Args:
+
+        path: Path of the file relative to the root of the data
+            directory.
+
+    Return:
+        A absolute local path pointing to the file.
+    """
+    local_path = _DATA_DIR / path
+    if not local_path.exists():
+        url = _DATA_URL + str(path)
+        with urllib.request.urlopen(url) as response:
+            with open(local_path, "wb") as output:
+                shutil.copyfileobj(response, output)
+
+    return local_path
+
+
+def get_model_path(kind, sensor, configuration):
+    """
+    Get path to local GPROF-NN model.
+
+    Args:
+        kind: The type of the GPROF-NN model ('1D' or '3D')
+        sensor: Sensor object representing the sensor for which
+            to retrieve the model.
+        configuration: The configuration of the model
+            ('ERA5' or 'GANAL').
+
+    Return:
+        Local path to the requested model.
+    """
+    kind = kind.lower()
+    if kind not in ["1d", "2d"]:
+        raise ValueError("'kind' must be one of: '1D', '3D'")
+    configuration = configuration.lower()
+    if configuration not in ["era5", "ganal"]:
+        raise ValueError("'configuration' must be one of: 'ERA5', 'GANAL'")
+
+    sensor_name = sensor.full_name.lower()
+    model_name = f"gprof_nn_{kind}_{sensor_name}_{configuration}.pckl"
+    path = Path("models") / model_name
+
+    try:
+        return get_file(path)
+    except urllib.error.HTTPError:
+        pass
+
+    sensor_name = sensor.sensor_name.lower()
+    model_name = f"gprof_nn_{kind}_{sensor_name}_{configuration}.pckl"
+    path = Path("models") / model_name
+    return get_file(path)

### 2021-12-08 setup.py
Added app_dir package to dependencies.

@@ -14,7 +14,7 @@ setup(
     url="https://github.com/simonpf/gprof_nn",
     author="Simon Pfreundschuh",
     author_email="simon.pfreundschuh@chalmers.se",
-    install_requires=["numpy", "xarray", "torch"],
+    install_requires=["numpy", "xarray", "torch", "appdirs"],
     entry_points = {
         'console_scripts': ['gprof_nn=gprof_nn.bin:gprof_nn'],
     },

### 2021-12-09 gprof_nn/data/__init__.py
Added test for gprof_nn.data module.

@@ -21,7 +21,8 @@ _DATA_DIR.mkdir(parents=True, exist_ok=True)
 _MODEL_DIR = _DATA_DIR / "models"
 _MODEL_DIR.mkdir(parents=True, exist_ok=True)
 
-_DATA_URL = "http://0.0.0.0:8000/"
+
+_DATA_URL = "http://rain.atmos.colostate.edu/gprof_nn/"
 
 
 def get_file(path):
@@ -73,10 +74,18 @@ def get_model_path(kind, sensor, configuration):
 
     try:
         return get_file(path)
-    except urllib.error.HTTPError:
+    except urllib.error.URLError:
         pass
 
     sensor_name = sensor.sensor_name.lower()
     model_name = f"gprof_nn_{kind}_{sensor_name}_{configuration}.pckl"
     path = Path("models") / model_name
-    return get_file(path)
+    try:
+        return get_file(path)
+    except urllib.error.HTTPError:
+        raise ValueError(
+            f"Couldn't find a model for sensor '{sensor.name}' and "
+            f"configuration '{configuration}' neither locally nor on "
+            f"the server. Maybe it doesn't exist yet?"
+        )
+

### 2021-12-09 test/test_data.py

@@ -1,0 +1,19 @@
+"""
+This file tests the 'gprof_nn.data' module.
+"""
+import gprof_nn.data
+from gprof_nn import sensors
+
+def test_get_model_path(tmpdir, monkeypatch):
+    """
+    Test that retrieving a model file from the server works as
+    expected.
+    """
+    path = tmpdir
+    monkeypatch.setattr(gprof_nn.data, "_DATA_DIR", tmpdir)
+    tmpdir.mkdir("models")
+
+    path = gprof_nn.data.get_model_path("1D", sensors.GMI, "ERA5")
+
+    assert path.exists()
+    assert (tmpdir / "models" / "gprof_nn_1d_gmi_era5.pckl").exists()

### 2021-12-11 article/manuscript.tex
Merge.
Manuscript writing.

--- 2021-05-28 article/manuscript.tex
@@
+\RequirePackage[l2tabu,orthodox]{nag} % turn on warnings because of bad style
+\documentclass[a4paper,11pt,bibtotoc]{scrartcl}
+\usepackage{natbib}
+
+\usepackage[utf8]{inputenc}
+
+\usepackage[T1]{fontenc}        % Tries to use Postscript Type 1 Fonts for better rendering
+\usepackage{lmodern}            % Provides the Latin Modern Font which offers more glyphs than the default Computer Modern
+\usepackage{amsmath} % Provides all mathematical commands
+\usepackage{amsfonts} % Provides all mathematical commands
+
+\usepackage{hyperref}           % Provides clickable links in the PDF-document for \ref
+\usepackage{grffile}            % Allow you to include images (like graphicx). Usage: \includegraphics{path/to/file}
+\usepackage{subcaption}
+
+% Allows to set units
+\usepackage[ugly]{units}        % Allows you to type units with correct spacing and font style. Usage: $\unit[100]{m}$ or $\unitfrac[100]{m}{s}$
+
+% Additional packages
+\usepackage{url}                % Lets you typeset urls. Usage: \url{http://...}
+\usepackage{breakurl}           % Enables linebreaks for urls
+\usepackage{xspace}             % Use \xpsace in macros to automatically insert space based on context. Usage: \newcommand{\es}{ESPResSo\xspace}
+\usepackage{xcolor}             % Obviously colors. Usage: \color{red} Red text
+\usepackage{booktabs}           % Nice rules for tables. Usage \begin{tabular}\toprule ... \midrule ... \bottomrule
+
+% Source code listings
+\usepackage{listings}           % Source Code Listings. Usage: \begin{lstlisting}...\end{lstlisting}
+\lstloadlanguages{python}       % Default highlighting set to "python"
+
+\renewcommand{\familydefault}{\sfdefault}
+\newcommand\todo[1]{\textcolor{red}{\textbf{TODO :: }#1}}
+\newcommand\note[1]{\textcolor{green}{\textbf{NOTE :: }#1}}
+
+\begin{document}
+
 \title{GPROF-NN: A neural-network based implementation of the Goddard Profiling Algorithm}
+\date{\today}
 
-\documentclass[12pt]{article}
-
-\usepackage{amsmath}
-\begin{document}
 \maketitle
@@
 \section{Introduction}
-\section{Methods}
 
-This section describes the original GPROF algorithm as well as the two newly
-developed neural-network-based versions.
+The Goddard Profiling Algorithm (GPROF, \citet{kummerow15}) is the operational
+precipitation retrieval algorithm for the passive microwave (PMW) observations
+from the constellation of satellites of the Global Precipitation Measurement
+(GPM, \citet{hou14}) mission, whose objective is to provide global measurements
+of precipitation at a temporal resolution of 3 hours. The precipitation that is
+retrieved using GPROF is used directly by meteorologists and climate scientists
+and serves as input for GPM level 3 retrieval products. The algorithm thus
+constitutes an essential component of the system of global observations that
+enables monitoring of the hydrological cycle for the benefit of science and
+society.
 
-\subsection{The original GPROF algorithm}
+The development of GPROF was originally motivated by the Tropical Rainfall
+Measurement Mission (TRMM, \citet{simpson96}), the precursor of the GPM mission,
+and thus dates back almost 30 years \citep{kummerow94_a, kummerow94_b}. Due to
+the conceptual and computational complexity of simulating PMW observations of
+clouds and precipitation, the algorithm was and remains based on a retrieval
+database consisting of observations and corresponding profiles of hydrometeors
+and precipitation rates. Nonetheless, the algorithm has undergone several
+updates since its conception: Methodologically, the most fundamental
+modification was the introduction of the Bayesian retrieval scheme in
+\citet{kummerow96}, which is used in the algorithm until today. Following this,
+algorithm updates were mostly focused on improving the retrieval database and
+the incorporation of ancillary data into the retrieval. While the first version
+of GPROF still used hand crafted hydrometeor profiles to generate the retrieval
+database, these were soon replaced by profiles from a meso-scale weather model
+\citep{kummerow96}. An important improvement was the replacement of the
+model-derived database by an observationally-generated database for the GPROF
+2010 algorithm \citep{kummerow11, kummerow15}, which helped reduce errors caused
+by misrepresentation of atmospheric states in the database. The 2014 version of
+GPROF \citep{kummerow15} introduced the first fully-parametric version of the
+algorithm, which was designed to be applicable to all sensors of the GPM
+constellation. This version of GPROF became the operational PMW
+precipitation retrieval of the GPM mission.
 
-The original GPROF algorithm is based on a Bayesian formulation of the retrieval
-problem. Given an observation $\mathbf{y}$, the result of the retrieval is
-defined as the posterior distribution $P(\mathbf{x} | \mathbf{y})$ of the
-atmospheric state $\mathbf{x}$.
+This study focuses on the computational method that is used to produce the
+retrieval results from the retrieval database used by GPROF. Since its
+introduction in \citet{kummerow96}, the currently used Bayesian method has not
+received much consideration, mainly because the database and the incorporation
+of ancillary data were deemed to be more relevant for improving the accuracy of
+the retrieval. However, two disadvantages of the current retrieval method have
+become apparent with the introduction of the much larger,
+observationally-generated retrieval databases into the algorithm
+\citep{elsaesser15}: Firstly, the retrieval database must be compressed into
+self-similar clusters to make calculation of the retrieval results sufficiently
+fast. This lossy compression may limit the extent to which the current algorithm
+can benefit from the size and representativeness of observationally generated
+retrieval databases. Secondly, the current retrieval method requires artificial
+inflation of the assumed database uncertainties to produce accurate results.
+Since there is no principled way to calculate these uncertainties, they need to
+be tuned heuristically for each sensor.
 
-To calculate the posterior $P(\mathbf{x} | \mathbf{y})$ the algorithm uses a
-database consisting of pairs
-$(\mathbf{y}_1, \mathbf{x}_1), \ldots (\mathbf{y}_N, \mathbf{x}_N)$
-of observations $\mathbf{y}_i$ and corresponding states
-$\mathbf{x}_i$. Assuming that the database describes the true a priori
-distribution of atmospheric states $\mathbf{x}$ sufficiently well, the expected
-value w.r.t to the posterior distribution $P(\mathbf{x} | \mathbf{x})$ of the
-atmospheric state $\mathbf{x}$ can be approximated using
-\begin{align}
-  \int_{\mathbf{x} } f(\mathbf{x})\: d\mathbf{x} \approx
-  \frac{\sum_i f(\mathbf{x}) P(\mathbf{y}|\mathbf{x}_i)}{\sum_i P(\mathbf{y}|\mathbf{x})}
+With the aim of overcoming these limitations, this study investigates the
+benefits of using the GPROF retrieval database to train a neural network to
+retrieve precipitation and hydrometeor profiles. Since the retrieval database
+has grown to a size of several hundred million entries it is perfectly suited
+for the application of deep neural networks, which scale very well to large
+amounts of data and are capable of learning highly complex relationships from
+them. In addition to this, a neural network based implementation has the
+advantage of allowing the integration of spatial information into the retrieval,
+which is not easily feasible with the current method. This study thus aims to
+answer the following two questions:
+\begin{enumerate}
+\item Can a deep learning based retrieval method with identical inputs improve
+  the accuracy of retrieved surface precipitation and vertical hydrometeor
+  profiles?
+  \item Can the incorporation of spatial information into the retrieval help to
+    improve the retrievals?
+\end{enumerate}
+
+To answer these questions, this study presents two novel, neural network based
+implementations of the GPROF algorithm:
+\begin{description}
+\item[GPROF-NN 1D:] This algorithm uses a fully-connected neural network to
+  retrieve single column hydrometeors and rain rates based on the observed
+  brightness temperature vector. Its inputs and outputs are thus identical to
+  those of the standard GPROF algorithm.
+\item[GPROF-NN 3D:] This algorithm extends the GPROF-NN 1D algorithm by
+  incorporating spatial information into the retrieval. It produces the same
+  output as GPROF and GPROF-NN 1D but processes all observations simultaneously,
+  thus allowing the algorithm to combine information from pixels across the
+  swath.
+\end{description}
+The retrieval performance of the two GPROF-NN implementations is assessed on a
+held-out part of the retrieval database and compared to that of the upcoming
+version of GPROF. The performance is assessed for two sensors of the GPM
+constellation: The GPM Microwave Imager (GMI) and the Microwave Humidity Sounder
+(MHS, \citet{bonsignori07}). In addition to this, a case study is presented that
+compares the retrieved surface precipitation from overpasses of the two sensors
+over Hurricane Harvey.
+
+\section{Data and methods}
+
+This section introduces the retrieval database that is used by the GPROF
+algorithm and serves as training data for the GPROF-NN algorithms. This
+is followed by a description of the current implementation of GPROF as well
+as the implementation of the two GPROF-NN retrievals.
+
+\subsection{The retrieval database}
+
+The GPROF retrieval database consists of the retrieval inputs, i.e. PMW observations
+and corresponding ancillary data, paired with corresponding retrieval outputs,
+i.e. the values of the retrieval quantities. GPROF retrieves surface
+precipitation, profiles and path integrals of other hydrometeors as well as
+latent heating profiles. A listing of all retrieval targets and corresponding
+units is provided in Tab.~\ref{tab:variables}.
+
+\begin{table}
+  \caption{Retrieval quantities in the retrieval database}
+  \label{tab:variables}
+  \centering
+  \begin{tabular}{l|rr}
+    Retrieval variable & Unit & Type\\
+    \hline
+    Surface precipitation & $\unit{mm\ h^{-1}}$ & Scalar \\
+    Convective precipitation & $\unit{mm\ h^{-1}}$ & Scalar \\
+    Cloud water path & $\unit{kg\ m^{-2}}$ & Scalar \\
+    Rain water path & $\unit{kg\ m^{-2}}$ & Scalar \\
+    Ice water path & $\unit{kg\ m^{-2}}$ & Scalar \\
+    Cloud water content & $\unit{g\ m^{-3}}$ & Profile \\
+    Rain water content & $\unit{g\ m^{-3}}$ & Profile \\
+    Snow water content & $\unit{g\ m^{-3}}$ & Profile \\
+    Latent heating & $\unit{K \ h^{-1}}$ & Profile \\
+  \end{tabular}
+\end{table}
+
+Values of the retrieval targets for the database of the upcoming version of
+GPROF are derived from one year (October 2018 to September 2019) of
+hydrometeor-profile retrievals from the GPM combined product \citep{grecu16}.
+This data is complemented with surface precipitation from the currently
+operational MIRS retrieval by adding light precipitation in areas where no echo
+is detected by the GPM Dual-Frequency Precipitation Radar. Observations over
+sea-ice and snow-covered surfaces are handled separately. For sea-ice,
+precipitation is derived from the ERA5 reanalysis \citep{hersbach20}. For
+snow-covered surfaces, precipitation is derived from four years of co-locations
+with the gauge-corrected radar observations from the Multi-Radar Multi-Sensor
+(MRMS) product.
+
+The retrieval targets are then combined with observations from each of the
+sensors of the GPM constellation: For GMI, this is done by simply remapping and
+averaging the retrieval outputs to the footprint of GMIs $23\ \unit{GHz}$
+channel. Since co-locations with other sensors of the constellation are rare or
+may not by available at all, observations for these sensors are simulated and
+the footprint averaging is adapted to the expected sensor resolution.
+
+Furthermore, ancillary data derive from global circulation models is added to
+the database. It comprises two meter temperature ($T_{\text{2m}}$), total column
+water vapor ($\text{TCWV}$), a surface class as well as an air lifting index
+(ALI) that encodes information on atmospheric convergence in mountainous areas.
+Retrieval databases are produced in two configurations for each sensor of the
+GPM constellation: One for near real time applications, which uses forecasts
+from the Japan Meteorological Agency (JMA) Global Analysis (GANAL) to derive
+ancillary data, and one for climatological applications, which uses the ERA5
+reanalysis \citep{hersbach20} to derive the ancillary data. Since this is more
+of an operational detail, this study only uses databases based on ERA5.
+
+\subsection{The GPROF algorithm}
+\label{sec:gprof}
+
+The current implementation of the GPROF retrieval consists of three components:
+(1) The Bayesian method that is used to calculate the retrieval results from
+retrieval inputs, (2) the compression required to handle very large databases
+and (3) the incorporation of ancillary data into the retrieval. An illustration
+of these three components is provided in Fig.~\ref{fig:gprof_legacy}.
+
+\begin{figure}[hbpt!]
+  \centering \includegraphics[width=\textwidth]{../figures/gprof_legacy}
+  \caption{The GPROF retrieval algorithm. Panel (a) illustrates the resampling
+    of database samples based on the Bayesian scheme that is the foundation of
+    the GPROF retrieval. The colored, filled curve on the right side of the plot
+    depicts the approximate PDF derived from the sample weights. Panel (b) shows
+    the clustering of observations that is performed to speed up the calculation
+    of retrieval results. Panel (c) illustrates the incorporation of ancillary
+    data through the application of a stratified retrieval database.}
+  \label{fig:gprof_legacy}
+\end{figure}
+
+The Bayesian scheme that is used in GPROF calculates the retrieval result as the
+posterior distribution $p(\mathbf{x} | \mathbf{y})$ of the observed atmospheric
+state $\mathbf{x}$ conditioned on the retrieval input $\mathbf{y}$, here a
+vector of PMW observations. The database that is used for that consists of pairs
+$(\mathbf{y}_1, \mathbf{x}_1), \ldots (\mathbf{y}_N, \mathbf{x}_N)$ of
+observations $\mathbf{y}_i$ and states $\mathbf{x}_i$. Assuming the database is
+distributed according to the the true a priori $p(\mathbf{x})$, i.e. the
+climatological distribution of the atmospheric state $\mathbf{x}$, the expected
+value of $\mathbf{x}$ w.r.t to the posterior distribution $p(\mathbf{x} |
+\mathbf{x})$ can be approximated using
+\begin{align}\label{eq:gprof_retrieval}
+  \int_{\mathbf{x} } \mathbf{x} p(\mathbf{x} | \mathbf{y})\: d\mathbf{x} =
+  \int_{\mathbf{x} } \mathbf{x}\: \frac{p(\mathbf{y} |
+    \mathbf{x})p(\mathbf{x})}{p(\mathbf{y})}\: d\mathbf{x} \approx \frac{\sum_i
+    \mathbf{x} P(\mathbf{y}|\mathbf{x}_i)}{\sum_i P(\mathbf{y}|\mathbf{x})}
 \end{align}
-The conditional probability of the observation given an atmospheric state
-from the database, $P(\mathbf{y}|\mathbf{x}_i)$, is approximated using a
-Gaussian error model
-\begin{align}
-  P(\mathbf{y}|\mathbf{x}_i) = \frac{1}{\sqrt(2\pi\text{det}(\mathbf{S}^{-1}))}
-  \exp \left \{
-  - \frac{1}{2}
-  (\mathbf{y} - \mathbf{y}_i})^T \mathbf{S}^{-1}  (\mathbf{y} - \mathbf{y}_i)}{2})
-  \right }
+The conditional probability $P(\mathbf{y} | \mathbf{x}_i)$ of the observation
+$\mathbf{y}$ given an atmospheric state from the database is modeled assuming
+unbiased Gaussian errors in the observations 
+  %
+\begin{align}\label{eq:gprof_error}
+  P(\mathbf{y}|\mathbf{x}_i) = \frac{1}{\sqrt{2\pi\text{det}(\mathbf{S}^{-1})}}
+  \exp \left \{ - \frac{1}{2} (\mathbf{y} - \mathbf{y}_i)^T \mathbf{S}^{-1}
+  (\mathbf{y} - \mathbf{y}_i) \right \}
 \end{align}
+using a diagonal covariance matrix $\mathbf{S}$ that descibes the errors due to
+sensor noise as well as other causes of deviations of real observation
+$\mathbf{y}$ from the observations in the database.
+Eq.~\ref{eq:gprof_retrieval} corresponds to a resampling of the states in the
+database with case-specific weights calculated using Eq.~\ref{eq:gprof_error}.
+This can be extended to approximate the probability density function of the
+posterior distribution or to derive probabilities of certain characteristics of
+the a posteriori state such as the presence of precipitation in a given
+observation.
 
-This approach can be extended to also estimate other statistics of the a
-posteriori distribution such as its quantiles.
+With the current size of the retrieval database, including all samples in the
+sum in Eq.~\ref{eq:gprof_retrieval} would make the retrieval prohibitively slow.
+To speed up the calculation observations are grouped into self-similar clusters
+and only mean observation and state as well as the number of observations is
+stored in the database. The clustering is accounted for in
+Eq.~\ref{eq:gprof_retrieval} by multiplying the weight
+$p(\mathbf{y_i}|\mathbf{x_i})$ with the number of profiles in the cluster.
+
+Finally, in order to constrain the retrieval further than what the PMW
+observations can do alone, the sum in Eq.~\ref{eq:gprof_retrieval} is limited to
+profiles from similar meteorological contexts as identified by the ancillary
+data. To this end the database is stratifed with respect to $T_\text{2m},
+\text{TCWV}$, surface type and air lifting index. In addition to further
+constraining the retrieval, this has the benefit of further speeding up the
+retrieval.
+
+\subsection{The GPROF-NN algorithms}
+
+The principal objective that guided the design of the GPROF-NN algorithms was to
+develop a neural network retrieval that provides the exact same output as GPROF
+so that it can potentially replace the current implementation in a future
+update. GPROF produces several outputs that are probabilistic: A probability of
+precipitation as well as the mode and terciles of the posterior distribution of
+precipitation. An implementation based on standard regression neural networks
+would provide no principled way to produce these probabilistic outputs due to
+incompatibility of (deterministic) regression with the Bayesian formulation used
+by GPROF. To overcome this limitation, the implementation of the GPROF-NN
+retrievals is based on quantile regression neural networks (QRNNs). As has been
+shown in \citep{pfreundschuh18}, when trained on a dataset that is distributed
+according to the a priori distribution of a Bayesian retrieval, QRNNs learn to
+the predict quantiles of the Bayesian posterior distribution. They thus provide
+a straight forward way to reconcile neural network retrievals with the Bayesian
+framework that is applied by GPROF.
+
+This approach can be generalized to predict a sequence of quantiles, from the
+cumulative distribution function (CDF) of the a posteriori distribution can be
+reconstructed. Since the distribution of a scalar variable is fully described by
+its CDF, the predicted CDF can be used to derive any relevant statistic of the a
+posteriori distribution. For the GPROF-NN retrievals the predicted CDF is used
+to derive maximum a posteriori and mean surface precipitation (the latter of
+which is identical to the solution that would have been obtained with common
+mean squared error regression), the terciles of the posterior distribution as
+well the probability of precipitation. An illustration this principle of the
+GPROF-NN retrievals is provided in Fig.~\ref{fig:gprof_nn_principle}: The
+retrieval employs a neural network to predict for each input pixel a vector of
+values. The elements of this vector correspond to a specific squence of
+quantiles of the a posteriori distribution. These quantiles are used to
+reconstruct a piece-wise linear approximation of the CDF of the a posteriori
+distribution from which the retrieval outputs are derived.
+
+\begin{figure}[hbpt]
+  \centering
+  \includegraphics[width=1.0\textwidth]{../plots/gprof_nn_principle}
+  \caption{The basic principle of the implementation of the GPROF-NN retrievals.
+    A Bayesian solution of the retrieval is obtained by predicting, for each
+    input pixel, a sequence of quantiles of the a posterior distribution that is
+    used to reconstruct its CDF. The predicted CDF is then used to derive the
+    retrieval results.}
+  \label{fig:gprof_nn_principle}
+\end{figure}
+
+\subsubsection{Training objectives}
+\label{sec:objectives}
+
+A neural network can learn to predict a quantiles $\hat{x}_\tau$ of
+a given conditional distribution by training it to minimize the quantile (or pinball) loss
+$\mathcal{L_\tau}$ function corresponding to the quantile
+fraction $\tau$ \citep{koenker01}:
+\begin{align}\label{eq:quantile_loss}
+  \mathcal{L}_\tau(\hat{x}_\tau, x) &= (\tau - \mathbb{I}_{x < \hat{x}_\tau})(x
+  - \hat{x}_\tau),
+\end{align}
+where $\hat{x}_\tau$ is the predicted quantile, $x$ is a sample of the
+conditional distribution from the training data and $\mathbb{I}$ is the
+indicator function taking the value 1 when $x < \hat{X}$ and 0 otherwise.
+
+This principle can be extended to a sequence of quantiles corresponding to
+quantile fractions $\tau_1, \ldots, \tau_N$ by minimizing the mean of the losses
+functions corresponding to each quantile fraction:
+\begin{align}\label{eq:loss_function}
+  \mathcal{L}_{\tau_1,\ldots, \tau_N}(\hat{\mathbf{x}}, x) &= \frac{1}{N} \sum_i
+  \mathcal{L}_{\tau_i}(\hat{x}_i, x)
+\end{align}
+where $\hat{x}_i$ is the $i$th component of the vector of predicted quantiles
+$\hat{\mathbf{x}}$.
+
+The GPROF-NN retrievals use the loss function Eq.~\ref{eq:loss_function} with
+128 equally spaced quantiles ranging from $\tau_1 = 0.001$ to $\tau_{128} =
+0.999$ for all scalar retrieval variables.
+
+Hydrometeor profiles, however, are handled differently: To reduce the number of
+network outputs, the posterior mean of hydrometeor profiles is predicted
+directly using mean squared error regression. Since the output of GPROF contains
+only the posterior mean of the hydrometeor concentrations, it was deemed
+unnecessary to predict their full posterior distribution at each level using
+quantile regression.
+
+\subsubsection{GPROF-NN 1D}
+
+The GPROF-NN 1D algorithm was designed to be functionally identical to the
+current GPROF algorithm. Just as GPROF, it uses PMW observations from a single
+pixel together with corresponding ancillary data to retrieve corresponding
+precipitation and hydrometeor profiles.
+
+\begin{figure}[hbpt]
+  \centering
+    \includegraphics[width=\textwidth]{../plots/gprof_nn_0d_arch}
+    \caption{Illustration of the neural network architecture used in the
+      GPROF-NN 1D algorithm. The network consists of a common body and
+      one head for each retrieval variable. Each block in body and head
+      consists of fully-connected layer, layer norm and GELU activation function.
+    }
+  \label{fig:gprof_nn_1d}
+\end{figure}
+
+The neural network architecture used for the GPROF-NN 1D retrieval is
+illustrated in Fig.~\ref{fig:gprof_nn_1d}. A single network is trained to
+predict all retrieval variables listed (c.f. Tab.~\ref{tab:variables}) using the
+training objectives described Sec.~\ref{sec:objectives}. The network consists of
+a shared body and a separate head for each retrieved variable. Body and heads
+are built-up of blocks consisting of a fully-connected layer followed by
+layer normalization \citep{ba16} and GELU \citep{hendrycks16} activation
+functions. During development we have experiment with different numbers of blocks
+in body ($N_b$) and each of the heads ($N_h$) but found only marginal impact on
+the retrieval performance and settled for a configuration with $N_b = 6$ and
+$H_h = 4$.
+
+The GPROF-NN 1D network is trained by simultaneously minimizing the sum of the
+losses of all retrieval variables. The training is performed over 60 epochs
+using the Adam optimizer \citep{kingma14} with an initial learning rate of
+$5\cdot10^{-4}$ and a cosine annealing learning rate schedule
+\citep{loshchilov16}. Warm restarts are performed after 10, 20 and 30 epochs.
+
+\subsubsection{GPROF-NN 3D}
+
+The GPROF-NN 3D retrieval extends the GPROF and GPROF-NN 1D algorithms by
+incorporating spatial information into the retrieval. To achieve this,
+the GPROF-NN employs a convolutional neural network (CNN) that performs
+the retrieval for all pixels in the swath simultaneously.
+
+\begin{figure}[hbpt]
+  \centering
+  \includegraphics[width=1.0\textwidth]{../plots/gprof_nn_2d_arch_compl}
+  \caption{
+    The neural network architecture of the GPROF-NN 3D retrieval. Illustration
+    (a) displays the structure of the Xception blocks \citep{chollet17} that
+    form the building blocks of the GPROF-NN 3D model. An Xception block
+    consists of two depthwise separable convolutions followed by a group
+    normalization layer and a GELU activation function. Illustration (b) shows
+    how the Xception blocks are used in an asymmetric encoder-decoder structure
+    that forms the body of the network. Output from the body is combined with
+    the ancillary data to form the inputs to the separate heads that predict the
+    retrieval results for each of the retrieved variables.
+  }
+  \label{fig:gprof_nn_3d_arch}
+\end{figure}
+
+The network architecture for the GPROF-NN 3D algorithm, illustrated in
+Fig.~\ref{fig:gprof_nn_3d_arch}, consists of an asymmetric encoder-decoder
+structure followed by a separate head for each retrieved variable. The stages of
+the en- and decoder are built up of what we refer to here as Xception blocks
+(Fig.~\ref{fig:gprof_nn_3d_arch} (a)) because they are based on the Xception
+architecture introduced in \citet{chollet17}. Each block consists of two
+depthwise separable convolutions with a kernel size of 3 followed by group
+normalization layers with 32 groups and GELU activation functions. The first
+block in each stage of the encoder additionally contains a $3\times3$
+max-pooling layer with a stride of 2 following the first $3\times3$ convolution
+layer. Each downsampling block in the encoder is followed by $N = 4$ standard
+Xception blocks. The stages of the decoder consist of bi-linear upsampling layer
+followed by a single Xception block. The network architecture was chosen in
+order to keep the memory footprint of the model low enough that a moderately
+deep model could still be trained on a single GPU with $24\ \unit{GB}$ of
+memory.
+
+\begin{figure}[hbpt]
+  \centering
+    \includegraphics[width=\textwidth]{../plots/data_augmentation}
+    \caption{
+      The effect of GMIs conical viewing geometry on observed features. Panel
+      (a) displays geolocated observations of the $10.6\ \unit{GHz}$ channel.
+      Grey squares mark equilaterals with a side length of $200\unit{km}$
+      oriented along the swath. The highlighted stripe located at the swath
+      center marks the region where the values of the retrieved variables are
+      available. Panel (b) shows the same observations viewed as an image on a
+      uniform grid. Panel (c) shows six synthetically generated training inputs
+      based on two input regions marked in Panel (b). The first row shows three
+      synthetic samples that simulate the effect of viewing the input in region
+      A at a different position across the GMI swath. The second row shows the
+      corresponding transformations for the input in region B.
+    }
+  \label{fig:data_augmentation}
+\end{figure}
+
+Due the special structure of the GPROF retrieval database, it was necessary to
+adapt the training scheme for the GPROF-NN 3D retrieval. For the most part, the
+retrieval database contains values for the retrieved variables only at the 20
+central pixels of the GMI swath (illustrated by the light stripe in
+Fig~\ref{fig:data_augmentation} (a)). That means that although the network
+predicts the retrieval variables at all pixels in the swath, the loss can be
+calculated only at the pixels where the reference data is available.
+Furthermore, the viewing geometry of PMW sensors breaks the translation symmetry
+that is common for digital images and which is one of the inductive biases of
+CNNs \citep{goodfellow16}. The reason for this is that the distance between and
+orientation of pixels varies across the swathes of the sensors. This is
+illustrated in Fig.~\ref{fig:data_augmentation} for the conical viewing geometry
+of GMI. When displayed on a uniform grid, one and the same shape appears
+differently depending on its position across the swath. To account for this in
+the training, training samples are transformed randomly to simulate the effect
+of seeing a given scene at different locations across the GMI swath. These
+transformations are applied randomly when a training sample is loaded thus
+ensuring that the network rarely or never sees a training scene from the same
+perspective. The transformations were furthermore designed to ensure that the
+relative location of the pixels at which values of retrieval variables are known
+varies across the full width of the swath instead of being located always at its
+center. Examples of the transformed inputs are displayed in
+Fig.~\ref{fig:data_augmentation} (c).
+
+The training of the GPROF-NN 3D algorithm for other sensors than GMI is further
+complicated by the fact that the retrieval database for these sensors contains
+simulated observations only at the central pixels of the GMI swath (the
+highlighted pixels in Fig.~\ref{fig:data_augmentation} (a)). To obtain
+two-dimensional training scenes that are sufficiently wide to train a CNN, we
+make use of an intermediate CNN based model to predict the simulated brightness
+temperatures across the full GMI swath. The extended simulated brightness
+temperatures are then remapped from the GMI viewing geometry to the viewing
+geometry of the target sensor in the same way as outlined in the paragraph
+above.
+
+\section{Results}
+
+The first part of this section presents the the evaluation GPROF and the
+GPROF-NN algorithms on a held-out test data set. The test data set consists of
+the observations from the retrieval database from the first three days of every
+month. It should be noted that we have deliberately limited this evaluation to
+data from the retrieval database in order to isolate the effect of the retrieval
+algorithm from that of the database.
+
+We conclude this section with a case study of overpasses of Hurricane Harvey.
+These results are based on real observations and thus provide an indication to
+what extent the performance on the retrieval database can be expected to
+generalize to real observations.
+
+\subsection{Precipitation and water columns}
+
+As described in Tab.~\ref{tab:variables}, the scalar variables retrieved by
+GPROF are surface and convective precipitation as well as the column-integrated
+concentrations of cloud droplets, rain and snow, denoted as cloud water path
+(CWP), rain water path (RWP) and ice water path (IWP), respectively. Scatter
+plots of the retrieval results for these 5 quantities are displayed in
+Fig.~\ref{fig:results_scatter_gmi} for GMI and
+Fig.~\ref{fig:results_scatter_mhs} for MHS. The frequencies in all plots have
+been normalized column-wise to ensure that results for high  values
+remain visible.
+
+\begin{figure}[hbpt]
+  \centering
+  \includegraphics[width=\textwidth]{../plots/results_scatter_gmi}
+  \caption{
+    Scatter plots of scalar retrieval targets for the three retrieval algorithms
+    for GMI. The row display the results for the GPROF, GPROF-NN 1D and GPROF-NN
+    3D algoriths, respectively. Columns display the results for different
+    retrieval targets. Frequencies in the plots have been normalized
+    column-wise, i.e. per bin of the reference value.
+  }
+  \label{fig:results_scatter_gmi}
+\end{figure}
+
+For surface precipitation retrieved by GMI, consistent improvements in retrieval
+accuracy are observed between the GPROF and GPROF-NN 1D as well as the GPROF-NN
+1D and GPROF-NN 3D algorithms. The improvements are most pronounced for very
+small rain rates between $10^{-2}$ and $10^{-1}\ \unit{mm\ h^{-1}}$ but are
+consistent across the full range of displayed values. For convective
+precipitation, the results of GPROF deviate noticeably from the diagonal.
+Slightly better results are provided by the GPROF-NN 1D algorithm, while the
+GPROF-NN 3D algorithm yields the best agreement with the reference data. For the
+path-integrated quantities GPROF already achieves good agreement with the
+reference values. Nonetheless, small but consistent improvements are observed
+for the GPROF-NN 1D and GPROF-NN 3D algorithms.
+
+\begin{figure}[hbpt]
+  \centering
+  \includegraphics[width=\textwidth]{../plots/results_scatter_mhs}
+  \caption{
+    Same as Fig.~\ref{fig:results_scatter_gmi} but for MHS.
+  }
+  \label{fig:results_scatter_mhs}
+\end{figure}
+
+The results for MHS, displayed in Fig~\ref{fig:results_scatter_mhs}, paint a
+similar picture. Although the overall accuracy of the retrieval is lower, the
+GPROF-NN 1D retrieval consistently yields more accurate results than GPROF. Also
+here the GPROF-NN 3D retrieval yields further, consistent improvements compared
+to the GPROF-NN 1D retrieval.
+
+Quantitative measures of the retrieval accuracy for all scalar retrieval targets
+and the three retrieval algorithms are displayed in Tab.~\ref{tab:metrics_gmi}
+for GMI and Tab.~\ref{tab:metrics_mhs} in the appendix for MHS.
+The tables display bias, mean absolute error (MAE), root mean squared error
+(RMSE) and the mean absolute percentage error (MAPE$_{t}$) for all test samples
+events with a reference value that exceeds a quantity-specific threshold $t$.
+The error metrics confirm the qualitative findings from
+Fig.~\ref{fig:results_scatter_gmi} and \ref{fig:results_scatter_mhs}: The neural
+network implementations outperform GPROF in terms of all considered metrics.
+Moreover, the GPROF-NN 3D algorithm further improves upon the performance of
+the GPROF-NN 1D algorithm. Although the errors are generally larger for MHS, the
+same tendency is observed.
+
+\begin{table}[hbpt!]
+
+  \centering
+  \caption{
+    Selected error metrics and standard deviations of retrieved scalar
+    variables for the three implementations of the GPROF algorithm applied to
+    the GMI sensor. The standard deviations were estimated using bootstrapping
+    and thus provide a measure of the uncertainty due to random sampling of the
+    test data.
+  }
+  \label{tab:metrics_gmi}
+\begin{subtable}{\textwidth} 
+ \caption{Surface precipitation} 
+\begin{tabular}{|c||p{3.5cm}|p{3.5cm}|p{3.5cm}|}
+\hline
+Metric &
+\multicolumn{1}{|c}{GPROF} &
+\multicolumn{1}{|c}{GPROF-NN 1D} &
+\multicolumn{1}{|c|}{GPROF-NN 3D} \\
+\hline\hline
+Bias \hfill [$\unit{mm\ h^{-1}}$] & \hfill $ -0.0091 \pm 0.0001$ &\hfill $ -0.0019 \pm 0.0001$ &\hfill $ -0.0012 \pm 0.0001$ \\
+MAE \hfill [$\unit{mm\ h^{-1}}$] & \hfill $  0.0787 \pm 0.0001$ &\hfill $  0.0579 \pm 0.0000$ &\hfill $  0.0462 \pm 0.0001$ \\
+RMSE \hfill [$\unit{mm\ h^{-1}}$] & \hfill $  0.4049 \pm 0.0078$ &\hfill $  0.3799 \pm 0.0078$ &\hfill $  0.3398 \pm 0.0078$ \\
+MAPE$_{0.1}$ \hfill [$\unit{\%}$] & \hfill $ 64.3838 \pm 0.0681$ &\hfill $ 58.0879 \pm 0.0247$ &\hfill $ 48.1921 \pm 0.0449$ \\
+\hline
+\end{tabular}
+\end{subtable}
+
+\begin{subtable}{\textwidth} 
+ \caption{Convective precipitation} 
+\begin{tabular}{|c||p{3.5cm}|p{3.5cm}|p{3.5cm}|}
+\hline
+Metric &
+\multicolumn{1}{|c}{GPROF} &
+\multicolumn{1}{|c}{GPROF-NN 1D} &
+\multicolumn{1}{|c|}{GPROF-NN 3D} \\
+\hline\hline
+Bias \hfill [$\unit{mm\ h^{-1}}$] & \hfill $ -0.0026 \pm 0.0001$ &\hfill $ -0.0008 \pm 0.0001$ &\hfill $ -0.0004 \pm 0.0000$ \\
+MAE \hfill [$\unit{mm\ h^{-1}}$] & \hfill $  0.0332 \pm 0.0002$ &\hfill $  0.0232 \pm 0.0000$ &\hfill $  0.0228 \pm 0.0001$ \\
+RMSE \hfill [$\unit{mm\ h^{-1}}$] & \hfill $  0.3946 \pm 0.0094$ &\hfill $  0.3550 \pm 0.0094$ &\hfill $  0.3104 \pm 0.0094$ \\
+MAPE$_{0.1}$ \hfill [$\unit{\%}$] & \hfill $ 90.7868 \pm 0.1886$ &\hfill $ 88.3793 \pm 0.1458$ &\hfill $ 82.5167 \pm 0.1474$ \\
+\hline
+\end{tabular}
+\end{subtable}
+
+\begin{subtable}{\textwidth} 
+ \caption{Rain water path} 
+\begin{tabular}{|c||p{3.5cm}|p{3.5cm}|p{3.5cm}|}
+\hline
+Metric &
+\multicolumn{1}{|c}{GPROF} &
+\multicolumn{1}{|c}{GPROF-NN 1D} &
+\multicolumn{1}{|c|}{GPROF-NN 3D} \\
+\hline\hline
+Bias \hfill [$\unit{mm\ h^{-1}}$] & \hfill $  0.0009 \pm 0.0000$ &\hfill $ -0.0004 \pm 0.0000$ &\hfill $ -0.0002 \pm 0.0000$ \\
+MAE \hfill [$\unit{mm\ h^{-1}}$] & \hfill $  0.0167 \pm 0.0000$ &\hfill $  0.0118 \pm 0.0000$ &\hfill $  0.0089 \pm 0.0000$ \\
+RMSE \hfill [$\unit{mm\ h^{-1}}$] & \hfill $  0.0913 \pm 0.0032$ &\hfill $  0.0955 \pm 0.0032$ &\hfill $  0.0702 \pm 0.0032$ \\
+MAPE$_{0.01}$ \hfill [$\unit{\%}$] & \hfill $ 78.5511 \pm 0.1381$ &\hfill $ 69.4666 \pm 0.0849$ &\hfill $ 59.6460 \pm 0.0470$ \\
+\hline
+\end{tabular}
+\end{subtable}
+
+\begin{subtable}{\textwidth} 
+ \caption{Ice water path} 
+\begin{tabular}{|c||p{3.5cm}|p{3.5cm}|p{3.5cm}|}
+\hline
+Metric &
+\multicolumn{1}{|c}{GPROF} &
+\multicolumn{1}{|c}{GPROF-NN 1D} &
+\multicolumn{1}{|c|}{GPROF-NN 3D} \\
+\hline\hline
+Bias \hfill [$\unit{mm\ h^{-1}}$] & \hfill $ -0.0029 \pm 0.0000$ &\hfill $ -0.0003 \pm 0.0000$ &\hfill $  0.0000 \pm 0.0000$ \\
+MAE \hfill [$\unit{mm\ h^{-1}}$] & \hfill $  0.0194 \pm 0.0000$ &\hfill $  0.0121 \pm 0.0000$ &\hfill $  0.0087 \pm 0.0000$ \\
+RMSE \hfill [$\unit{mm\ h^{-1}}$] & \hfill $  0.1126 \pm 0.0034$ &\hfill $  0.1118 \pm 0.0034$ &\hfill $  0.0739 \pm 0.0034$ \\
+MAPE$_{0.01}$ \hfill [$\unit{\%}$] & \hfill $ 76.1349 \pm 0.1828$ &\hfill $ 66.5753 \pm 0.1338$ &\hfill $ 54.5458 \pm 0.0679$ \\
+\hline
+\end{tabular}
+\end{subtable}
+
+\begin{subtable}{\textwidth} 
+ \caption{Cloud water path} 
+\begin{tabular}{|c||p{3.5cm}|p{3.5cm}|p{3.5cm}|}
+\hline
+Metric &
+\multicolumn{1}{|c}{GPROF} &
+\multicolumn{1}{|c}{GPROF-NN 1D} &
+\multicolumn{1}{|c|}{GPROF-NN 3D} \\
+\hline\hline
+Bias \hfill [$\unit{mm\ h^{-1}}$] & \hfill $ -0.0018 \pm 0.0000$ &\hfill $ -0.0001 \pm 0.0000$ &\hfill $  0.0006 \pm 0.0000$ \\
+MAE \hfill [$\unit{mm\ h^{-1}}$] & \hfill $  0.0258 \pm 0.0000$ &\hfill $  0.0155 \pm 0.0000$ &\hfill $  0.0116 \pm 0.0000$ \\
+RMSE \hfill [$\unit{mm\ h^{-1}}$] & \hfill $  0.0483 \pm 0.0031$ &\hfill $  0.0387 \pm 0.0031$ &\hfill $  0.0302 \pm 0.0031$ \\
+MAPE$_{0.01}$ \hfill [$\unit{\%}$] & \hfill $ 48.0013 \pm 0.0309$ &\hfill $ 29.9094 \pm 0.0188$ &\hfill $ 25.0521 \pm 0.0203$ \\
+\hline
+\end{tabular}
+\end{subtable}
+\label{tab:surface_precip_metrics}
+\end{table}
+
+\begin{figure}[hbpt]
+  \centering
+  \includegraphics[width=\textwidth]{../plots/results_surface_types}
+  \caption{
+    Bias, RMSE, MAE and MAPE$_{0.1}$ of the retrieved surface precipitation for
+    w.r.t. surface type and retrieval algorithm. The results for the GMI sensor
+    are displayed in the first column, results for the MHS sensor in the
+    second column. Error bars mark one standard deviation around the mean
+    calculated using bootstrap and thus quantify the uncertainty due to the
+    random sampling of the test data.
+  }
+  \label{fig:results_surface_types}
+\end{figure}
+
+Since the surface type has a considerable effect on the lower-frequency
+observations used in the retrieval, its impact on the retrieval of surface
+precipitation is assessed in Fig.~\ref{fig:results_surface_types}. The figure
+displays the Bias, RMSE, MAE and MAPE for the different surface classes that are
+distinguished in the retrieval. Even when different surface types are considered
+separately, the results of the surface precipitation retrieval show the same
+pattern as the scatter plots in Fig.~\ref{fig:results_scatter_gmi}: The results of
+the GPROF-NN 1D retrieval are more accurate than those of the GPROF algorithm
+and the results of the GPROF-NN 3D algorithm are in turn better than those of
+the GPROF-NN 1D algorithm. These results are mostly consistent across all
+surface types and considered metrics, however for some surface types the
+GPROF-NN 1D algorithms yields lower MAPE than GPROF-NN 3D algorithm.
+
+Fig.~\ref{fig:results_spatial} displays the spatial distribution of bias, RMSE
+and MAPE for the three retrievals in $5\ \unit{^{\circ}} \times
+5\ \unit{^{\circ}}$ boxes. As could be expected from the previous results, the
+magnitudes of the biases of GPROF are considerably larger than for the other two
+algorithms. Furthermore, the GPROF exhibits consistent biases across
+geographical regions such as the tropics in South America and Africa as well as
+east Asia, which is less the case for the neural network algorithms. Although
+spatial distribution of the RMSE mostly reflects the global distribution of
+precipitation, a gradual decrease in RMSE can be observed between the results of
+GPROF and GPROF-NN 1D as well as GPROF-NN 1D and GPROF-NN 3D. More consistent
+patterns are visible in the MAPE: All three retrievals exhibit increased MAPE
+over land surfaces, which likely reflects the decrease in information content
+due to the reduced contrast in the lower frequency channels. Over Ocean, the
+MAPE is generally increased in the sub-tropics and tropics compared to higher
+latitudes. Although these patterns are observed in the results of all
+algorithms, a clear and globally consistent decrease in MAPE can be observed
+between the GPROF, GPROF-NN 1D and GPROF-NN 3D retrievals.
+
+\begin{figure}[hbpt]
+  \centering
+  \includegraphics[width=\textwidth]{../figures/results_spatial}
+  \caption{
+    Spatial distributions of Bias (Column 1), RMSE (Column 2) and MAPE (Column
+    3) error statistics for the GPROF (Row 1), GPROF-NN 1D (Row 2), GPROF-NN 3D
+    (Row 3) algorithms in $5\ \unit{^{\circ}} \times 5\ \unit{^{\circ}} $ grid
+    boxes.
+  }
+  \label{fig:results_spatial}
+\end{figure}
+
+\subsubsection{Predicted retrieval uncertainties and probabilistic rain detection}
+
+In addition to quantitative precipitation estimates, GPROF produces estimates of
+the first and second tercile of the posterior distribution of surface
+precipitation, which provide an uncertainty estimate for the retrieved mean
+surface precipitation, as well as a probabilistic classification of pixels into
+raining and non-raining pixels based on an estimate probability of
+precipitation. Due to their probabilistic nature, similar results can be
+produced using GPROF-NN algorithms. From the predicted quantiles, the terciles
+can be predicted by simply interpolating them to the fractions $\frac{1}{3}$ and
+$\frac{2}{3}$, respectively. The probability of precipitation is calculated by
+using the predicted posterior distribution to calculate the probability of the
+retrieved surface precipitation to be larger than the smallest non-zero rain
+rate in the training data.
+
+\begin{table}
+  \caption{
+    Calibration of the predicted terciles of the posterior distribution of
+    surface precipitation on the test set for the three retrieval algorithms and
+    the GMI and MHS sensors.
+  }
+  \label{tab:terciles}
+  \begin{subtable}{\textwidth} 
+    \caption{GMI} 
+    \centering
+  \begin{tabular}{l|rrrr}
+    \multicolumn{1}{c|}{Tercile} &
+    \multicolumn{1}{c}{Nominal} &
+    \multicolumn{1}{c}{GPROF} &
+    \multicolumn{1}{c}{GPROF-NN 1D} &
+    \multicolumn{1}{c}{GPROF-NN 3d} \\
+    \hline
+    First & 0.333 & 0.371 & 0.339 & 0.336 \\
+    Second & 0.667 & 0.71 & 0.648 & 0.656 \\
+  \end{tabular}
+  \end{subtable}
+  \begin{subtable}{\textwidth} 
+    \caption{MHS} 
+    \centering
+    \begin{tabular}{l|rrrr}
+      \multicolumn{1}{c|}{Tercile} &
+      \multicolumn{1}{c}{Nominal} &
+      \multicolumn{1}{c}{GPROF} &
+      \multicolumn{1}{c}{GPROF-NN 1D} &
+      \multicolumn{1}{c}{GPROF-NN 3d} \\
+      \hline
+      First & 0.333 & 0.392 & 0.339 & 0.336 \\
+      Second & 0.667 & 0.743 & 0.648 & 0.656 \\
+    \end{tabular}
+  \end{subtable}
+
+\end{table}
+
+To assess the accuracy of the uncertainty estimates from GPROF and the GPROF-NN
+algorithms, Tab.~\ref{tab:terciles} lists the calibration, i.e. the frequency
+with which each tercile was larger than the true surface precipitation. Although
+for both sensors and algorithms the observed frequencies are close to the
+nominal frequencies $\frac{1}{3}$ and $\frac{2}{3}$, the GPROF-NN algorithms
+yield calibrations that are closer to the nominal values.
+
+\begin{figure}[hbpt!]
+  \centering
+  \includegraphics[width=0.75\textwidth]{../plots/results_pop}
+  \caption{
+    Calibration (Panel (a)) and receiver-operating characteristic (ROC, Panel(b)) for
+    the predicted probability of precipitation.
+  }
+  \label{fig:results_pop}
+\end{figure}
+
+The quality of the rainin/non-raining classification is assessed in
+Fig.~\ref{fig:results_pop}, which displays the calibration of the predicted
+probability and the receiver-operating characteristic (ROC) curve. Similar as
+for the terciles, the predicted probabilities are fairly well calibrated for all
+algorithms and sensors. However, also here the GPROF-NN algorithms seem to
+yields results slightly closer to the diagonal. The results for the ROC curves
+are analogous: The GPROF-NN 1D algorithm yields better precipitation detection
+than GPROF and the GPROF-NN 3D retrieval in turn yields slightly better
+performance than the 1D version. In terms of classification skill, worse
+performance is achieved for GMI than for MHS by all algorithms.
+
+\subsubsection{Effective resolution}
+
+Next, we aim to assess the impact of the retrieval method on the resolution of
+the retrieved precipitation fields. For this, we adopt the approach from
+\citet{guilloteau17}, who have studied the effective resolution of the previous
+version of GPROF for the GMI and the TRMM Microwave Imager sensors. A
+1-dimensional Haar wavelet decomposition in along-track direction over all 128
+pixel long sequences in the test data is performed To calculate the effective
+resolution. We do not consider observations for different surface types
+separately. Following, \citet{guilloteau17} we examine energy spectra as well as
+correlation coefficients and Nash-Sutcliffe (NS) efficiency of the coefficients
+of the wavelet decomposition for the reference and retrieved surface
+precipitation. The results are displayed in Fig.~\ref{fig:resolution}.
+
+\begin{figure}[hbpt!]
+  \centering \includegraphics[width=\textwidth]{../plots/resolution}
+  \caption{ Spatial variability of retrieved and reference fields. Panel (a)
+    shows the average of the total energy defined as the sum of the squared
+    wavelet coefficient at different length scales for the reference and
+    retrieved surface precipitation fields. Panel (b) shows the correlation
+    coefficient between the coefficients of the reference and the retrieved
+    field. Panel (c) shows the corresponding Nash-Sutcliffe efficiency. }
+  \label{fig:resolution}
+\end{figure}
+
+An obvious difference to the results from \citet{guilloteau17} is that the
+energy spectra of the reference precipitation field is not monotonically
+decreasing. The reason for this is that the reference precipitation field in the
+retrieval database is smoothed using an averaging filter adapted to the
+footprint size of the respective sensor. The GPROF-NN 3D algorithm has the
+highest variability in the retrieved precipitation field, followed by the
+GPROF-NN 1D algorithm and GPROF. However, for all retrievals the variability
+remains lower than that of the reference field. In terms of correlation
+coefficients of the wavelet coefficients at different scales (Panel (b)), the
+GPROF-NN 3D algorithm exhibits the highest correlation with the coefficients of
+the reference field, again followed by the GPROF-NN 1D algorithm and GPROF. The
+same pattern is observed for the NS efficiency. In term of effective resolution,
+defined as the smallest scale at which the NS efficiency exceeds 0.5, the
+GPROF-NN 3D algorithm for the GMI sensor achieves a resolution solution of
+$13.5\ \unit{km}$, which is the distance between consecutive pixels in
+along-track direction and corresponds to the smallest spatial scale that can be
+resolved in this analysis. The effective resolutions for the GPROF-NN 1D
+algorithm is $16.2\ \unit{km}$ and for GPROF $21.4\ \unit{km}$.
+
+\note{I will add the results for MHS for GPROF and GPROF-NN 1D here when these
+  results become available.}
+
+\subsubsection{Profile retrieval variables}
+
+In addition to precipitation fields and path-integrated hydrometeor
+concentrations, GPROF retrieves concentration profiles of rain, snow and cloud
+water as well as profiles of latent heating rates. Error statistics for these
+varaibles are displayed in Fig.~\ref{fig:results_profiles}. Qualitatively, the
+results are similar to those observed for the scalar retrieval variables: The
+neural network retrievals yield significantly lower biases than GPROF with the
+GPROF-NN 3D algorithm performing slightly better than the 1D version. Although
+the relative improvements are smaller, similar patterns are observed for RMSE
+and MAPE. These results are mostly consistent across metrics and retrieval
+variables with the exception of the MAPE of cloud water content, for which no
+improvement for the GPROF-NN 3D over the GPROF-NN 1D retrieval is observed. Also
+here the retrievals based on MHS yield higher errors but the performances of the
+three retrievals remain qualitatively similar to what is observed for GMI.
+
+\begin{figure}[hbpt]
+  \centering
+  \includegraphics[width=\textwidth]{../figures/results_profiles}
+  \caption{
+    Error statistics of the retrieved profile variables. Columns show the
+    errors for the different retrieved variables, whereas rows show 
+    altitude averaged Bias, RMSE and MAPE, respectively.
+  }
+  \label{fig:results_profiles}
+\end{figure}
+
+%\subsubsection{Retrieval results beyond the posterior mean}
+%
+%Most quantitative retrieval results of GPROF correspond to the mean of the a
+%posteriori distribution. This is the case for the surface precipictation,
+%convective precipitation, all path-integrated quantities as well as the
+%retrieved profiles. In its current form, GPROF produces only 5 outputs that make
+%use of information from the posterior distribution beyond the posterior mean:
+%The estimated probability of precipitation, the precipitation flag, the 1st and
+%3rd terciles of the surface precipitation (provided as a measure of uncertainty)
+%and the most likely precipitation.
+%
+%
+%The neural-network version of GPROF predicts the full posterior distribution for
+%scalar retrieval variables and storing only the posterior mean discards a large
+%part of this information. Yet, storing the full posterior distribution would
+%require excessive amounts of disk storage and would likely overwhelm the average
+%user. As a storage-efficient alternative, we propose to include a random sample
+%from the posterior distribution in the retrieval output. Although these samples
+%by construction neglect spatial correlations, the results should be able to
+%truthfully reproduce the distribution of precipitation when used in a
+%climatological analysis.
+%
+%As an illustration of the benefits of randomly-sampled retrieval estimates,
+%Fig.~\ref{fig:results_zonal} displays the zonal distributions of surface
+%precipitation on $1^\circ$-degree bins. The reference distribution of the test
+%data (filled contours) is displayed in the background and the quantiles of the
+%retrieved distributions (shaded lines) are drawn on top. Note that the color
+%scaling has been chosen so that if the retrieved and reference distribution were
+%similar, the shaded lines would lie on top of the boundaries between the filled
+%contours in the background.
+%
+%The first row of panels shows the distributions for the case that the posterior
+%mean is used as the observed precipitation value. In this case, the distribution
+%of the observed values does not match the corresponding distribution of the test
+%data well. The distribution of the observed precipitation is systematically
+%biased towards light precipitation. When instead of the posterior mean random
+%samples are used as precipitation estimates, the agreement between the retrieved
+%and the test distribution is significantly improved. Although e discrepancies
+%remain for the lower quantiles of the distribution, very good agreement is found
+%for the high quantiles of the distribution.
+%
+%  \begin{figure}[hbpt!]
+%    \centering
+%    \includegraphics[width=\textwidth]{../figures/zonal_distributions}
+%    \caption{
+%      Zonal distributions of test data and corresponding retrieved values. Colored, filled contours
+%      in the background show the CDFs of the zonal distribution of precipitation in the test data.
+%      Grey-shaded line display the quantiles of the corresponding retrieved distributions based on
+%      the posterior mean (Row 1) and random samples from the posterior distribution (Row 2).
+%    }
+%    \label{fig:results_zonal}
+%  \end{figure}
+%
+%  For another perspective on these results, Fig.\ \ref{fig:scatter_percentiles}
+%  displays scatter plots of the retrieved and the true 95th and 99th percentiles
+%  of the distributions of surface precipitation on a grid of size
+%  $5\ \unit{^{\circ}} \times 5\ \unit{^{\circ}}$. When the posterior mean is
+%  used to calculate the distribution of surface precipitation, the obtained
+%  percentiles systematically underestimate the true percentiles. This systematic
+%  underestimation is corrected when instead of the posterior mean random samples
+%  are used to calculate the surface precipitation distributions.
+%
+%  \begin{figure}[hbpt!]
+%    \centering
+%    \includegraphics[width=\textwidth]{../figures/scatter_percentiles}
+%    \caption{
+%      95th (Row 1) and 99th (Row 2) percentiles of the distribution
+%      of surface precipitation in $5^\circ \times 5^\circ$ boxes in the
+%      test data scattered against the corresponding retrieved values. Filled
+%      contours in the background show the distribution obtained by using the
+%      posterior mean as retrieved value and colored contours the distribution
+%      that is obtained when a random sample from the posterior distribution
+%      is used as retrieval value. The columns show the results for the
+%      GPROF-NN 0D (Column 1) and the GPROF-NN 3D algorithm  (Column 2).
+%    }
+%    \label{fig:scatter_percentiles}
+%  \end{figure}
+
+\subsection{Case study: Hurricane Harvey}
+
+So far, all of the presented results were based on a test dataset with the same
+statistics as the retrieval database. When the algorithms are applied in
+operations, however, the input data may not necessarily follow the distribution
+of the retrieval database. While for GMI the observations can be expected to be
+consistent with those in the retrieval database, this is not necessarily the
+case for other sensors for which the retrieval database contains mostly
+simulated observations. While a comprehensive analysis of the retrieval
+performance on real observations is outside the scope of this paper, this
+section presents retrieval results from two overpasses over Hurricane Harvey to
+provide an indication as to whether the performance characteristics of the
+retrieval algorithm can be expected to carry over to real observations.
+
+\begin{figure}[hbpt]
+  \centering
+  \includegraphics[width=\textwidth]{../plots/surface_precip_harvey_gmi_ir}
+  \caption{ Retrieved surface precipitation from the GPM combined product
+    (Panels (a), (c)), the GPROF algorithm (Panels (b), (d)), and the neural
+    network versions GPROF-NN 1D (Panels (e), (g)) and GPROF-NN 3D (Panels (f),
+    (h)) in and around Hurricane Harvey on 2017-08-25 at 11:50:00 UTC. Panels
+    (a), (b), (e), (f) display the precipitation on a logarithmic scale over an
+    area spanning $20\ 000\ \unit{km}$ in zonal and meridional directions.
+    Panels (c), (d), (g), (f) display the surface precipitation on a linear
+    color scale in an enlarged area spanning $5\ 000\ \unit{km}$ around the
+    Hurricane. The background show $10.3\ \unit{\mu m}$ brightness temperatures
+    from the GOES 16 observations closest to the overpass.
+  }
+  \label{fig:harvey_gmi}
+\end{figure}
+
+First, we consider an overpass of the GPM Core Observatory over Hurricane Harvey
+that occurred on August 25, 2017 at 11:50 UTC. Retrieved surface precipitation
+is displayed in Fig.~\ref{fig:harvey_gmi}. In addition to the retrieved
+surface precipitation from GPROF and the GPROF-NN algorithms, the figure also
+displays the surface precipitation from the GPM combined product averaged to the
+resolution of the $18.7\ \unit{GHz}$ channel. When considered at log-scale,
+obvious structural differences in the retrieved precipitation are visible: The
+GPROF retrieval produces large swaths of low precipitation that cover most of
+the scene, which are not present in the combined retrieval. These features are
+significantly reduced in the results of GPROF-NN 1D algorithm and even more so
+in the results of the GPROF-NN 3D retrieval. When considered at linear scale,
+all three retrievals capture the global structure of the precipitation but
+tend to overestimate its magnitude compared to the combined retrieval. This
+effect is strongest for GPROF but is also apparent for the neural network
+based retrievals. Nonetheless, both neural-network based retrievals better
+reproduce the fine-scale structure that is visible in the results from
+the combined retrieval.
+
+\begin{figure}[hbpt]
+  \centering
+  \includegraphics[width=\textwidth]{../plots/surface_precip_harvey_mhs}
+  \caption{Surface precipitation retrieved from ground radar, (Panels (a), (c)),
+    and MHS PMW observations using the GPROF algorithm (Panels (b), (d)), and
+    the neural network versions GPROF-NN 1D (Panels (e), (g)) and GPROF-NN 3D
+    (Panels (f), (h)) in Hurricane Harvey on 2017-08-25 at 13:58:00
+    UTC. Panels (a), (b), (e), (f) display the precipitation on a logarithmic
+    scale over an area spanning $20\ 000\ \unit{km}$ in zonal and meridional
+    directions. Panels (c), (d), (g), (f) display the surface precipitation on a
+    linear color scale in an enlarged area spanning $5\ 000\ \unit{km}$ around
+    the Hurricane. In the background the closest true-color composite from GOES
+    16 is shown. }
+  \label{fig:harvey_mhs}
+\end{figure}
+
+Finally, we also present results from an overpass of the MHS sensor on board the
+NOAA-18 satellite over the same storm at 13:58 UTC in
+Fig.~\ref{fig:harvey_mhs}. The figure also presents radar derived
+precipitation rates from NOAA's MRMS product suite as reference data. Similar as
+for GMI, the GPROF algorithms predicts low precipitation rates across large
+parts of the scene and even in areas that are cloud free. Albeit to a lesser
+degree, this seems to be an issue also for the GPROF-NN 1D algorithm. The
+GPROF-NN 3D algorithm is less affected by this issue and also yields better
+agreement with the MRMS precipitation rates over land. Regarding the high rain
+rates at the center of the Hurricane, the overall structure again agrees well
+with the radar measurements. GPROF yields the highest rain rates, followed by
+the GPROF-NN 3D and 1D algorithms. Considering the distance of the radar
+observations from the coast, however, it is unclear how reliable the ground
+radar measurements are.
+
+\section{Discussion}
+
+This study presented two novel, neural network based implementations of the
+GPROF retrieval algorithm and evaluated their performance for the GMI and MHS
+sensors against the current implementation.
+
+\subsection{Retrieval performance}
+
+The presented results show that retrieval accuracy across a wide range of
+metrics (Tab.~\ref{tab:metrics_gmi}, Tab.~\ref{tab:metrics_mhs}) is consistently
+improved by  $10$ to $25\ \unit{\%}$ across most retrieved quantities and
+considered error metrics simply by replacing the current retrieval method with a
+fully-connected neural network that uses the same input as the current GPROF
+algorithm. At the same time the effective resolution of the retrieval increases
+by about $25\ \unit{\%}$ from $21\ \unit{km}$ to $16\ \unit{km}$
+(Fig.~\ref{fig:resolution}). Although both GPROF and the GPROF-NN 1D algorithm
+are based on the same retrieval database and use the information as retrieval
+input, the neural network based retrieval provides more accurate results. We
+suspect that this may be cause by two aspects of the current implementation of
+GPROF: Firstly, the assumption of state independent, uncorrelated Gaussian
+errors that affect the observations and, secondly, the clustering of
+observations to compress the retrieval database. It was found by
+\citet{elsaesser15} that even small errors in the assumed uncertainties can
+cause significant deviations at the pixel level. From the uncertainties that
+affect the observations, only the sensor noise conforms to the Gaussian error
+model employed by the retrieval method. For sensors other than GMI, the
+observations are additionally affected by the modeling error in the simulated
+observations, which is difficult to estimate and unlikely to be well described
+by the Gaussian error model. In practice the uncertainties are additionally
+inflated to account for the sparsity of the retrieval database and the effects
+of the clustering. We therefore hypothesize that the ability of the GPROF-NN 1D
+retrieval to learn to account for these uncertainties from the data directly
+allows it to perform better than the current retrieval scheme even when both
+retrievals limited to the same amount of information.
+
+By further extending the retrieval to incorporate structural information, its
+accuracy can be further improved by another $10$ to $25\ \unit{\%}$
+compared to the baseline retrieval at the same time as the effective resolution
+in along track direction is decreased to its lower limit of $13.5\ \unit{km}$.
+The use of structural information for precipitation retrievals is common
+practice in algorithms based on infrared observations \citep{sorooshian00,
+  hong04} and the potential benefits of CNN based retrieval have been shown in
+\citet{sadeghi19}. Our results indicate that also retrievals based on PMW
+observations can benefit substantially from the incorporation of structural
+information into the retrieval.
+
+\subsection{Limitations}
+
+It is important to consider the limitations of the results presented in this
+study. We have deliberately limited the evaluation of retrieval accuracy to test
+data with the same statistical properties as the training data. This was done to
+isolate the effect of the retrieval method from potential aliasing effects that
+would be introduced by the use of external validation data. The performance
+improvements presented here must therefore be understood as upper limits on the
+improvements that can be realized in operational use. Since the GMI retrieval
+can be trained using real observations, the tightness of this bound depends only
+on the representativeness of the retrieval database and thus the accuracy of the
+sources used to generate it. For other sensors, however, the retrieval algorithm
+is trained using simulated observations. Before these simulated observations can
+be used to train the GPROF-NN 3D algorithm, they have to be first extended to
+the full GMI swath using a dedicated simulator network and then remapped to the
+viewing geometry of the sensor. Since, at least for MHS, the distribution of
+simulated observations were found to not match the real observations very well,
+an additional quantile-matching correction is applied to all observations. For
+those sensors it is thus not at all clear to what extent the potential
+improvements carry over to operational application of the retrieval.
+
+\subsection{Operational considerations}
+
+This study constitutes a first step towards the development of a neural network
+based implementation of GPROF that may potentially replace the current
+implementation. However, some challenges remain in the transition of the
+GPROF-NN algorithms from experimental to operational stage. As described above,
+it is not clear to what extent the benefits of the neural network based
+implementation can be realized in operational use. Furthermore, there are other
+aspects that affect the suitability of a retrieval for operational use. One of
+them is certainly the computational effort required to perform the retrieval.
+For the current neural network implementations of GPROF this is not an issue as
+they in fact run faster than the current implementation even when run on a
+single CPU core (c.f. Fig.~\ref{fig:timings} in the appendix). Another issue,
+however, is the detection of anomalous inputs. The current implementation of
+GPROF has the advantage that anomalous inputs can potentially be detected
+through the missing of suitable matches from the retrieval database. The neural
+network implementation lacks this capability and will produce precipitation
+rates for all inputs that are not otherwise flagged by quality control. That
+being said, it is not clear either how effective the current GPROF
+implementation is at detecting small shifts in the retrieval input.
+
+\subsection{Conclusions}
+
+The results presented in this study demonstrate the potential of the two new
+neural network based implementations of GPROF to improve the accuracy as well as
+the effective resolution of retrievals of precipitation and hydrometeor profiles
+from the PMW observations of the satellites of the GPM constellation. In
+addition to the benefits of simply replacing the current retrieval scheme with a
+neural network, we found similar, additional improvements by incorporating
+spatial information into the retrieval through the use of a CNN based
+implementation.
+
+Nonetheless, the most relevant question remains to be answered: How well can the
+retrievals generalize to real observations and independent validation data? This
+will depend on the accuracy of the retrieval database as well as the realism of
+the simulations that are used to produce the input data for sensors other than
+GMI. While we aim to answer this question in a follow-up study, the results
+presented here have value also in their own right. Should the improvements
+observed here not carry over to operational use, it would point to serious
+issues with the retrieval database.
+
+Should, on the contrary, these improvements carry over to operational use these
+novel GPROF implementations would constitute a simple and cost efficient way to
+improve global measurements of precipitation. With this the algorithms would
+constitute a small step towards improving our understanding of the global
+hydrological cycle and its changes in a warming climate.
+
+%The GPROF-NN algorithms and their performance assessment that were presented in
+%this study indicate that the retrieval accuracy of the PMW precipitation retrievals
+%of the GPM constellation can be improved significantly by replacing the current
+%algorithm with a neural network based implementation. Although the presented
+%results should be interpreted as an upper bound on the true retrieval
+%performance, our results remain encouraging: Replacing the retrieval method has
+%the potential to be a simple and cheap way to improve global precipitation
+%measurements.
+%
+%The next step towards such an upgrade will be to extend the implementation of
+%the GPROF-NN 0D and GPROF-NN 3D algorithms to the other sensors of the GPM
+%constellation and run them operationally along side GPROF. The three methods
+%will then be validated and compared against gauge corrected ground based radar
+%measurements in a follow up study. This is to ensure that the good performance of
+%the neural network based retrieval does indeed carries over to the operational
+%processing.
+%
+%In addition to potential improvements in retrieval accuracy, we presented two
+%novel applications enabled by neural-network based retrievals:
+%
+%\begin{enumerate}
+%\item The inclusion of random samples from the posterior distribution in the
+%  retrieval output is proposed as an efficient way of representing retrieval
+%  uncertainties that helps to improve the representation of extreme events in
+%  the retrieval results.
+%\item Our results show that incorporating structural information into the
+%  retrieval yields larger performance benefits than incorporating ancillary
+%  information. This allows simplifying the retrieval pipeline to the point that
+%  only observations are required as input, which opens up the possibility
+%  letting the user run the retrieval on his own computer, i.e. to perform the
+%  inference 'on the edge'. This may be beneficial for retrievals with very high
+%  resolution, i.e. whose storage requirement are much larger than those of the
+%  L1C data.
+%\end{enumerate}
+%
+%Building on the approach proposed in \citet{pfreundschuh18}, this work presents
+
+%pipelines built on Bayesian retrieval algorithms. We find that the machine
+%learning based retrieval, is easiert to implement, runs faster and performs
+%better than the hand-crafted retrieval algorithm. Although it remains to be seen
+%to which extent this will improve retrievals with respect to independent
+%validation data, it seems reasonable to expect at least incremental improvements
+%in global precipitation measurements as well as interesting opportunities for
+%the novel retrieval applications.
+
+%\subsection{Retrieval performance}
+%
+%Our results show a consistent improvement of the GPROF-NN retrievals over the
+%original GPROF algorithm. This is an encouraging result considering that all
+%three retrievals are based on exactly the same retrieval database, indicating
+%that the GPROF retrievals can be improved simply by switching to a different
+%retrieval method.
+%
+%Nonetheless, this raises the question why the neural network based retrievals
+%perform better than the original method? To investigate this, we have calculated
+%the gradients of the retrieved surface precipitation with respect to the
+%brightness temperatures for the GPROF and GPROF-NN 0D retrievals
+%(Fig.~\ref{fig:sensitivity}). Comparison of the distributions of the gradients
+%reveals that the GPROF-NN retrievals exhibit larger gradients with more
+%variability than the GPROF retrieval. This shows that GPROF is less
+%sensitive to the input observations than the neural network versions.
+%
+%Based on the formulation of GPROF (c.f. ~\ref{sec:gprof}), the retrieval may be
+%viewed as nearest-neighbor regression with a Gaussian kernel and a diagonal
+%covariance matrix. Although the theoretical formulation of the retrieval assumes
+%that the elements of the covariance matrix to correspond to the standard deviations
+%of the random errors in the observations, these errors are in practice inflated
+%to account for sparsity of the retrieval database. Our results suggest that this
+%may limit the ability of the method to reproduce structure of the retrieval
+%manifold leading to a loss of sensitivity.
+%
+%In addition to the improvement in retrieval accuracy between GPROF and the
+%GPROF-NN algorithms, our results showed further improvements in retrieval
+%accuracy are obtained when spatial information is included in the retrieval.
+%Intuitively, this makes sense since a human observer can easily identify
+%characteristic rain structures in PMW imagery. At this point we can not
+%make any statements regarding whether or not the patters learned by the
+%neural network have physical significance but we plan to address this in
+%a follow up study.
+%
+%\subsection{The importance of ancillary data}
+%
+%As described above, GPROF relies on several ancillary inputs to improve
+%retrieval performance. Since the GPROF-NN algorithms were designed to be
+%equivalent to GPROF, they, too, make use of ancillary data. However, the
+%improved performance of the neural network based retrievals raises the question
+%to what extent the inclusion of ancillary data is necessary. To answer this
+%question we have tested two variants (GPROF-NN 0D NA and GPROF-NN 3D NA) of the
+%GPROF-NN 0D and GPROF-NN 3D algorithms that don't make use of ancillary data and
+%assessed their performance.
+%
+%\begin{table}[hbpt!]
+%  \centering
+%  \caption{Accuracy metrics of the surface precipitation retrieval for the
+%    retrievals with and without ancillary data.}
+%  \begin{tabular}{lrrrr}
+%    \toprule
+%        {} &      Bias $[\unit{mm\ h^{-1}}]$ &       MAE $[\unit{mm\ h^{-1}}]$ &       RMSE $[\unit{mm\ h^{-1}}]$ &       MAPE $[\unit{\%}]$ \\
+%        Algorithm   &           &           &           &            \\
+%        \midrule
+%        GPROF       &  0.008458 &  0.079320 & 0.489126  &  66.335999 \\
+%        GPROF-NN 0D & -0.001904 &  0.057892 & 0.380972  &  58.060650 \\
+%        GPROF-NN 0D NA & \textbf{0.000092} &	0.061978 & 	0.379741 & 	60.272976 \\
+%        GPROF-NN 3D & -0.001239 &  \textbf{0.046224} &  0.338950 &  \textbf{48.223190} \\
+%        GPROF-NN 3D NA & -0.000270 & 0.048000 &	 \textbf{0.328201} &	51.544792  \\
+%        \bottomrule
+%  \end{tabular}
+%
+%  \label{tab:surface_precip_metrics_na}
+%\end{table}
+%
+%The results are presented in Tab.~\ref{tab:surface_precip_metrics_na} and in
+%Fig.~\ref{fig:results_scatter_na} - \ref{fig:results_surface_types_na} in the
+%appendix. Although omitting the ancillary data has a negative impact on the
+%retrieval performance, both retrievals still perform better than GPROF.
+%Moreover, the GPROF-NN 3D retrieval without ancillary data performs only
+%marginally worse than the retrieval with ancillary data and better than both
+%variants of the GPROF-NN 0D retrieval.
+%
+%Although this may not have any direct consequences on the operational
+%application of GPROF, it still has interesting consequences for the development
+%of future retrievals. Firstly, this obviously means that incorporating spatial
+%information into a PMW precipitation retrieval improves the retrieval to a
+%larger extent than adding ancillary information from secondary sources without
+%adding latency due to data availability constraints. Secondly, this also opens
+%up the possibility of distributing the retrieval algorithm instead of the
+%retrieval and letting the users run the retrieval themselves. The neural network
+%model for the GPROF-NN 3D algorithm has a size of about $70\ \unit{MB}$,
+%which makes it roughly as large as the level 1C file required to run the
+%retrieval. This use case may be interesting for the development of specialized
+%retrievals that have a lower general interest but require large amounts of
+%storage space. An example would be a high-resolution hydrometeor profile
+%retrieval, which may be of interest for case studies but whose output would be
+%too large to archive operationally.
+%
+%
+%\section{Conclusions}
+%
+%The GPROF-NN algorithms and their performance assessment that were presented in
+%this study indicate that the retrieval accuracy of the PMW precipitation retrievals
+%of the GPM constellation can be improved significantly by replacing the current
+%algorithm with a neural network based implementation. Although the presented
+%results should be interpreted as an upper bound on the true retrieval
+%performance, our results remain encouraging: Replacing the retrieval method has
+%the potential to be a simple and cheap way to improve global precipitation
+%measurements.
+%
+%The next step towards such an upgrade will be to extend the implementation of
+%the GPROF-NN 0D and GPROF-NN 3D algorithms to the other sensors of the GPM
+%constellation and run them operationally along side GPROF. The three methods
+%will then be validated and compared against gauge corrected ground based radar
+%measurements in a follow up study. This is to ensure that the good performance of
+%the neural network based retrieval does indeed carries over to the operational
+%processing.
+%
+%In addition to potential improvements in retrieval accuracy, we presented two
+%novel applications enabled by neural-network based retrievals:
+%
+%\begin{enumerate}
+%\item The inclusion of random samples from the posterior distribution in the
+%  retrieval output is proposed as an efficient way of representing retrieval
+%  uncertainties that helps to improve the representation of extreme events in
+%  the retrieval results.
+%\item Our results show that incorporating structural information into the
+%  retrieval yields larger performance benefits than incorporating ancillary
+%  information. This allows simplifying the retrieval pipeline to the point that
+%  only observations are required as input, which opens up the possibility
+%  letting the user run the retrieval on his own computer, i.e. to perform the
+%  inference 'on the edge'. This may be beneficial for retrievals with very high
+%  resolution, i.e. whose storage requirement are much larger than those of the
+%  L1C data.
+%\end{enumerate}
+%
+%Building on the approach proposed in \citet{pfreundschuh18}, this work presents
+%a way of integrating and reconciling machine learning methods with retrieval
+%pipelines built on Bayesian retrieval algorithms. We find that the machine
+%learning based retrieval, is easiert to implement, runs faster and performs
+%better than the hand-crafted retrieval algorithm. Although it remains to be seen
+%to which extent this will improve retrievals with respect to independent
+%validation data, it seems reasonable to expect at least incremental improvements
+%in global precipitation measurements as well as interesting opportunities for
+%the novel retrieval applications.
+
+\section{Appendix}
+
+\begin{table}[hbpt!]
+  \centering
+  \caption{
+    Same as Tab.~\ref{tab:metrics_gmi} but for the MHS sensor.
+  }
+  \label{tab:metrics_mhs}
+\begin{subtable}{\textwidth} 
+ \caption{Surface precipitation} 
+\begin{tabular}{|c||p{3.5cm}|p{3.5cm}|p{3.5cm}|}
+\hline
+Metric &
+\multicolumn{1}{|c}{GPROF} &
+\multicolumn{1}{|c}{GPROF-NN 1D} &
+\multicolumn{1}{|c|}{GPROF-NN 3D} \\
+\hline\hline
+Bias \hfill [$\unit{mm\ h^{-1}}$] & \hfill $  0.0061 \pm 0.0001$ &\hfill $ -0.0027 \pm 0.0001$ &\hfill $ -0.0020 \pm 0.0001$ \\
+MAE \hfill [$\unit{mm\ h^{-1}}$] & \hfill $  0.0898 \pm 0.0001$ &\hfill $  0.0618 \pm 0.0000$ &\hfill $  0.0502 \pm 0.0001$ \\
+RMSE \hfill [$\unit{mm\ h^{-1}}$] & \hfill $  0.2570 \pm 0.0001$ &\hfill $  0.2032 \pm 0.0001$ &\hfill $  0.1289 \pm 0.0001$ \\
+MAPE$_{0.1}$ \hfill [$\unit{\%}$] & \hfill $ 81.8390 \pm 0.0782$ &\hfill $ 68.5625 \pm 0.0472$ &\hfill $ 59.1714 \pm 0.0668$ \\
+\hline
+\end{tabular}
+\end{subtable}
+
+\begin{subtable}{\textwidth} 
+ \caption{Convective precipitation} 
+\begin{tabular}{|c||p{3.5cm}|p{3.5cm}|p{3.5cm}|}
+\hline
+Metric &
+\multicolumn{1}{|c}{GPROF} &
+\multicolumn{1}{|c}{GPROF-NN 1D} &
+\multicolumn{1}{|c|}{GPROF-NN 3D} \\
+\hline\hline
+Bias \hfill [$\unit{mm\ h^{-1}}$] & \hfill $  0.0068 \pm 0.0001$ &\hfill $ -0.0045 \pm 0.0001$ &\hfill $  0.0027 \pm 0.0002$ \\
+MAE \hfill [$\unit{mm\ h^{-1}}$] & \hfill $  0.0366 \pm 0.0001$ &\hfill $  0.0268 \pm 0.0001$ &\hfill $  0.0225 \pm 0.0002$ \\
+RMSE \hfill [$\unit{mm\ h^{-1}}$] & \hfill $  0.1760 \pm 0.0002$ &\hfill $  0.1459 \pm 0.0002$ &\hfill $  0.0931 \pm 0.0002$ \\
+MAPE$_{0.1}$ \hfill [$\unit{\%}$] & \hfill $ 86.1388 \pm 0.3012$ &\hfill $ 77.8337 \pm 0.1587$ &\hfill $ 93.3863 \pm 0.7875$ \\
+\hline
+\end{tabular}
+\end{subtable}
+
+\begin{subtable}{\textwidth} 
+ \caption{Rain water path} 
+\begin{tabular}{|c||p{3.5cm}|p{3.5cm}|p{3.5cm}|}
+\hline
+Metric &
+\multicolumn{1}{|c}{GPROF} &
+\multicolumn{1}{|c}{GPROF-NN 1D} &
+\multicolumn{1}{|c|}{GPROF-NN 3D} \\
+\hline\hline
+Bias \hfill [$\unit{mm\ h^{-1}}$] & \hfill $  0.0035 \pm 0.0000$ &\hfill $ -0.0000 \pm 0.0000$ &\hfill $  0.0008 \pm 0.0001$ \\
+MAE \hfill [$\unit{mm\ h^{-1}}$] & \hfill $  0.0231 \pm 0.0001$ &\hfill $  0.0150 \pm 0.0000$ &\hfill $  0.0121 \pm 0.0000$ \\
+RMSE \hfill [$\unit{mm\ h^{-1}}$] & \hfill $  0.0176 \pm 0.0000$ &\hfill $  0.0129 \pm 0.0000$ &\hfill $  0.0072 \pm 0.0000$ \\
+MAPE$_{0.01}$ \hfill [$\unit{\%}$] & \hfill $121.7047 \pm 0.2060$ &\hfill $ 86.6833 \pm 0.1344$ &\hfill $ 79.5081 \pm 0.4045$ \\
+\hline
+\end{tabular}
+\end{subtable}
+
+\begin{subtable}{\textwidth} 
+ \caption{Ice water path} 
+\begin{tabular}{|c||p{3.5cm}|p{3.5cm}|p{3.5cm}|}
+\hline
+Metric &
+\multicolumn{1}{|c}{GPROF} &
+\multicolumn{1}{|c}{GPROF-NN 1D} &
+\multicolumn{1}{|c|}{GPROF-NN 3D} \\
+\hline\hline
+Bias \hfill [$\unit{mm\ h^{-1}}$] & \hfill $  0.0055 \pm 0.0001$ &\hfill $ -0.0006 \pm 0.0000$ &\hfill $  0.0008 \pm 0.0001$ \\
+MAE \hfill [$\unit{mm\ h^{-1}}$] & \hfill $  0.0274 \pm 0.0000$ &\hfill $  0.0123 \pm 0.0000$ &\hfill $  0.0112 \pm 0.0001$ \\
+RMSE \hfill [$\unit{mm\ h^{-1}}$] & \hfill $  0.0208 \pm 0.0001$ &\hfill $  0.0105 \pm 0.0001$ &\hfill $  0.0065 \pm 0.0001$ \\
+MAPE$_{0.01}$ \hfill [$\unit{\%}$] & \hfill $109.1287 \pm 0.5944$ &\hfill $ 73.1542 \pm 0.0480$ &\hfill $ 73.5670 \pm 0.2495$ \\
+\hline
+\end{tabular}
+\end{subtable}
+
+\begin{subtable}{\textwidth} 
+ \caption{Cloud water path} 
+\begin{tabular}{|c||p{3.5cm}|p{3.5cm}|p{3.5cm}|}
+\hline
+Metric &
+\multicolumn{1}{|c}{GPROF} &
+\multicolumn{1}{|c}{GPROF-NN 1D} &
+\multicolumn{1}{|c|}{GPROF-NN 3D} \\
+\hline\hline
+Bias \hfill [$\unit{mm\ h^{-1}}$] & \hfill $  0.0051 \pm 0.0000$ &\hfill $  0.0015 \pm 0.0000$ &\hfill $  0.0042 \pm 0.0000$ \\
+MAE \hfill [$\unit{mm\ h^{-1}}$] & \hfill $  0.0293 \pm 0.0000$ &\hfill $  0.0202 \pm 0.0000$ &\hfill $  0.0167 \pm 0.0000$ \\
+RMSE \hfill [$\unit{mm\ h^{-1}}$] & \hfill $  0.0032 \pm 0.0000$ &\hfill $  0.0017 \pm 0.0000$ &\hfill $  0.0013 \pm 0.0000$ \\
+MAPE$_{0.01}$ \hfill [$\unit{\%}$] & \hfill $ 57.3852 \pm 0.0340$ &\hfill $ 42.5588 \pm 0.0266$ &\hfill $ 37.8377 \pm 0.0820$ \\
+\hline
+\end{tabular}
+\end{subtable}
+
+\end{table}
+
+\clearpage
+
+\begin{figure}[hbpt]
+  \centering
+  \includegraphics[width=\textwidth]{../figures/results_spatial}
+  \caption{
+    As \ref{fig:results_spatial} but for MHS.
+  }
+  \label{fig:results_spatial}
+\end{figure}
+
+\begin{figure}[hbpt]
+  \centering
+  \includegraphics[width=\textwidth]{../plots/timings}
+  \caption{
+    Single-core processing times of GPROF and the GPROF-NN algorithms for a
+    single orbit of observations.
+    }
+  \label{fig:timings}
+\end{figure}
+
+%\begin{figure}[hbpt]
+%  \centering
+%  \includegraphics[width=\textwidth]{../figures/sensitivity}
+%  \caption{
+%    Relative gradients with respect to the input observations of the three retrievals.
+%    Each panel shows bar plots representing the distribution of the relative gradients
+%    for each of the 13 channels and 4 different amounts of surface precipitation (RR).
+%    Colored boxes display the interquartile range (IQR), whiskers extent to outermost points that
+%    lie within a distance $1.5 \cdot \text{IQR}$ of the limits of the IQR, outliers are
+%    not displayed. Note the difference in y-axis range between the results for GPROF and
+%    the GPROF-NN retrievals.
+%    }
+%  \label{fig:sensitivity}
+%\end{figure}
+
+%\begin{figure}[hbpt]
+%  \centering
+%  \includegraphics[width=\textwidth]{../figures/results_scatter_na}
+%  \caption{
+%    Like Fig.~\ref{fig:results_scatter} but comparing the GPROF-NN algorithms
+%    with (GPROF-NN 0D, GPROF-NN 3D) and without ancillary data (GPROF-NN 0D NA,
+%    GPROF-NN 3D NA).
+%  }
+%  \label{fig:results_scatter_na}
+%\end{figure}
+%
+%\begin{figure}[hbpt]
+%  \centering
+%  \includegraphics[width=\textwidth]{../figures/results_surface_types_na}
+%  \caption{
+%    Like Fig.~\ref{fig:results_surface_types} but comparing the GPROF-NN algorithms
+%    with (GPROF-NN 0D, GPROF-NN 3D) and without ancillary data (GPROF-NN 0D NA,
+%    GPROF-NN 3D NA).
+%  }
+%  \label{fig:results_surface_types_na}
+%\end{figure}
+
+\clearpage
+
+\bibliographystyle{plainnat}
+\bibliography{references}
 

### 2021-12-13 test/test_augmentation.py
Refactored augmentation module.

@@ -7,13 +7,15 @@ import numpy as np
 import xarray as xr
 
 from gprof_nn import sensors
-from gprof_nn.augmentation import (Swath,
-                                   get_center_pixels,
-                                   get_transformation_coordinates,
-                                   offset_x,
-                                   get_center_pixel_input,
-                                   M,
-                                   N)
+from gprof_nn.augmentation import (
+    Swath,
+    get_center_pixels,
+    get_transformation_coordinates,
+    get_center_pixel_input,
+    M,
+    N
+)
+from gprof_nn.data.training_data import decompress_and_load
 
 
 def test_gmi_geometry():
@@ -50,8 +52,8 @@ def test_swath_geometry():
     geometry are reversible.
     """
     path = Path(__file__).parent
-    input_file = path / "data" / "gprof_nn_mhs_era5_5.nc"
-    input_data = xr.load_dataset(input_file)
+    input_file = path / "data" / "mhs" / "gprof_nn_mhs_era5.nc.gz"
+    input_data = decompress_and_load(input_file)
 
     lats = input_data.latitude.data[0]
     lons = input_data.longitude.data[0]
@@ -96,8 +98,8 @@ def test_transformation_coordinates():
     center of the swath.
     """
     path = Path(__file__).parent
-    input_file = path / "data" / "gprof_nn_mhs_era5_5.nc"
-    input_data = xr.load_dataset(input_file)
+    input_file = path / "data" / "mhs" / "gprof_nn_mhs_era5.nc.gz"
+    input_data = decompress_and_load(input_file)
 
     lats = input_data.latitude.data[0]
     lons = input_data.longitude.data[0]
@@ -106,4 +108,4 @@ def test_transformation_coordinates():
         lats, lons, geometry, 64,
         64, 0.5, 0.5, 0.5
     )
-    assert np.all(np.isclose(c[1, 0, :], np.arange(110 - 32, 110 + 32), atol=0.5))
+    assert np.all(np.isclose(c[1, 0, :], np.arange(110 - 32, 110 + 32), atol=2.0))

### 2021-12-13 test/test_preprocessor.py
Refactoring preprocessor module.

@@ -9,14 +9,15 @@ import pytest
 from quantnn.normalizer import Normalizer
 
 from gprof_nn import sensors
-from gprof_nn.data.preprocessor import (PreprocessorFile,
-                                        has_preprocessor,
-                                        run_preprocessor,
-                                        calculate_frozen_precip,
-                                        ERA5,
-                                        GANAL)
-from gprof_nn.data.training_data import (GPROF_NN_0D_Dataset,
-                                         write_preprocessor_file)
+from gprof_nn.data.preprocessor import (
+    PreprocessorFile,
+    has_preprocessor,
+    run_preprocessor,
+    calculate_frozen_precip,
+    ERA5,
+    GANAL,
+)
+from gprof_nn.data.training_data import GPROF_NN_1D_Dataset, write_preprocessor_file
 from gprof_nn.data.l1c import L1CFile
 
 
@@ -62,7 +63,7 @@ def test_read_preprocessor_mhs():
     Tests reading a GMI preprocessor file.
     """
     path = Path(__file__).parent
-    input_file = PreprocessorFile(path / "data" / "mhs" / "MHS.pp")
+    input_file = PreprocessorFile(path / "data" / "mhs" / "pp" / "MHS.pp")
     input_data = input_file.to_xarray_dataset()
 
     assert input_file.n_pixels == 90
@@ -100,23 +101,25 @@ def test_write_preprocessor_file(tmp_path):
     path = Path(__file__).parent
     input_file = path / "data" / "gmi" / "gprof_nn_gmi_era5.nc"
 
-    targets = ["surface_precip",
-               "ice_water_path",
-               "rain_water_path",
-               "cloud_water_path",
-               "latent_heat",
-               "rain_water_content",
-               "snow_water_content",
-               "cloud_water_content"]
+    targets = [
+        "surface_precip",
+        "ice_water_path",
+        "rain_water_path",
+        "cloud_water_path",
+        "latent_heat",
+        "rain_water_content",
+        "snow_water_content",
+        "cloud_water_content",
+    ]
 
-    dataset = GPROF_NN_0D_Dataset(input_file,
-                             batch_size=1,
[…]

### 2021-12-13 gprof_nn/coordinates.py
Removed unused module.

@@ -1,52 +1,0 @@
-"""
-====================
-gprof_nn.coordinates
-====================
-
-Module providing helper function to convert between coordinate
-systems.
-"""
-import numpy as np
-
-try:
-    from pyproj import Transformer
-
-    LATLON_TO_ECEF_TRANSFORMER = Transformer.from_crs("epsg:4326", "epsg:4978")
-except:
-    raise ModuleNotFoundError(
-        """
-        This functionality of the 'gprof_nn' package requires the 'pyproj'
-        package to be installed on your system.
-        """
-    )
-    LATLON_TO_ECEF_TRANSFORMER = None
-
-
-def latlon_to_ecef(longitudes, latitudes, altitudes=None):
-    """
-    Convert latitudes and longitudes to Earth-centered earth-fixed (ECEF)
-    euclidean coordinates.
-
-    Args:
-        longitudes: Array containing the longitudes which to transform
-            to ECEF coordinates
-        longitudes: Array containing the latitudes which to transform
-            to ECEF coordinates
-        altitudes: Optional array of altitudes corresponding to the given
-            longitudes and latitudes. If not provided an altitude of
-            of 0 is assumed for all points.
-
-    Returns:
-        Tuple ``(x, y, z)`` consisting of the arrays ``x``, ``y``, ``z``
-        with each holding the corresponding component of the converted
-        3D coordinates.
-    """
-    longitudes = np.array(longitudes)
-    latitudes = np.array(latitudes)
-
-    if altitudes is None:
-        altitudes = np.zeros(latitudes.shape)
-    else:
-        altitudes = np.array(altitudes)
-
-    return LATLON_TO_ECEF_TRANSFORMER.transform(latitudes, longitudes, altitudes)

### 2021-12-13 test/test_coordinates.py

@@ -1,16 +1,0 @@
-"""
-Test for the coordinates submodule.
-"""
-import numpy as np
-from gprof_nn.coordinates import latlon_to_ecef
-
-def test_latlon_to_ecef():
-
-    lons = [0, 90, 180, 270, 360]
-    lats = [0, 0, 0, 0, 0]
-
-    x, y, z = latlon_to_ecef(lons, lats)
-
-    assert np.all(np.isclose(x[[1, 3]], 0.0))
-    assert np.all(np.isclose(y[[0, 2, 4]], 0.0))
-    assert np.all(np.isclose(z, 0.0))

### 2021-12-13 gprof_nn/coordinates.py
Restored files.

@@ -1,0 +1,52 @@
+"""
+====================
+gprof_nn.coordinates
+====================
+
+Module providing helper function to convert between coordinate
+systems.
+"""
+import numpy as np
+
+try:
+    from pyproj import Transformer
+
+    LATLON_TO_ECEF_TRANSFORMER = Transformer.from_crs("epsg:4326", "epsg:4978")
+except:
+    raise ModuleNotFoundError(
+        """
+        This functionality of the 'gprof_nn' package requires the 'pyproj'
+        package to be installed on your system.
+        """
+    )
+    LATLON_TO_ECEF_TRANSFORMER = None
+
+
+def latlon_to_ecef(longitudes, latitudes, altitudes=None):
+    """
+    Convert latitudes and longitudes to Earth-centered earth-fixed (ECEF)
+    euclidean coordinates.
+
+    Args:
+        longitudes: Array containing the longitudes which to transform
+            to ECEF coordinates
+        longitudes: Array containing the latitudes which to transform
+            to ECEF coordinates
+        altitudes: Optional array of altitudes corresponding to the given
+            longitudes and latitudes. If not provided an altitude of
+            of 0 is assumed for all points.
+
+    Returns:
+        Tuple ``(x, y, z)`` consisting of the arrays ``x``, ``y``, ``z``
+        with each holding the corresponding component of the converted
+        3D coordinates.
+    """
+    longitudes = np.array(longitudes)
+    latitudes = np.array(latitudes)
+
+    if altitudes is None:
+        altitudes = np.zeros(latitudes.shape)
+    else:
+        altitudes = np.array(altitudes)
+
+    return LATLON_TO_ECEF_TRANSFORMER.transform(latitudes, longitudes, altitudes)

### 2022-07-11 gprof_nn/bin/__init__.py
Fixed spelling.

@@ -3,9 +3,9 @@
 gprof_nn.bin
 ============
 
-This sub-module implements the top-level 'gprof_nn' command line applitcation.
-The top-level implementation delegates the processing to the sub-commands
-that are defined in the sub-module of the 'gprof_nn.bin' module.
+This sub-module implements the top-level 'gprof_nn' command line application.
+Its task is to delegate the processing to the sub-commands that are defined
+in the sub-module of the 'gprof_nn.bin' module.
 """
 import argparse
 import sys

### 2022-07-11 gprof_nn/data/mrms.py
Fix definition of dataset variables.

@@ -345,14 +345,8 @@ def resample_to_swath(mrms_data, sensor, l1c_data):
     )
 
     results = xr.Dataset()
-    results["latitude"] = (
-        ("along_track", "across_track"),
-        lats_5
-    )
-    results["longitude"] = (
-        ("along_track", "across_track"),
-        lons_5
-    )
+    results["latitude"] = lats_5
+    results["longitude"] = lons_5
 
     if "precip_rate" in mrms_data.variables:
         # Smooth and interpolate surface precip
@@ -378,10 +372,7 @@ def resample_to_swath(mrms_data, sensor, l1c_data):
             method="nearest",
             kwargs={"fill_value": np.nan}
         )
-        results["surface_precip"] = (
-            ("along_track", "across_track"),
-            surface_precip
-        )
+        results["surface_precip"] = (("along_track", "across_track"), surface_precip.data)
         surface_precip_avg = smooth_reference_field(
             sensor,
             surface_precip.data,
@@ -401,7 +392,7 @@ def resample_to_swath(mrms_data, sensor, l1c_data):
         )
         results["radar_quality_index"] = (
             ("along_track", "across_track"),
-            rqi
+            rqi.data
         )
 
     results.attrs["sensor"] = sensor.sensor_name

### 2022-07-12 article/gprof_nn_manuscript_revised.tex
First round of peer review.

%% Copernicus Publications Manuscript Preparation Template for LaTeX Submissions
%% ---------------------------------
%% This template should be used for copernicus.cls
%% The class file and some style files are bundled in the Copernicus Latex Package, which can be downloaded from the different journal webpages.
%% For further assistance please contact Copernicus Publications at: production@copernicus.org
%% https://publications.copernicus.org/for_authors/manuscript_preparation.html

%% Please use the following documentclass and journal abbreviations for preprints and final revised papers.

%% 2-column papers and preprints
\documentclass[journal abbreviation, manuscript]{copernicus}

%% Journal abbreviations (please use the same for preprints and final revised papers)

% Advances in Geosciences (adgeo)
% Advances in Radio Science (ars)
% Advances in Science and Research (asr)
% Advances in Statistical Climatology, Meteorology and Oceanography (ascmo)
% Annales Geophysicae (angeo)
% Archives Animal Breeding (aab)
% Atmospheric Chemistry and Physics (acp)
% Atmospheric Measurement Techniques (amt)
% Biogeosciences (bg)
% Climate of the Past (cp)
% DEUQUA Special Publications (deuquasp)
% Drinking Water Engineering and Science (dwes)
% Earth Surface Dynamics (esurf)
% Earth System Dynamics (esd)
% Earth System Science Data (essd)
% E&G Quaternary Science Journal (egqsj)
% European Journal of Mineralogy (ejm)
% Fossil Record (fr)
% Geochronology (gchron)
% Geographica Helvetica (gh)
% Geoscience Communication (gc)
% Geoscientific Instrumentation, Methods and Data Systems (gi)
% Geoscientific Model Development (gmd)
% History of Geo- and Space Sciences (hgss)
% Hydrology and Earth System Sciences (hess)
% Journal of Bone and Joint Infection (jbji)
% Journal of Micropalaeontology (jm)
% Journal of Sensors and Sensor Systems (jsss)
% Magnetic Resonance (mr)
% Mechanical Sciences (ms)
% Natural Hazards and Earth System Sciences (nhess)
% Nonlinear Processes in Geophysics (npg)
% Ocean Science (os)
% Polarforschung - Journal of the German Society for Polar Research (polf)
% Primate Biology (pb)
% Proceedings of the International Association of Hydrological Sciences (piahs)
% Safety of Nuclear Waste Disposal (sand)
% Scientific Drilling (sd)
% SOIL (soil)
% Solid Earth (se)
% The Cryosphere (tc)
% Weather and Climate Dynamics (wcd)
% Web Ecology (we)
% Wind Energy Science (wes)

%% \usepackage commands included in the copernicus.cls:
%\usepackage[german, english]{babel}
%\usepackage{tabularx}
%\usepackage{cancel}
%\usepackage{multirow}
%\usepackage{supertabular}
%\usepackage{algorithmic}
%\usepackage{algorithm}
%\usepackage{amsthm}
%\usepackage{float}
%\usepackage{subfig}
%\usepackage{rotating}

\begin{document}

\title{GPROF-NN: A neural network based implementation of the Goddard Profiling Algorithm}

% \Author[affil]{given_name}{surname}

\Author[1, 2]{Simon}{Pfreundschuh}
\Author[2]{Paula J.}{Brown}
\Author[2]{Christian D.}{Kummerow}
\Author[1]{Patrick}{Eriksson}
\Author[3]{Teodor}{Norrestad}

\affil[1]{Department of Space, Earth and Environment, Chalmers University of Technology, 41296 Gothenburg, Sweden}
\affil[2]{Department of Atmospheric Science, Colorado State University, Fort Collins, CO 80523}
\affil[3]{Previously Department of Space, Earth and Environment, Chalmers University of Technology, 41296 Gothenburg, Sweden}

%% The [] brackets identify the author with the corresponding affiliation. 1, 2, 3, etc. should be inserted.

%% If an author is deceased, please mark the respective author name(s) with a dagger, e.g. "\Author[2,$\dag$]{Anton}{Smith}", and add a further "\affil[$\dag$]{deceased, 1 July 2019}".

%% If authors contributed equally, please mark the respective author names with an asterisk, e.g. "\Author[2,*]{Anton}{Smith}" and "\Author[3,*]{Bradley}{Miller}" and add a further affiliation: "\affil[*]{These authors contributed equally to this work.}".

\correspondence{Simon Pfreundschuh (simon.pfreundschuh@chalmers.se)}

\runningtitle{A neural network based implementation of the Goddard Profiling Algorithm}

\runningauthor{Simon Pfreundschuh}

\received{}
\pubdiscuss{} %% only important for two-stage journals
\revised{}
\accepted{}
\published{}

%% These dates will be inserted by Copernicus Publications during the typesetting process.

\firstpage{1}

\maketitle

\begin{abstract}

The Global Precipitation Measurement (GPM) mission measures global precipitation
at a temporal resolution of a few hours to enable close monitoring of the global
hydrological cycle. GPM achieves this by combining observations from a
space-borne precipitation radar, a constellation of passive microwave (PMW)
sensors and geostationary satellites. The Goddard Profiling Algorithm (GPROF) is
used operationally to retrieve precipitation from all PMW sensors of the GPM
constellation. Since the resulting precipitation rates serve as input for many
of the level 3 retrieval products, GPROF constitutes an essential component
of the GPM processing pipeline.

This study investigates ways to improve GPROF using modern machine learning
methods. We present two neural network based, probabilistic implementations of
GPROF: GPROF-NN 1D, which, just as the current GPROF implementation, processes
pixels individually, and GPROF-NN 3D, which employs a convolutional neural
network to incorporate structural information into the retrieval. The accuracy
of the retrievals is evaluated using a test dataset consistent with the data
used in the development of the GPROF and GPROF-NN retrievals. This allows
assessing the accuracy of the retrieval method isolated from the
representativeness of the training data, which remains a major source of
uncertainty in the development of precipitation retrievals. Despite using the
same input information as GPROF, the GPROF-NN 1D retrieval improves the accuracy of the
retrieved surface precipitation for the GPM Microwave Imager (GMI) from
$0.079\ \unit{mm h^{-1}}$ to $0.059\ \unit{mm h^{-1}}$ in terms of
mean absolute error (MAE), from $76.1 \ \unit{\%}$ to
$69.5\ \unit{\%}$ in terms of symmetric mean absolute percentage error (SMAPE)
and from $0.797$ to $0.847$ in terms of correlation. The improvements for the Microwave
Humidity Sounder (MHS) are from $0.085\ \unit{mm h^{-1}}$ to $0.061\ \unit{mm h^{-1}}$ in terms
of MAE, from  $81\ \unit{\%}$ to $70.1\ \unit{\%}$ 
for SMAPE and from $0.724$ to $0.804$ in terms of correlation. Comparable improvements are
found for the retrieved hydrometeor profiles and  their column integrals as well
as the detection of precipitation. Moreover, the ability of the retrievals to
resolve small-scale variability is improved by more than $40\ \unit{\%}$ for
GMI and $29\ \unit{\%}$ for MHS. The GPROF-NN 3D retrieval further improves
the  MAE  to $0.043\ \unit{mm h^{-1}}$, the SMAPE to $48.67\ \unit{\%}$
  and the correlation to $0.897$ for GMI and $0.043\ \unit{mm h^{-1}}$,
  $63.42\ \unit{\%}$ and $0.83$ for MHS.

Application of the retrievals to GMI observations of hurricane Harvey shows
moderate improvements when compared to co-located GPM combined and ground-based
radar measurements indicating that the improvements at least partially carry
over to assessment against independent measurements. Similar retrievals for MHS
do not show equally clear improvements, leaving the validation against independent
measurements for future investigation.

Both GPROF-NN algorithms make use of the same input and output data as the
original GPROF algorithm, and may thus replace the current implementation in a
future update of the GPM processing pipeline. Despite their superior accuracy,
the single-core runtime required for the operational processing of an orbit of
observations is lower than that of GPROF. The GPROF-NN algorithms promise to be
a simple and cost-efficient way to increase the accuracy of the PMW
precipitation retrievals of the GPM constellation and thus improve the
monitoring of the global hydrological cycle.

%The novel GPROF-NN algorithms presented here were designed to be functionally
%equivalent to the current implementation, which enables the neural network
%approach to replace GPROF Bayesian scheme in a future update. Their training
%framework was designed in a modular way so that it can be extended to all
%sensors of the GPM constellation. Despite their superior retrieval accuracy, the
%single CPU core runtime required for the operational processing of an orbit of
%observations is lower than that of GPROF. The GPROF-NN algorithms thus promise
%to be a simple and cost effective way to improve the accuracy of the PMW
%precipitation retrievals of the GPM constellation and thus help improve the
%monitoring of the global hydrological cycle. Together with this study we release
%the implementation of the training framework and operational retrieval as a
%publicly available software package which includes all trained neural network
%models.
\end{abstract}

\introduction  %% \introduction[modified heading if necessary]

The Goddard Profiling Algorithm (GPROF, \citet{kummerow15}) is the operational
precipitation retrieval algorithm for the passive microwave (PMW) observations
from the radiometer constellation of the Global Precipitation Measurement (GPM,
\citet{hou14}), whose objective is to provide consistent global measurements of
precipitation at a temporal resolution of a few hours. The precipitation
retrieved by GPROF serves as input for the Integrated Multi-Satellite Retrievals
for GPM (IMERG, \citeauthor{huffman20}, \citeyear{huffman20}), which can be
considered the state-of-the-art of global precipitation measurements. The
algorithm thus constitutes an essential component of the global observation
system that enables to monitoring the hydrological cycle for the benefit of
science and society.

The development of GPROF was originally motivated by the Tropical Rainfall
Measurement Mission (TRMM, \citet{simpson96}), the precursor of the GPM mission,
and thus dates back almost 30 years \citep{kummerow94_a, kummerow94_b}. Due to
the conceptual and computational complexity of simulating PMW observations of
clouds and precipitation, the algorithm was and remains based on a retrieval
database consisting of  observations and corresponding
profiles of hydrometeors and precipitation rates. Nonetheless, the algorithm has
undergone several updates since its conception: Methodologically, the most
fundamental modification was the introduction of the Bayesian retrieval scheme
in \citet{kummerow96}, which is used in the algorithm until today. Following
this, algorithm updates were mostly focused on improving the retrieval database
and the incorporation of ancillary data into the retrieval. While the first
version of GPROF still used hand crafted hydrometeor profiles to generate the
retrieval database, these were soon replaced by profiles from a meso-scale
weather model \citep{kummerow96}. An important improvement was the replacement
of the model-derived database by an observationally-generated database for the
GPROF 2010 algorithm \citep{kummerow11, kummerow15}, which helped reduce errors
caused by misrepresentation of atmospheric states in the database. The 2014
version of GPROF \citep{kummerow15} introduced the first fully-parametric
version of the algorithm, which was designed to be applicable to all sensors of
the GPM constellation. This version of GPROF became the operational PMW
precipitation retrieval of the GPM mission.

This study focuses on the computational method that is used to produce the
retrieval results from the retrieval database used by GPROF. Since its
introduction in \citet{kummerow96}, the currently used Bayesian method has not
received much consideration, mainly because the database and the incorporation
of ancillary data were deemed to be more relevant for improving the accuracy of
the retrieval. However, two disadvantages of the current retrieval method have
become apparent with the introduction of the much larger,
observationally-generated retrieval databases into the algorithm
\citep{elsaesser15}: Firstly, the retrieval database must be compressed into
self-similar clusters to reduce the processing time. This lossy compression may
limit the extent to which the current algorithm can benefit from the size and
representativeness of observationally generated retrieval databases. This is
expected to  affect retrievals of high rain rates due to their
scarcity in the retrieval database. Secondly, the accuracy of the retrieval
results depends on the uncertainties assigned to the database observations.
Since there is no principled way to calculate these uncertainties, they need to
be tuned heuristically for each sensor.

While GPROF is currently based on a data-driven method to solve Bayesian inverse
problems, more general machine learning techniques have recently gained
popularity for application in precipitation retrievals. Deep neural networks
(DNNs), which have enabled a number of significant breakthroughs in different
scientific fields \citep{silver16, jumper21}, have in recent years been explored
for retrieving precipitation from satellite observations. Especially
convolutional neural networks (CNNs) are appealing for this application because
of their ability to leverage spatial patterns in image data. This property sets
them apart from traditional retrieval methods and shallow machine-learning
techniques, which are limited in their ability to use this information by
computational complexity \citep{duncan19} or the need for feature engineering or
manual incorporation of spatial information through techniques such as
convective-stratiform discrimination \citep{kaushik10}.

Shallow neural networks have long been used to retrieve precipitation
from PMW observations \citep{staelin00, surssavadee08}. The Passive microwave
Neural network Precipitation Retrieval (PNPR) presented in \citet{sano15,
  sano16, sano18} and the work by \citet{tang18} are among the more recent
algorithms that use neural networks for retrieving precipitation from PMW
observations. They employ relatively shallow neural networks and retrieve
precipitation in a pixel-wise manner, thus neglecting spatial structure of the
observations. Other recent work demonstrates the ability of  CNNs to
leverage spatial information in  satellite observations. Examples of this are
IR-based retrievals by \citet{sadeghi19}, PMW-based precipitation detection
\citep{li21} and retrievals combining PMW with IR observations \citep{gorooh22}
and gauge measurements \citep{moraux19}.

A shortcoming of the aforementioned studies is that none of them
addresses the inherent uncertainty of the precipitation retrievals. Retrieving
precipitation from PMW observations constitutes an inverse problem, whose
ill-posed character leads to significant uncertainties in the retrieval results.
Traditionally, these uncertainties are handled using Bayesian statistics.
However, because the algorithms mentioned above neglect the probabilistic
character of the retrieval, there is no way to reconcile them with the Bayesian
approach.

Moreover, existing precipitation retrievals that make use of DNNs
\citep{moraux19, sadeghi19, li21, gorooh22} are experimental retrievals that are
currently not used operationally. The design of an operational retrieval
algorithm for the GPM PMW observations needs to address a number of additional
requirements, such as the handling of observations from different sensors and
the retrieval of multiple output variables. Furthermore, because GPM is an
ongoing mission, continuity of the output variables must be ensured, which
further constrains the design of the retrieval algorithm.

This study explores the use of DNNs for the operational retrieval of
precipitation rates and hydrometeor profiles from the PMW observations from the
GPM constellation. To this end, we present two PMW precipitation retrieval
algorithms that provide probabilistic precipitation estimates and can be used in
the operational processing pipeline for the GPM PMW observations.

\begin{description}
\item[GPROF-NN 1D] uses a fully-connected neural network to
  retrieve single column hydrometeor profiles and rain rates based on the
  observed brightness temperature vector. It thus uses exactly the same input
  data as the standard GPROF algorithm.
\item[GPROF-NN 3D]  extends the GPROF-NN 1D algorithm by
  incorporating spatial information into the retrieval using a CNN. It produces
  the same output as GPROF and GPROF-NN 1D but processes all observations
  simultaneously, thus allowing the algorithm to combine information from pixels
  across the swath.
\end{description}

The proposed algorithms are based on quantile regression neural networks (QRNNs,
\citeauthor{pfreundschuh18}, \citeyear{pfreundschuh18}), which can be used to
predict the posterior distribution of a Bayesian solution of the retrieval,
given that the assumed a priori distribution of the Bayesian solution is the
same as the distribution of the neural network's training data. Because of this, the
GPROF-NN retrievals can produce all of GPROF's retrieval outputs, which include
a probability of precipitation and an uncertainty estimate of the predicted
precipitation in the form of terciles of the posterior distribution.

Before a retrieval can replace the current operational version of GPROF, it is
imperative to establish its ability to improve the retrieval accuracy to avoid
degradation of the GPM data products. A balanced evaluation of the accuracy of
precipitation retrievals is difficult because it depends on the statistics of
the data used in the assessment. Data-driven retrievals generally yield the most
accurate results when evaluated on data with the same distribution as the data
used for their training. At the same time, evaluation against independent
measurements may distort the evaluation when these measurements deviate
significantly from the training data. In this study, the retrieval performance
of the GPROF-NN algorithms is evaluated and compared to that of GPROF using a
held-out part of the retrieval database. This provides the most direct estimate
of the benefits of the neural-network-based retrievals because it avoids the
distorting effects of using test data from a different origin. Moreover, the
nominal accuracy of both the GPROF and GPROF-NN algorithms provides a reference
for future validation against independent measurements. More specifically, this
study employs the newly developed GPROF-NN algorithms to answer the following
two questions:
\begin{enumerate}
\item Can a DNN that uses the same input information as GPROF
  provide more accurate  retrievals of surface precipitation
  and vertical
  hydrometeor profiles?
\item What is the potential of using a CNN to incorporate spatial
  information into the retrieval to further improve improve the
  accuracy of the retrievals within
  the current processing pipeline?
\end{enumerate}

%Since GPROF
%was designed to process each pixel independently, the currently used retrieval
%database is more difficult to use to train a CNN than a neural network that
%neglects the spatial structure of the satellite observations. The principal
%reason for this is that the current retrieval database only consists of
%observations from a narrow region at the center of the GPM microwave imager
%(GMI) and thus provides only limited information in the across-track
%direction. However, since it has been shown that spatial information can
%improve the accuracy of precipitation retrievals, we wanted to nonetheless
%explore the potential of such a retrieval within the constraints of the
%current processing pipeline. The two GPROF-NN algorithms were therefore
%designed as follows:

This study uses the upcoming version of the GPROF algorithm, GPROF 2021, also
known as GPROF V7 in the GPM Precipitation Processing System \citep{pps}. The
retrieval performance is assessed for two sensors of the GPM constellation: The
GPM Microwave Imager (GMI) and the Microwave Humidity Sounder (MHS,
\citet{bonsignori07}). In addition to the evaluation against the data from the
retrieval data base, the study also presents a case study of the retrieved
surface precipitation from overpasses of both GMI and MHS, which are compared to
reference measurements from the GPM combined product (CMB, \citeauthor{grecu16},
\citeyear{grecu16}) and ground-based radar measurements from the Multi-Radar
Multi-Sensor (MRMS, \citeauthor{smith16}, \citeyear{smith16}) product suite.

\section{Data and methods}

The GPROF-NN algorithms make use of the same data as the original GPROF
algorithm. This data, which we refer to as the retrieval database, defines, for
all three algorithms, the input for the retrieval and the precipitation and
hydrometeor profiles that the retrieval aims to reproduce. Because of its
fundamental importance for all retrievals considered here, this section provides
an overview of the retrieval database. This is followed by a brief description
of the current GPROF algorithm and the implementation of the GPROF-NN
retrievals.

\subsection{The retrieval database}

The GPROF retrieval database is made up of pairs of retrieval input data and
corresponding output. The input comprises PMW observations and
ancillary data. The output consists of the values of the retrieved quantities.
GPROF's retrieval outputs include surface precipitation, profiles and path
integrals of rain, snow and cloud water as well as latent heating profiles. A
listing of all retrieval targets and corresponding units is provided in
Tab.~\ref{tab:variables}.

\begin{table}
  \caption{Retrieval quantities in the retrieval database}
  \label{tab:variables}
  \centering
  \begin{tabular}{l|rr}
    Retrieval variable & Unit & Type\\
    \hline
    Surface precipitation & $\unit{mm\ h^{-1}}$ & Scalar \\
    Convective precipitation & $\unit{mm\ h^{-1}}$ & Scalar \\
    Cloud water path & $\unit{kg\ m^{-2}}$ & Scalar \\
    Rain water path & $\unit{kg\ m^{-2}}$ & Scalar \\
    Ice water path & $\unit{kg\ m^{-2}}$ & Scalar \\
    Cloud water content & $\unit{g\ m^{-3}}$ & Profile \\
    Rain water content & $\unit{g\ m^{-3}}$ & Profile \\
    Snow water content & $\unit{g\ m^{-3}}$ & Profile \\
    Latent heating & $\unit{K \ h^{-1}}$ & Profile \\
  \end{tabular}
\end{table}

Since the available channels and the viewing geometries vary between the
sensors of the GPM constellation, a separate database is generated for each
sensor type. A crucial difference between the retrieval databases for GMI and
the other sensors of the GPM constellation is that the database for GMI uses
real observations, while the databases for the other sensors are constructed
using simulations. The varying resolutions and viewing geometries of different
sensors are taken into account by resampling and averaging the simulated
observations and retrieval results to the observation footprints of the
corresponding sensor. The channels of the GMI and MHS sensors that are used in
this study are listed in Tab.~\ref{tab:channels}.

\begin{table}[hbpt]
  \caption{Channels of the GMI and MHS sensors used for the retrievals in this study.}
  \label{tab:channels}
  \begin{tabular}{lrr}
    Channel & Freq. [$\unit{GHz}$] & Pol. \\
    \hline
   GMI-1  & $10.6$ & V  \\
   GMI-2  & $10.6$ & H  \\
   GMI-3  & $18.7$ & V  \\
   GMI-4  & $18.7$ & H  \\
   GMI-5  & $23$   & V  \\
   GMI-6  & $37$   & V  \\
   GMI-7  & $37$   & H  \\
   GMI-8  & $89$   & V  \\
   GMI-9  & $89$   & H  \\
   GMI-10 & $166$ & V \\
   GMI-11 & $166$ & H  \\
   GMI-12 & $183 \pm 3$ \ & V \\
   GMI-13 & $183 \pm 7$ \ & V \\
  \end{tabular}%
  \hspace{1cm}
    \begin{tabular}{lrr}
      Sensor & Freq. [$\unit{GHz}$] & Pol. \\
      \hline
      MHS-1 &  89 & V \\
      MHS-2 &  157 & V \\
      MHS-3 &  $183 \pm 1$ & H \\
      MHS-4 &  $183 \pm 3$ & H \\
      MHS-5 &  $190.31$ & V \\
    \end{tabular}
\end{table}

%For brevity, we refer to the pairs that make up the retrieval datase as
%'database profiles' or simply 'profiles' due to them being derived from or
%corresponding to specific vertical profiles of atmospheric variables.

The databases for GPROF 2021 are derived from one year (October 2018 to
September 2019) of retrieved hydrometeor profiles from the GPM CMB product
\citep{grecu16}. This data is complemented with surface precipitation from the
currently operational Microwave Integrated Retrieval System \citep{boukabara11},
which adds light precipitation in areas where no echo is detected by the GPM
Dual-Frequency Precipitation Radar. Observations over sea-ice and snow-covered
surfaces are handled separately. For sea-ice, precipitation is derived from the
ERA5 reanalysis \citep{hersbach20}. For snow-covered surfaces, precipitation is
derived from several years of co-locations with gauge-corrected radar measurements
from MRMS \citep{smith16}.

The ancillary data that serve as additional retrieval inputs are derived from
reanalysis datasets. They consist of two meter temperature ($T_{\text{2m}}$),
total column water vapor ($\text{TCWV}$), the surface type as well as an air
lifting index (ALI) that encodes information on atmospheric convergence in
mountainous areas. The ancillary data for the databases used in this study were
derived from the ERA5 reanalysis \citep{hersbach20}.

A detailed description of the retrieval database and the derivation of the data
it contains can be found in the GPROF ATBD \citep{atbd}. The training data for the
GPROF-NN retrievals consists of the data from the retrieval database. The
training data is stored in an intermediate format to simplify the loading of the
data during training of the neural network. The format and the creation process
of the training data is described in detail in Sec.~\ref{sec:gprof_nn_training_data}
in the appendix.

\subsection{The GPROF algorithm}
\label{sec:gprof}

The current implementation of GPROF uses a Bayesian scheme to retrieve
precipitation and hydrometeor profiles, which works by resampling the profiles
in the database based on the similarity of the observations and ancillary data.
GPROF uses ancillary data to split the database into separate bins. This reduces
the number of profiles for which weights must be computed and helps to constrain
the retrieval. Moreover, the profiles in each bin are clustered to limit the
number of profiles that need to be processed. A detailed description of the
implementation of GPROF is provided in Sec.~\ref{app:gprof} in the appendix.

\subsection{The GPROF-NN algorithms}

The principal objective guiding the design of the GPROF-NN algorithms was to
develop a neural-network-based retrieval that operates on the same input data
and provides the same output as GPROF so that it can replace the current
implementation in a future update. Although GPROF's retrieval scheme is defined
on independent pixels, the algorithm processes full orbits of observations and
corresponding ancillary data. Both GPROF-NN retrievals were therefore designed
to process the same input format as GPROF, which corresponds to each sensor's
level 1C observations in their native spatial sampling, which, where required,
is remapped to a common grid. The output from all retrievals is on the same grid
as the input.

GPROF produces multiple probabilistic outputs: A probability of precipitation
and the mode and terciles of the posterior distribution of precipitation. An
implementation based on standard regression neural networks would not provide
any principled way to produce these probabilistic outputs due to the
incompatibility of deterministic regression with the Bayesian retrieval
formulation used in GPROF. The implementation of the GPROF-NN retrievals uses
quantile regression neural networks (QRNNs) to overcome this limitation. As
shown in \citet{pfreundschuh18}, when trained on a dataset distributed according
to the a priori distribution of a Bayesian retrieval, QRNNs learn to predict
quantiles of the Bayesian posterior distribution. They thus provide a simple and
efficient way to reconcile neural network retrievals with the Bayesian framework
employed by GPROF.

QRNNs predict a sequence of quantiles, which allows reconstructing the
cumulative distribution function (CDF) of the a posteriori distribution of any
scalar retrieval quantity. Since the distribution of a scalar variable is fully
described by its CDF, any relevant statistic of the a posteriori distribution
can be derived from the predicted CDF. The GPROF-NN retrievals use the predicted
CDF to derive the most likely and mean surface precipitation (the latter of
which is identical to the solution that would have been obtained with standard
mean squared error regression), the terciles of the posterior distribution, and
the probability of precipitation. Fig.~\ref{fig:gprof_nn_principle} illustrates
the principle of the GPROF-NN retrievals: The retrieval employs a neural network
to predict a vector of values for each pixel in the input observations. The
elements of this vector correspond to a sequence of quantiles of the a
posteriori distribution. These quantiles are used to reconstruct a piece-wise
linear approximation of the CDF of the distribution from which the retrieval
results are derived.

\begin{figure}[hbpt!]
  \centering \includegraphics[width=1.0\textwidth]{figs_revised/fig01}
  \caption{The basic principle of the implementation of the GPROF-NN retrievals.
    A Bayesian solution of the retrieval is obtained by predicting, for each
    input pixel, a sequence of quantiles of the a posteriori distribution that
    is used to reconstruct its CDF. The predicted CDF is then used to derive the
    scalar retrieval results.}
  \label{fig:gprof_nn_principle}
\end{figure}

\subsubsection{Training objectives}
\label{sec:objectives}

A neural network can be trained to predict a quantile $\hat{x}_\tau$ of a given
conditional distribution by training it to minimize the quantile 
loss function $\mathcal{L_\tau}$ corresponding to the quantile
fraction $\tau$ \citep{koenker01}:
\begin{align}\label{eq:quantile_loss}
  \mathcal{L}_\tau(\hat{x}_\tau, x) &= (\tau - \mathbb{I}_{x \leq \hat{x}_\tau})(x
  - \hat{x}_\tau),
\end{align}
where $\hat{x}_\tau$ is the predicted quantile, $x$ is the reference value from the
training data and $\mathbb{I}_{x \leq \hat{x}}$ is the indicator function taking
  the value 1 when the condition $x \leq \hat{x}$ is true and 0 otherwise.

This principle can be extended to a sequence of quantiles corresponding to
quantile fractions $\tau_1, \ldots, \tau_N$ by minimizing the mean of the loss
functions corresponding to each quantile fraction:
\begin{align}\label{eq:loss_function}
  \mathcal{L}_{\tau_1,\ldots, \tau_N}(\hat{\mathbf{x}}, x) &= \frac{1}{N}
  \sum_{i=0}^N
  \mathcal{L}_{\tau_i}(\hat{x}_i, x)
\end{align}
where $\hat{x}_i$ is the $i$th component of the vector of predicted quantiles
$\hat{\mathbf{x}}$. The GPROF-NN retrievals use this loss function with 128
equally spaced quantiles ranging from $\tau_1 = 0.001$ to $\tau_{128} = 0.999$
for all scalar retrieval variables.

A difficulty with predicting quantiles of precipitation is that that lower
quantiles may become degenerate due to the high probability of no precipitation.
For example, it is impossible to predict empirical quantiles with $0 < \tau <
0.5$ for a pixel with $50\ \unit{\%}$ probability of precipitation. To allow
monitoring of the ability of the network to correctly predict retrieval
uncertainty up to the degeneracy induced by non-raining pixels, we replace rain
rates of non-raining pixels with random values from a log-uniform distribution
that are smaller than the smallest rain rate in the training data. During the
retrieval, predicted precipitation rates that are smaller than this threshold
are set to zero. The threshold is chosen as $10^{-4}\ \unit{mm\ h^{-1}}$ and
thus has negligible impact on mean or accumulated precipitation.

An additional advantage of the application of the quantile loss function is that
the training can be performed on transformed retrieval outputs without changing
the statistical properties of the network predictions given that the
transformation function is strictly monotonic. The training of all scalar,
non-negative retrieval quantities uses a log-linear transformation function of
the form
\begin{align}\label{eq:log_linear}
  f(x) &= \begin{cases}
    \log(x) & \text{if } x < 1 \\
    x - 1 & \text{if otherwise}.
  \end{cases}
\end{align}
In addition to avoiding the prediction of negative values, we found this to
slightly increase retrieval accuracy for quantities that vary by multiple
orders of magnitude, which precipitation rates and hydrometeor concentrations
typically do.

For hydrometeor profiles, the retrieval is implemented in a slightly different
manner. To reduce the number of network outputs, the posterior mean of
hydrometeor profiles is predicted directly using mean squared error regression.
Since the output of GPROF contains only the posterior mean of the hydrometeor
concentrations, it was deemed unnecessary to predict their full posterior
distribution at each level using quantile regression. To avoid the prediction
of negative concentrations ReLU activation functions are applied to the
network outputs corresponding to hydrometeor concentrations.

\subsubsection{GPROF-NN 1D}

The GPROF-NN 1D retrieval was designed to use the same input information and
produce the same output as the Bayesian scheme used by GPROF. GPROF-NN 1D thus
operates on single pixels of brightness temperatures and  ancillary
data and predicts the corresponding precipitation and hydrometeor profiles.

\begin{figure}[hbpt]
  \centering
    \includegraphics[width=\textwidth]{figs_revised/fig02}
    \caption{Illustration of the neural network architecture used in the
      GPROF-NN 1D algorithm. The network consists of a common body and
      one head for each retrieval variable. Each block in body and head
      consists of fully-connected layer, layer norm and GELU activation function.
    }
  \label{fig:gprof_nn_1d}
\end{figure}

The neural network architecture used for the GPROF-NN 1D retrieval is
illustrated in Fig.~\ref{fig:gprof_nn_1d}. A single network is trained to
predict all retrieval variables (c.f. Tab.~\ref{tab:variables}) using the
training objectives described in Sec.~\ref{sec:objectives}. The network consists of
a shared body and a separate head for each retrieved variable. Body and heads
are built-up of blocks consisting of a fully-connected layer followed by layer
normalization \citep{ba16} and GELU \citep{hendrycks16} activation functions.
During development we have experimented with different numbers of blocks in body
($N_b$) and each of the heads ($N_h$) but found only marginal impact on the
retrieval performance and settled for a configuration with $N_b = 6$ and $H_h =
4$.

%The GPROF-NN 1D network is trained by simultaneously minimizing the sum of the
%$losses of all retrieval variables. The training is performed over 70 epochs
%$using the Adam optimizer \citep{kingma14} with an initial learning rate of
%$$5\cdot10^{-4}$ and a cosine annealing learning rate schedule
%$\citep{loshchilov16}. Warm restarts are performed after 10, 30 and 50 epochs.

Detailed descriptions of the neural network training and the retrieval
processing for GPROF-NN 1D are provided in Sec.~\ref{sec:gprof_nn_training} and
Sec.~\ref{sec:gprof_nn_processing}, respectively.

\subsubsection{GPROF-NN 3D}

The GPROF-NN 3D retrieval extends the GPROF and GPROF-NN 1D algorithms by
incorporating structural information into the retrieval. To achieve this, the
GPROF-NN 3D algorithm employs a CNN that performs the retrieval for all pixels
in the swath simultaneously.

\begin{figure}[hbpt]
  \centering
  \includegraphics[width=1.0\textwidth]{figs_revised/fig03}
  \caption{
    The neural network architecture of the GPROF-NN 3D retrieval. Illustration
    (a) displays the structure of the Xception blocks \citep{chollet17} that
    form the building blocks of the GPROF-NN 3D model. An Xception block
    consists of two depthwise separable convolutions followed by a group
    normalization layer and a GELU activation function. Illustration (b) shows
    how the Xception blocks are used in an asymmetric encoder-decoder structure
    that forms the body of the network. Output from the body is combined with
    the ancillary data to form the inputs to the separate heads that predict the
    retrieval results for each of the retrieved variables.
  }
  \label{fig:gprof_nn_3d_arch}
\end{figure}

The network architecture for the GPROF-NN 3D algorithm, illustrated in
Fig.~\ref{fig:gprof_nn_3d_arch}, consists of an asymmetric encoder-decoder
structure followed by a separate head for each retrieved variable. The stages of
the en- and decoder are built up of what we refer to here as Xception blocks
(Fig.~\ref{fig:gprof_nn_3d_arch} (a)) because they are based on the Xception
architecture introduced in \citet{chollet17}. Each block consists of two
depthwise separable convolutions with a kernel size of 3 followed by group
normalization layers with 32 groups and GELU activation functions. The first
block in each stage of the encoder additionally contains a $3\times3$
max-pooling layer with a stride of 2 following the first $3\times3$ convolution
layer. Each downsampling block in the encoder is followed by $N = 4$ standard
Xception blocks. The stages of the decoder consist of a bi-linear upsampling
layer followed by a single Xception block. The network architecture was chosen
with the aim of maximizing the depth and width of the network while keeping the
time required for processing an orbit low. Symmetric padding is performed before
all convolution operations with a kernel size larger than one in order to
conserve the input size.

%The fully-convolutional network architecture employed by the GPROF-NN 3D
%retrieval enables the model to process inputs of arbitrary sizes. This allows
%processing of a whole orbit of observations as a single input, thus in principle
%omitting the need for splitting up the observations into overlapping tiles to
%avoid edge effects.

\begin{figure}[hbpt]
  \centering
    \includegraphics[width=\textwidth]{figs_revised/fig04}
    \caption{
      The effect of GMIs conical viewing geometry on observed features. Panel
      (a) displays geolocated observations of the $10.6\ \unit{GHz}$ channel
      (colored background). Grey squares mark equilaterals with a side length of
      $200\unit{km}$ oriented along the swath. The highlighted stripe located at
      the swath center marks the region where the values of the retrieved
      variables are known. Panel (b) shows the same observations viewed as an
      image on a uniform grid. Panel (c) shows six synthetically generated
      training inputs based on two input regions marked in Panel (b). The first
      row shows three synthetic samples that simulate the effect of viewing the
      input in region A at a different position across the GMI swath. The second
      row shows the corresponding transformations for the input in region B.
    }
  \label{fig:data_augmentation}
\end{figure}

Additional complexity in the training of the GPROF-NN 3D retrieval derives from
the requirement to operate on the same data as GPROF, which means that input and
output data must be on the native observation grid of each sensor. This is
problematic because the viewing geometries of PMW sensors break the
translational symmetry of digital images that constitutes one of the inductive
biases of CNNs \citep{goodfellow16}. For example, geo-located pixels of conical
scanners do not lie on a rectangular grid, which causes shapes to appear
differently depending on their position in the swath.
Fig.~\ref{fig:data_augmentation} illustrates this for GMI observations. The
rectangular shapes shown in panel (a) are distorted when the observations are
plotted on a uniform grid (panel (b)).

Moreover, because GPROF currently only uses the central 21 pixels of the CMB
product for the generation of the retrieval database, the values of the
retrieval targets in the GPROF database are known only at the central pixels of
the GMI swath. The location of these pixels is marked by the light stripe in
Fig.~\ref{fig:data_augmentation}. The neural network can thus learn the spatial
structure of precipitation only from the central part of the GMI observations.

The training of the GPROF-NN 3D retrieval employs a customized data augmentation
scheme to account for the aforementioned characteristics of the training data.
Training samples for the GPROF-NN 3D retrieval are transformed to simulate the
effect of observing each training scene at varying locations of the sensor
swath. The transformations are applied randomly when a training sample is
loaded, thus ensuring that the network rarely or never sees a training scene
from the same perspective. The transformations also vary the relative location
of the pixels at which values of retrieval variables are known across the full
width of the swath instead of always being located at its center. Examples of
transformed inputs for GMI are displayed in Fig.~\ref{fig:data_augmentation}
(c).

A detailed description of the  training and the retrieval
processing for the GPROF-NN 3D retrieval are provided in
Sec.~\ref{sec:gprof_nn_training} and Sec.~\ref{sec:gprof_nn_processing},
respectively.

\subsubsection{Extension to other sensors}
\label{sec:other_sensors}

The GPROF retrieval for GMI is special because it is the only sensor of the GPM constellation for which the retrieval inputs used in the database correspond to real observations. For the other sensors, the observations used to construct the retrieval database are simulated. Since the simulations take into account the effect of the different viewing geometries and resolutions, GPROF-NN 1D inherits its ability to handle observations from different sensors directly from the design of retrieval database.

For the GPROF-NN 3D algorithm this is not the case. The problem for sensors other than GMI is that the retrieval database contains simulated observations only at the central pixels of the GMI swath (the highlighted pixels in Fig.~\ref{fig:data_augmentation} (a)). To obtain two-dimensional training scenes that are sufficiently wide to train a CNN, we make use of an intermediate CNN based model to 'retrieve' simulated brightness temperatures across the full GMI swath. The extended simulated brightness temperatures are then remapped from the GMI viewing geometry to the viewing geometry of the target sensor. While this approach is certainly not ideal with respect to the realism of the generated scenes, it was the simplest and currently only feasible way to extend the GPROF-NN 3D retrieval to other sensors than GMI using only currently available data from the GPROF database. A detailed description of the procedures involved in generating the training data for different sensors is provided in Sec.~\ref{sec:gprof_nn_training_data} in the appendix.

Moreover, since the databases for other sensors rely on simulations, it is not guaranteed that the distribution of brightness temperatures in the database matches those of actual observations. The simulations are therefore corrected using a surface type and total TCWV-dependent correction that matches the quantiles of the conditional distributions of simulated and real observations. The GPROF algorithm's correction distinguishes three different surface types. However, the GPROF-NN algorithms use a correction with all 18 surface types because the correction used by GPROF was found to be too crude over land surfaces leading to artifacts in the retrieval results.

\section{Results}

This section presents the results of the evaluation of GPROF and the novel
GPROF-NN algorithms. The first part evaluates the retrievals using a held-out
test data set. The remainder of this section presents a case study of
precipitation retrievals from hurricane Harvey followed by a brief assessment of
the processing times of the different algorithms.

\subsection{Assessment on held-out test data}
\label{sec:results_test}

The held-out test data comprises observations from the retrieval database from the first three days of every month. This data has not been used for training the neural network retrievals. It is, however, derived from the same data sources and thus stems from the same distribution as the training data.

Tab.~\ref{tab:test_data} lists the number of pixels with precipitation information used for testing the retrievals. The evaluation of the GPROF-NN 3D retrieval uses spatially contiguous scenes of the same size as the ones used during its training. Since these scenes typically do not cover all of the pixels with precipitation information, the test data for the GPROF-NN 3D retrievals contain fewer pixels that can be used for evaluation. The lower number of test pixels for MHS is due to the coarser resolution of the observations, which leads to a smaller number of observations over sea-ice and snow and an additional reduction of the pixels available for evaluation of the GPROF-NN 3D retrieval.
\begin{table}[hbpt]
  \caption{
    Number of pixels with precipitation information in the test datasets
    used to evaluate the retrievals.
  }
  \label{tab:test_data}
  \begin{tabular}{|l||r|r|}
    \hline
    Sensor & GPROF \& GPROF-NN 1D & GPROF-NN 3D \\
    \hline
    \hline
    GMI & $50\,435\,584$ & $14\,218\,203$ \\
    \hline
    MHS & $24\,975\,877$ & $4\,945\,165$ \\
    \hline
  \end{tabular}
\end{table}

\subsubsection{Precipitation and hydrometeor paths}

As described in Tab.~\ref{tab:variables}, the scalar variables retrieved by
GPROF are surface and convective precipitation as well as the column-integrated
concentrations of cloud droplets, rain and snow. They are denoted as cloud water
path (CWP), rain water path (RWP) and ice water path (IWP), respectively.
Scatter plots of the retrieval results for these five quantities evaluated over all
surfaces are displayed in Fig.~\ref{fig:results_scatter_gmi} for GMI and
Fig.~\ref{fig:results_scatter_mhs} for MHS. The frequencies in all plots have
been normalized column-wise to ensure that results for high reference values
remain visible.

\begin{figure}[hbpt]
  \centering
  \includegraphics[width=\textwidth]{figs_revised/fig05.png}
  \caption{
    Scatter plots of scalar retrieval targets for the three retrieval algorithms
    for GMI. Rows display the results for the GPROF, GPROF-NN 1D and GPROF-NN
    3D algorithms, respectively. Columns display the results for different
    retrieval targets. Frequencies in the plots have been normalized
    column-wise, i.e. per bin of the reference value.
  }
  \label{fig:results_scatter_gmi}
\end{figure}

Consistent improvements in the accuracy of the surface precipitation retrieved
by GMI are observed between GPROF and GPROF-NN 1D as well as GPROF-NN 1D and
GPROF-NN 3D. The improvements are most pronounced for light rates between
$10^{-2}$ and $10^{-1}\ \unit{mm\ h^{-1}}$ but are consistent across the full
range of values. The comparably bad performance of GPROF for light precipitation
is likely due to the tuning of the assigned uncertainties to yield good results
for heavier rain that is more relevant for rainfall accumulation.

For convective precipitation, the results of GPROF deviate noticeably from the
diagonal. The results of GPROF-NN 1D slightly improve upon those  of GPROF. Although
the mode of the distribution is still displaced from the diagonal, the GPROF-NN
3D algorithm yields the best agreement with the reference data. For the
path-integrated quantities, similar improvements between GPROF and GPROF-NN 1D
as well as GPROF-NN 1D and GPROF-NN 3D are observed. Large cloud water path
values are underestimated by all retrievals which is likely because these values
are associated with precipitation but difficult to distinguish from it. Due to
the lack of a cloud water path signal in raining profiles, all algorithms resort
to predicting the climatology in the presence of significant rain.

\begin{figure}[hbpt]
  \centering
  \includegraphics[width=\textwidth]{figs_revised/fig06}
  \caption{
    Like Fig.~\ref{fig:results_scatter_gmi} but for MHS.
  }
  \label{fig:results_scatter_mhs}
\end{figure}

The results for MHS, displayed in Fig~\ref{fig:results_scatter_mhs}, paint a
similar picture. Although the overall accuracy of all retrievals is lower than
for GMI, GPROF-NN 1D consistently yields more accurate results than GPROF. Also
here the GPROF-NN 3D retrieval yields further, consistent improvements compared
to the GPROF-NN 1D retrieval.

Quantitative measures of the retrieval accuracy for surface precipitation of the
three retrieval algorithms are displayed in Tab.~\ref{tab:metrics_gmi} for GMI
and Tab.~\ref{tab:metrics_mhs} for MHS. Similar tables for the other retrieval
quantities are provided in
Tab.~\ref{tab:metrics_gmi_convective}-\ref{tab:metrics_mhs_cwp} in the appendix.
Each table displays bias, mean absolute error (MAE), mean squared error (MSE),
the symmetric mean absolute percentage error (SMAPE$_{t}$) for all test samples
with a reference value that exceeds a quantity-specific threshold $t$ and the
correlation. The error metrics confirm the qualitative findings from
Fig.~\ref{fig:results_scatter_gmi} and \ref{fig:results_scatter_mhs}: The neural
network implementations outperform GPROF in terms of all considered metrics.
Moreover, the GPROF-NN 3D algorithm further improves upon the performance of the
GPROF-NN 1D algorithm. The same tendency is observed for MHS, albeit with lower
overall accuracy.

\begin{table}[hbpt!]
  \caption{Mean error metrics and estimated standard deviation for surface
    precipitation retrieved from GMI observations.}
  \label{tab:metrics_gmi}
  \begin{tabular}{|c||p{3.5cm}|p{3.5cm}|p{3.5cm}|}
    \hline
    Metric &
    \multicolumn{1}{|c}{GPROF} &
    \multicolumn{1}{|c}{GPROF-NN 1D} &
    \multicolumn{1}{|c|}{GPROF-NN 3D} \\
    \hline\hline
    Bias \hfill [$\unit{mm\ h^{-1}}$] & \hfill $ -0.0029 \pm 0.0001$ &\hfill $ -0.0024 \pm 0.0001$   &\hfill $ -0.0006 \pm 0.0001$ \\
    MAE \hfill [$\unit{mm\ h^{-1}}$] & \hfill $  0.0788 \pm 0.0001$ &\hfill $  0.0585 \pm 0.0001$    &\hfill $  0.0444 \pm 0.0001$ \\
    MSE \hfill [$\unit{mm\ h^{-1}}$] & \hfill $  0.1965 \pm 0.0001$ &\hfill $  0.1379 \pm 0.0001$    &\hfill $  0.0983 \pm 0.0001$ \\
    SMAPE$_{0.01}$ \hfill [$\unit{\%}$] & \hfill $ 76.0598 \pm 0.0139$ &\hfill $ 69.5382 \pm 0.0127$ &\hfill $ 56.0040 \pm 0.0181$ \\
    Correlation & \hfill $ 0.7971$ &\hfill $  0.8470$ &\hfill $ 0.8966$ \\
    \hline
  \end{tabular}

\end{table}

\begin{table}[hbpt!]
  \caption{Mean error metrics and estimated standard deviation for surface
    precipitation retrieved from MHS observations.}
  \label{tab:metrics_mhs}
  \begin{tabular}{|c||p{3.5cm}|p{3.5cm}|p{3.5cm}|}
    \hline
    Metric &
    \multicolumn{1}{|c}{GPROF} &
    \multicolumn{1}{|c}{GPROF-NN 1D} &
    \multicolumn{1}{|c|}{GPROF-NN 3D} \\
    \hline\hline
    Bias \hfill [$\unit{mm\ h^{-1}}$] & \hfill $ -0.0110 \pm 0.0001$ &\hfill $ -0.0066 \pm 0.0001$ &\hfill $ -0.0018 \pm 0.0001$ \\
    MAE \hfill [$\unit{mm\ h^{-1}}$] & \hfill $  0.0846 \pm 0.0001$ &\hfill $  0.0609 \pm 0.0001$ &\hfill $  0.0487 \pm 0.0001$ \\
    MSE \hfill [$\unit{mm\ h^{-1}}$] & \hfill $  0.2317 \pm 0.0001$ &\hfill $  0.1682 \pm 0.0001$ &\hfill $  0.1087 \pm 0.0001$ \\
    SMAPE$_{0.01}$ \hfill [$\unit{\%}$] & \hfill $ 80.8641 \pm 0.0190$ &\hfill $ 68.4961 \pm 0.0185$ &\hfill $ 62.3086 \pm 0.0377$ \\
    Correlation & \hfill $  0.7239 \pm 0.0000$ &\hfill $  0.8040 \pm 0.0000$ &\hfill $  0.8400 \pm 0.0000$ \\
    \hline
  \end{tabular}
\end{table}

\begin{figure}[hbpt]
  \centering
  \includegraphics[width=\textwidth]{figs_revised/fig07.pdf}
  \caption{
    Bias, MSE, MAE and SMAPE$_{0.01}$ and correlation of the retrieved surface
    precipitation w.r.t. surface type and retrieval algorithm. The results for
    the GMI sensor are displayed in the first column, results for the MHS sensor
    in the second column. Error bars mark one standard deviation around the
    mean.}
  \label{fig:results_surface_types}
\end{figure}

Since the surface type has a considerable effect on the lower-frequency observations used in the retrieval, its impact on the retrieval of surface precipitation is assessed in Fig.~\ref{fig:results_surface_types}. The figure displays bias, MSE, MAE, SMAPE, and correlation for principal surface types. For the analysis, original GPROF surface types have been grouped into ocean (surface type 1), dense vegetation (surface types 3 - 5), sparse vegetation (6 - 7), snow (surface types 8-11), and coast (surface types 12-15). Even when the different surface types are considered separately, the results of the surface precipitation retrieval show the same pattern as the scatter plots in Fig.~\ref{fig:results_scatter_gmi}. The results of the GPROF-NN 1D retrieval are generally more accurate than those of GPROF, and the results of the GPROF-NN 3D algorithm are more accurate than those of the GPROF-NN 1D algorithm. These findings are mostly consistent across the considered surface types and both sensors. Exceptions are the biases of the GPROF-NN 1D algorithm for GMI over densely vegetated surfaces and for MHS over snow, which are larger than those of GPROF, and the MSE of the GPROF-NN 3D algorithm for MHS over snow, which is slightly larger than that of the GPROF-NN 1D retrieval. We suspect this is caused by the relative scarcity of the observations in the retrieval database.

Figure~\ref{fig:results_spatial_gmi} displays the geographical distribution of
bias, MSE and SMAPE for GMI in $5\ \unit{^{\circ}} \times 5\ \unit{^{\circ}}$
boxes. As could be expected from the previous results, the magnitudes of the
biases of GPROF are considerably larger than for the other two algorithms.
Furthermore, GPROF exhibits consistent biases across geographical regions such
as the Northwest Atlantic and Northwest Pacific, which is less the case for the
neural network algorithms. Although spatial distribution of the MSE mostly
reflects the global distribution of precipitation, a gradual decrease in MSE can
be observed between the results of GPROF and GPROF-NN 1D as well as GPROF-NN 1D
and GPROF-NN 3D. More consistent patterns are visible in the SMAPE: The largest
errors for all three retrievals occur over land surfaces, which likely reflects
the decrease in information content due to the reduced contrast in the lower
frequency channels. Over Ocean, errors are generally higher in the sub-tropics and
tropics compared to higher latitudes. Although these patterns are observed in
the results of all algorithms, a clear and globally consistent decrease in SMAPE
can be observed between the GPROF, GPROF-NN 1D and GPROF-NN 3D retrievals.

The corresponding results for MHS are provided in
Fig.~\ref{fig:results_spatial_mhs}. Again, although the errors are slightly
larger, the results are qualitatively similar.

\begin{figure}[hbpt]
  \centering
  \includegraphics[width=\textwidth]{figs_revised/fig08}
  \caption{
    Spatial distributions of bias (Column 1), MSE (Column 2), SMAPE$_{0.01}$
    (Column 3), and the counts in each $5\ \unit{^\circ} \times
    5\ \unit{^\circ}$ degree box (Column 4) for the GPROF (Row 1), GPROF-NN 1D
    (Row 2), GPROF-NN 3D (Row 3) algorithms for GMI.
  }
  \label{fig:results_spatial_gmi}
\end{figure}

\subsubsection{Predicted retrieval uncertainties and probabilistic rain detection}

In addition to quantitative precipitation estimates, GPROF produces estimates of
the first and second tercile of the posterior distribution of surface
precipitation, which provide an uncertainty estimate for the retrieved mean
surface precipitation, as well as a probabilistic classification of pixels into
raining and non-raining pixels based on an estimated probability of
precipitation. Due to their probabilistic nature, similar results can be
produced using the GPROF-NN algorithms. From the predicted quantiles, the terciles
can be inferred by interpolating them to the fractions $\frac{1}{3}$ and
$\frac{2}{3}$, respectively. The probability of precipitation is calculated by
using the predicted posterior distribution to calculate the probability of the
retrieved surface precipitation to be larger than the smallest non-zero rain
rate in the training data.

\begin{table}
  \caption{
    Calibration of the predicted terciles of the posterior distribution of
    surface precipitation  for the three retrieval algorithms and the GMI
    and MHS sensors.
  }
  \label{tab:terciles}
    \centering
  \begin{tabular}{|l|r|rrr|rrr|}
    \hline
    \multicolumn{1}{|c|}{\multirow{2}{*}{Tercile}} &
    \multicolumn{1}{c|}{\multirow{2}{*}{Nominal}} &
    \multicolumn{3}{c|}{GMI} &
    \multicolumn{3}{c|}{MHS} \\
    & & GPROF & GPROF-NN 1D & GPROF-NN 3D & GPROF & GPROF-NN 1D & GPROF-NN 3D \\
    \hline
    First & 0.333 & 0.461 & 0.351 & 0.349  & 0.274 & 0.34 & 0.326 \\
    Second & 0.667 & 0.514 & 0.652 & 0.654  & 0.480 & 0.664 & 0.649 \\
    \hline
  \end{tabular}

\end{table}

To assess the accuracy of the uncertainty estimates from GPROF and the GPROF-NN
algorithms, Tab.~\ref{tab:terciles} lists the calibration, i.e. the frequency
with which each predicted tercile was larger than the true surface
precipitation. The evaluation of the results for the GPROF-NN algorithms was
performed with the replacement of non-raining values described in
Sec.~\ref{sec:objectives}, which allows to account for degenerate quantiles.
Since no such mechanism is available for GPROF it is not possible to evaluate
the calibration of the predicted terciles without their effect. At least
partially because of this, the calibration of GPROF deviates from the nominal
frequencies. For the GPROF-NN algorithms, however, both algorithms yield
frequencies that are close to the expected frequencies of $\frac{1}{3}$ and
$\frac{2}{3}$, respectively.

\begin{figure}[hbpt!]
  \centering
  \includegraphics[width=0.75\textwidth]{figs_revised/fig09}
  \caption{
    Calibration (Panel (a)) and receiver-operating characteristic (ROC, Panel(b)) for
    the predicted probability of precipitation.
  }
  \label{fig:results_pop}
\end{figure}

The quality of the raining/non-raining classification is assessed in
Fig.~\ref{fig:results_pop}, which displays the calibration of the predicted
probability and the receiver-operating characteristic (ROC) curve. The predicted
probabilities are fairly well calibrated for all algorithms and sensors.
Nonetheless, the GPROF-NN algorithms yield results that are slightly closer to
the diagonal. The results for the ROC curves are analogous: The GPROF-NN 1D
algorithm yields better precipitation detection than GPROF and the GPROF-NN 3D
retrieval in turn yields slightly better performance than the 1D version. In
terms of classification skill, worse performance is achieved for GMI than for
MHS by all algorithms, but again the relative performance of the retrievals
is the same.

\subsubsection{Effective resolution}
\label{sec:effective_resolution}

Next, we aim to assess the impact of the retrieval method on the effective
resolution of the retrieved precipitation fields which is important for hydrologic
applications. For this, we adopt the approach from \citet{guilloteau17}, who
have studied the effective resolution of the previous version of GPROF for the
GMI and the TRMM Microwave Imager sensors. A 1-dimensional Haar wavelet
decomposition in along-track direction over all 128 pixel long sequences in the
test data is performed to calculate the effective resolution. We do not consider
observations for different surface types separately. Following,
\citet{guilloteau17} we examine energy spectra as well as correlation
coefficients and Nash-Sutcliffe (NS) efficiency of the coefficients of the
wavelet decomposition for the reference and retrieved surface precipitation. The
results are displayed in Fig.~\ref{fig:resolution}.

\begin{figure}[hbpt!]
  \centering \includegraphics[width=\textwidth]{figs_revised/fig10}
  \caption{ Spatial variability of retrieved and reference fields. Panel (a)
    shows the average of the total energy defined as the sum of the squared
    wavelet coefficients at different length scales for the reference and
    retrieved surface precipitation fields. Panel (b) shows the correlation
    coefficient between the coefficients of the reference and the retrieved
    precipitation field. Panel (c) shows the corresponding Nash-Sutcliffe
    efficiency.}
  \label{fig:resolution}
\end{figure}

An obvious difference to the results from \citet{guilloteau17} is that the
energy spectrum of the reference precipitation field is not monotonically
decreasing. The reason for this is that the reference precipitation field in the
retrieval database is smoothed using an averaging filter adapted to the
footprint size of the respective sensor. For GMI, the GPROF-NN 3D algorithm has
the highest variability in the retrieved precipitation field, followed by the
GPROF-NN 1D algorithm and GPROF. However, the variability of all retrievals
remains lower than that of the reference field. The correlation of the wavelet
coefficients at different scales (Panel (b)) is highest for the GPROF-NN 3D
algorithm, followed by the GPROF-NN 1D algorithm and GPROF. The same pattern is
observed for the NS efficiency. In terms of effective resolution, defined,
following \citet{guilloteau17}, as the smallest scale at which the NS efficiency
exceeds 0.5, the GPROF-NN 3D algorithm for the GMI sensor achieves a resolution
solution of $13.5\ \unit{km}$, which is the distance between consecutive pixels
in along-track direction and thus the smallest spatial scale that can be
resolved in this analysis. The effective resolutions for the GPROF-NN 1D
algorithm is $14.1\ \unit{km}$ and for GPROF $23.1\ \unit{km}$.

For MHS, the effective resolutions of GPROF and GPROF-NN 1D with
$104\ \unit{km}$ and $73\ \unit{km}$, respectively, is significantly higher than
for GMI. Since the resolution is averaged over the viewing angles of the cross
track scanner and because of its generally lower sensitivity to precipitation, a
certain degradation of the resolution was expected. Despite this, the GPROF-NN
3D algorithm achieves a resolution of $22.3\ \unit{km}$, which is close to that
of GPROF for GMI.

\subsubsection{Profile retrieval variables}

In addition to precipitation fields and path-integrated hydrometeor
concentrations, GPROF retrieves concentration profiles of rain, snow and cloud
water. The retrieval database also contains latent heating rates as target
variable but there is currently no plan to include them in the operational
output of GPROF 2021. For this study, latent heating rates were nonetheless
included in the output of the GPROF-NN retrievals to investigate the feasibility
of the retrieval.

The error statistics for the profile retrievals are displayed in
Fig.~\ref{fig:results_profiles}. For GMI, the results are qualitatively similar
to those observed for the scalar retrieval variables: The GPROF-NN 1D retrieval
has slightly lower biases than than GPROF with the GPROF-NN 3D algorithm
yielding the lowest biases. Similar patterns are observed for MSE, SMAPE and
correlation throughout most of the atmosphere. For rain and snow water content
the SMAPE of the GPROF-NN 3D retrieval increases and even exceeds that of
GPROF-NN 1D at the topmost levels where the hydrometeors are present. This is
presumably due the scarcity of profiles with hydrometeors at these altitudes,
leading to decreased accuracy for the more complex neural network model employed
by the GPROF-NN 3D algorithm.

For MHS, the retrievals exhibit slightly lower accuracy but qualitatively the results
are very similar to those from GMI. One exception are the biases for cloud water
content which are slightly larger than those of GPROF. It is not quite clear what
causes this but given that the biases remain comparable to those of GPROF and the results
for GMI, we do not consider these deviations critical.

\begin{figure}[hbpt]
  \centering
  \includegraphics[width=\textwidth]{figs_revised/fig11}
  \caption{
    Error statistics of the retrieved profile variables. Columns show the
    errors for the different retrieved variables, whereas rows show 
    altitude averaged bias, RMSE, SMAPE, and correlation, respectively.
  }
  \label{fig:results_profiles}
\end{figure}

\subsection{Case study: Hurricane Harvey}

All of the results presented above were based on a test dataset with the same
statistics as the retrieval database. While for GMI the observations can be
expected to be consistent with those in the database, this is not necessarily
the case for  sensors for which the retrieval database contains mostly
simulated observations. While a comprehensive analysis of the retrieval
performance on real observations is outside the scope of this study, this
section presents retrieval results from two overpasses over hurricane Harvey to
provide an indication as to whether the performance characteristics of the
retrieval algorithm can be expected to carry over to real observations.

\begin{figure}[hbpt]
  \centering
  \includegraphics[width=\textwidth]{figs_revised/fig12.png}
  \caption{Surface precipitation in hurricane Harvey on 2017-08-25 at 11:50:00
    UTC retrieved from GMI. Panel (a) shows a GOES 16 Night IR composite
    (generated using the \texttt{night\_ir\_with\_background\_hires} composite
    in Satpy \citep{satpy}), which merges infrared observations from the
    Advanced Baseline Imager (ABI, \citeauthor{schmit05}, \citeyear{schmit05}) on
    GOES 16 and NASA black marble imagery \citep{black_marble}. Panel shows (b)
    ground-based precipitation measurements from MRMS for which the radar
    quality index exceeds 0.8. Panel (c) shows retrieved surface precipitation
    from the CMB product. Panel (d), (e) and (f) show the retrieved surface
    precipitation fields from GPROF, GPROF-NN 1D and GPROF-NN 3D, respectively.}
  \label{fig:harvey_gmi}
\end{figure}

\begin{table}[hbpt]
  \caption{Accuracy metrics for surface precipitation retrieved from GMI PMW
    observations of hurricane Harvey for the overpass on 2017-08-25 at 11:50:00
    UTC. Each metric is calculated with respect to the surface precipitation from
    the CMB product as well as  the surface precipitation
    from MRMS as reference.}
  \label{tab:hurricane_harvey_gmi}
  \begin{tabular}{|l||rr|rr|rr|rr|rr|}
    \hline
    \multicolumn{1}{|c||}{Retrieval} &
    \multicolumn{2}{c|}{Bias [$\unit{mm\ h^{-1}}$]} &
    \multicolumn{2}{c|}{MSE [$(\unit{mm\ h^{-1}})^2$]} &
    \multicolumn{2}{c|}{Correlation} &
    \multicolumn{2}{c|}{Precision} & \multicolumn{2}{c|}{Recall}\\
    & CMB & MRMS & CMB & MRMS & CMB & MRMS & CMB & MRMS & CMB & MRMS\\
    \hline
    \hline
    GPROF       & 0.346 & 0.355 & 2.691 & 8.299 & 0.892 & 0.651 & 0.9 & 0.82 & 0.82 & 0.81 \\
    GPROF-NN 1D & 0.245 & 0.145 & 1.944 & 4.927 & 0.914 & 0.701 & 0.95 & 0.9 & 0.90 & 0.75 \\
    GPROF-NN 3D & 0.248 & 0.184 & 1.953 & 6.12  & 0.923 & 0.676 & 0.95 & 0.9 & 0.90 & 0.87 \\
    \hline
    \end{tabular}
  \end{table}

The first considered overpass is from the GPM Core Observatory over hurricane
Harvey and occurred on August 25, 2017, at 11:50 UTC. Fig.~\ref{fig:harvey_gmi}
shows the retrieved surface precipitation and reference measurements from the
CMB product and MRMS. The retrieved precipitation fields exhibit marked
differences in structure: The GPROF retrieval produces large areas of low
precipitation covering large parts of the scene but not present in the CMB or
MRMS measurements. This is consistent with the overestimation of light
precipitation observed in Fig.~\ref{fig:results_scatter_gmi}. These artifacts
are reduced in the results of the GPROF-NN 1D algorithm and practically absent
in the results of the GPROF-NN 3D retrieval.

A quantitative assessment of the retrieval results is provided in
Tab.~\ref{tab:hurricane_harvey_gmi}, which shows bias, MSE, and correlation, as
well as the precision and recall of the retrieved precipitation flag. The
precision is the fraction of correctly detected raining pixels of all pixels
predicted to be raining, and the recall is the fraction of all truly raining
that is correctly detected.

All statistics were calculated using the CMB product and the MRMS ground-based
measurements as a reference. The reference measurements were averaged to the
footprint of the GMI $18.7\ \unit{GHz}$ channel, taking into account the
rotation of the pixels across the swath. Only measurements with a radar quality
index is at least 0.8 were used for the comparison against MRMS retrievals.

The accuracy of all retrievals is lower when compared to MRMS than when compared
to CMB. This is likely because all GPROF retrievals are designed to reproduce
the retrieval database, which is to a large extent derived from the CMB product.
The GPROF-NN retrievals yield more accurate results than GPROF across all
considered metrics except for the recall, which is lower for GPROF-NN 1D than
for GPROF. Interestingly, GPROF-NN 1D achieves lower MAE, MSE, and bias as well
as higher correlation in the comparison against MRMS, while the two perform
similarly in the comparison against CMB.

\begin{figure}[hbpt]
  \centering
  \includegraphics[width=\textwidth]{figs_revised/fig13}
    \caption{Surface precipitation in hurricane Harvey on 2017-08-25 at 13:58:00
      UTC retrieved from MHS on NOAA-18. Panel (a) shows a natural color composite
      from the ABI on GOES 16 (generated using the \texttt{natural\_color} composite in Satpy
      \citep{satpy}). Panel shows (b) ground-based precipitation measurements
      from MRMS for which the radar quality index is at least 0.8. Panel (c) shows
      retrieved surface precipitation from the CMB product. Panel (d), (e) and
      (f) show the retrieved surface precipitation fields from GPROF, GPROF-NN
      1D and GPROF-NN 3D, respectively.}
  \label{fig:harvey_mhs}
\end{figure}

\begin{table}[hbpt!]
  \caption{Accuracy metrics for surface precipitation retrieved from MHS PMW
    observations of hurricane Harvey for the overpass on 2017-08-25 at 13:58 UTC.
    The metrics calculated  against the MRMS surface precipitation estimates.}
  \label{tab:hurricane_harvey_mhs}
  \begin{tabular}{|l||r|r|r|r|r|}
    \hline
    Retrieval &
    Bias [$\unit{mm\ h^{-1}}$] &
    MSE [$(\unit{mm\ h^{-1}})^2$] &
    Correlation &
    Precision &
    Recall \\
    \hline
    \hline
    GPROF       & 0.11   & 2.602  & 0.749  & 0.88 & 0.12 \\
    GPROF-NN 1D & 0.259  & 4.031  & 0.751  & 0.9057 & 0.094 \\
    GPROF-NN 3D & 0.152  & 3.168  & 0.759  & 0.948 & 0.052 \\
    \hline
    \end{tabular}
  \end{table}

Fig.~\ref{fig:harvey_mhs} presents retrieved surface precipitation from an
overpass of the MHS sensor on board the NOAA-18 satellite over the same storm at
13:58 UTC. Because no co-located CMB measurements are available for this
overpass, only the MRMS measurements are shown as reference measurements. GPROF
predicts low precipitation rates across large parts of the scene and even in
cloud-free areas. This tendency is reduced in the results of the GPROF-NN 1D
retrieval and even more so in the results of GPROF-NN 3D. The GPROF-NN
retrievals also generally yield better agreement with MRMS over land.
Furthermore, the rain bands of the hurricane are better defined in the results
of the GPROF-NN 3D retrievals, which is consistent with the increased effective
resolution of the retrievals.

Accuracy metrics for comparing the MHS retrievals with MRMS are shown in
Tab.~\ref{tab:hurricane_harvey_mhs}. The MRMS measurements were averaged to the
MHS observation footprints taking into account the changes in footprint size and
shape across the swath. For MHS, GPROF has the lowest Bias, MAE, and MSE and
higher recall than GPROF-NN 1D. These results do not show any clear improvements
for the GPROF-NN retrievals. However, the GPROF-NN 3D retrievals improve the
retrieval in terms of all metrics compared to GPROF-NN 1D, suggesting that the
GPROF-NN 3D can make use of the spatial information in the observations despite
being trained on simulated observations.

\subsection{Processing time}

GPROF is used to process PMW observations from a constellation of sensors
spanning several decades of observations. Therefore, the processing time must
not be excessively high. Although neural networks are generally  efficient to
evaluate, this often assumes dedicated hardware, which can not yet be expected
to be available at the processing centers.

We measure the processing time required for retrieving precipitation from a full
orbit of observations using a single CPU core of an Intel Xeon Gold 6234 CPU to
assess the computational complexity of the three retrievals. The processing time
here includes all steps from reading a GPROF input file to writing the
corresponding output file. The input and output files are the same for all three
algorithms, excluding, of course, differences in the retrieval results.

The results are displayed in Fig.~\ref{fig:processing_time}. The processing of a
single GMI file takes about 4 minutes for GPROF but only about 2 minutes for the
GPROF-NN retrievals. Because of the lower number of pixels in a single orbit,
all retrievals are significantly faster for MHS. However, also here the GPROF-NN
retrievals are significantly faster than GPROF. This shows that, even in the
absence of dedicated hardware, the GPROF-NN retrievals process observations
faster than the current implementation.

\begin{figure}[hbpt]
  \centering
  \includegraphics[width=0.5\textwidth]{figs_revised/fig14}
  \caption{Single CPU core processing time for an orbit of observations  for
    the three retrieval algorithms for GMI and MHS. Error bars show the range of
  one standard deviation around the mean for five executions of each retrieval.}
  \label{fig:processing_time}
\end{figure}

\section{Discussion}

This study presented two novel, neural network based implementations of the
GPROF retrieval algorithm and evaluated their performance for the GMI and MHS
sensors against the current implementation.

\subsection{Retrieval performance}

The evaluation of the GPROF-NN 1D algorithm against GPROF, showed that retrieval
accuracy as well as effective resolution can be improved by replacing the
current retrieval method with a fully-connected neural network. Although both
GPROF and the GPROF-NN 1D algorithm are based on the same retrieval database and
use the same information as retrieval input, the neural network provides more
accurate results. A potential explanation for this may lie in the way the two
algorithms handle observation uncertainties, which were shown by
\citet{elsaesser15} to have a significant effect on the retrieval accuracy. For
GPROF, these uncertainties must be specified manually. Apart from sensor noise,
observations from other sensors than GMI are affected by modeling errors in the
simulated observations, which are difficult to estimate and unlikely to be well
described by the assumed Gaussian error model. In addition to this,
uncertainties are inflated to account for the sparsity of the retrieval database
and the effects of clustering. Since samples with low precipitation rates are
generally better represented in the database, this likely makes the
uncertainties too large for the retrieval of low precipitation rates, which may
explain the inferior performance of GPROF for low precipitation rates observed
in Fig.~\ref{fig:results_scatter_gmi} and Fig.~\ref{fig:results_scatter_mhs}.
The neural network based algorithms infer observation errors directly from the
data, and can thus handle arbitrary observation errors, given that they are
accurately represented in the training data.

%This aspect highlights an important aspect concerning the potential of the
%GPROF-NN algorithm to improve the retrieval on real observations: For the
%improvements to carry over to real observations it is crucial that the
%observation error is represented correctly in the training data. Although this
%does not mean that the simulations must be error free, it requires care to be
%taken to accurately represent simulation uncertainties. While this is given for
%GMI because the observations used during training are real observations, it is
%much more difficult for sensors for which the observations are simulated.

Another advantage of the neural network retrievals that may explain the improved
accuracy is that they scale more easily to large retrieval databases. While the
GPROF algorithm requires compressing the retrieval database in a way that causes
information loss, the training of the neural networks uses the full database.
However, even in the absence of clustering, \citet{pfreundschuh18}
provided empirical evidence that neural network based retrievals are less
affected by the curse of dimensionality, which means they yield more accurate
results when limited data is available. Although the GPROF database is fairly
large, heavy precipitation events are likely still underrepresented, which may
be exacerbated by the clustering performed by GPROF. In this context, it may
also be worth pointing out that, since the database size influences only the
training time of the GPROF-NN algorithms, they can potentially be applied with
even larger retrieval databases than the one currently used, which may help to
further improve the retrieval accuracy in the future.

The second important finding from this study is that by extending the retrieval
to incorporate structural information, its accuracy can be further improved by
about $20\ \unit{\%}$ in terms of MAE, MSE and SMAPE and $5\ \unit{\%}$ in terms
of correlation compared to the GPROF-NN 1D retrieval at the same time as the
effective resolution in along track direction is decreased to its lower limit of
$13.5\ \unit{km}$ for GMI and improved by $70\ \unit{\%}$ for MHS. Because
precipitation exhibits distinct spatial patterns in satellite observations, many
algorithms make use of this information to improve precipitation retrievals
\citep{kummerow94, sorooshian00, hong04, kaushik10}. Our results confirm that CNNs learn to
leverage this information directly from the satellite imagery and that it can
notably improve the retrieval accuracy, which is in agreement with the findings
from other precipitation retrievals that employ CNNs \citep{tang18, sadeghi19,
  gorooh22, sano18}.

As a concluding remark regarding the retrieval performance, it should also be
noted that this study focused on the development of a generic retrieval
algorithm applicable to all sensors of the GPM constellation within the
operational constraints of the current GPROF retrieval. This means that the used
neural network models were not optimized exhaustively and that the performance
of neural network based PMW precipitation retrievals can likely be improved
further by dedicated tuning of the architecture.

\subsection{Limitations}

It is important to consider the limitations of the results presented in this
study. We have deliberately limited the evaluation of the retrieval accuracy to
test data with the same statistical properties as the retrieval database. This
was done to isolate the effect of the retrieval method from potential aliasing
effects that would be introduced by the use of external validation data. The
presented retrieval accuracy should therefore be interpreted as an upper bound
on the accuracy that can be achieved with respect to external validation data.
Since the GMI retrieval is trained using real observations, the performance on
real observations can be expected to be close to the results presented, which
was confirmed by the results from the GMI overpass over Hurricane Harvey
(Fig.~\ref{fig:harvey_gmi}).

For other sensors, however, the observations in the database can only be
simulated and may deviate significantly from true observations. As described in
Sec.~\ref{app:training_data}, the training of the GPROF-NN 3D retrieval requires
an additional neural network model to generate simulated observations of
sufficiently large extent to train a CNN. The results presented in
Sec.~\ref{sec:results_test} should therefore be seen as an assessment of the
potential benefits of a CNN-based retrieval given a perfect retrieval database
rather than the real-world retrieval accuracy.

The quantitative assessment of the accuracy of the MHS retrievals of hurricane
Harvey did not show any clear improvements for the GPROF-NN retrievals compared
to GPROF. This can be due to multiple reasons. Firstly, the hurricane
constitutes an extreme event and it is likely that the instantaneous MRMS
precipitation rates used as reference measurements are themselves affected by
considerable uncertainties. Secondly, given that the bulk of the precipitation
in the considered scene is intense and over ocean, GPROF can be expected to work
quite well. This makes it less likely to find clear improvements in this
particular scenario. Finally, the accuracy of the neural-network based
retrievals may be limited by the modeling error of the simulations in the
retrieval database. In principle, simulation errors could even cause the
GPROF-NN retrievals to be less accurate than GPROF for real observations. Should
this really be the case, the demonstrated potential of the GPROF-NN retrievals
would imply that the quality of the simulations in the GPROF database limits the
accuracy of the GPM PMW precipitation measurements and that future work to
should focus on improving the simulations.

\conclusions  %% \conclusions[modified heading if necessary]

The results presented in this study clearly demonstrate the potential of a
neural-network-based implementation of GPROF to improve accuracy and
effective resolution of retrievals of precipitation and hydrometeor profiles.
Both GPROF-NN retrievals have been designed as a drop-in replacement for GPROF
and can be directly used in the operational GPM processing pipeline. The results
presented in this study show that, given a perfect retrieval database,
considerable improvements in the accuracy of GPROF can be achieved  by
replacing the current Bayesian scheme with a deep neural network that processes
pixels independently. In addition to that, further improvements of similar
magnitude can be achieved with a CNN-based implementation, which incorporates
structural information into the retrieval.

Although the results presented here cannot fully answer the question to what
extent the improvements observed for the GPROF-NN algorithms carry over to
operational application of the retrievals, they show the potential of the
neural-network-based PMW retrievals. Upgrading GPROF to a neural-network-based
retrieval thus has the potential of being a very cost efficient way to improve
global measurements of precipitation with the added advantage of being
applicable even to historical observations. Furthermore, the results provide an
important reference point, which, together with a future evaluation of the
retrievals against independent measurements, is required to inform further
development aiming to improve the accuracy of GPM PMW retrievals.

The GPROF implementations presented in this study constitute a first step
towards a potential upgrade of GPROF to a neural-network-based implementation.
The next step will be to run the GPROF-NN retrievals along side GPROF 2021 for
all sensors of the GPM constellation and to validate the retrieval results
against independent validation data. The Python based software package that
implements the retrieval and training framework is made available together with
all trained models as free software \citep{gprof_nn}.

Although the effective improvements that will be achieved in operational use
still remain to be investigated, we take the results presented here as a
promising indication of the potential of the GPROF-NN retrievals to improve PMW
retrievals from the sensors of the GPM constellation. These algorithms may thus
constitute a step towards improving our ability to measure the global
hydrological cycle and its changes in a warming climate.

%% The following commands are for the statements about the availability of data sets and/or software code corresponding to the manuscript.
%% It is strongly recommended to make use of these sections in case data sets and/or software code have been part of your research the article is based on.

\codeavailability{
  The implementation of the GPROF-NN retrievals is published is as free
  software online \citep{gprof_nn}.
} %% use this section when having only software code available

\appendix

\section{The GPROF Bayesian retrieval scheme}    %% Appendix A
\label{app:gprof}

At the base of the GPROF is a Bayesian retrieval method based on Monte Carlo
integration of the profiles in the retrieval database. The database is split
into bins using the ancillary data to reduce the number of profiles that must be
processed for each pixel and better constrain the retrieval. Moreover, the
profiles are clustered to further reduce the number of profiles to process.
Fig.~\ref{fig:gprof_legacy} illustrates the three components of the GPROF
retrieval.

\begin{figure}[hbpt!]
  \centering \includegraphics[width=\textwidth]{figs_revised/fig15}
  \caption{Components of the GPROF retrieval algorithm. Panel (a) illustrates
    the binning of the database with respect to the ancillary data, which
    consists of two meter temperature ($T_\text{2m}$), total column water vapor
    (TCWV), surface type and airlifting index. Panel (b) illustrates the
    clustering of each database bin into self-similar clusters with the size of
    the markers representing the number of profiles in each cluster. Panel (c)
    illustrates the Bayesian scheme that is used to approximate the posterior
    distribution of the retrieval, which corresponds to the filled curve to the
    right, by weighting the samples in the database. }
  \label{fig:gprof_legacy}
\end{figure}

The binning of the profiles in the database is performed with respect to all
ancillary variables, that is $T_\text{2m}$, $\text{TCWV}$, surface type and
airlifting index (Fig.~\ref{fig:gprof_legacy} (a)). Each bin covers a range of
$1\ \unit{K}$ in $T_\text{2m}$ and $1\ \unit{kg} m^{-2}$ in $\text{TCWV}$ around
the closest, corresponding integral value. If a bin contains less than 30 000 profiles, its
profiles are combined with those from bins with neighboring $T_\text{2m}$ and
$\text{TCWV}$ values.

The profiles in each  bin are combined into self-similar clusters and only
the mean of the observations and retrieval targets as well as the number of
observations is retained (Fig.~\ref{fig:gprof_legacy} (b)). The hierarchical
clustering merges profiles with similar observations until the number of clusters
per bin is less than 800.

The binning and clustering of the database is performed offline, i.e. during the
development phase of the retrieval. The following scheme is applied to retrieve
precipitation and hydrometeor profiles from the clustered database bins. The
first step in the retrieval is the determination of the database bins to be
used. The central bin is found from the surface type and airlifting index and
the rounded input $T_\text{2m}$ and $\text{TCWV}$ values. Profiles from this
central bin and the two neighboring $T_\text{2m}$ bins are considered for the
retrieval.

Let $(\mathbf{y}_1, \mathbf{x}_1, n_1), \ldots, (\mathbf{y}_N, \mathbf{x}_N,
n_N)$ denote the profile clusters in the selected database bin, where
$\mathbf{y}_i$ and $\mathbf{x}_i$ are, respectively, the centroids of the
observation and state vector and $n_i$ the corresponding number of profiles in
the $i$th cluster .
Assuming that the  profiles in the database bins are distributed
according to the a priori distribution $p(\mathbf{x})$, the expected value of
$\mathbf{x}$ with respect to the corresponding posterior distribution
$p(\mathbf{x} | \mathbf{y})$ can be approximated using
\begin{align}\label{eq:gprof_retrieval}
  \int_{\mathbf{x} } \mathbf{x} p(\mathbf{x} | \mathbf{y})\: d\mathbf{x} =
  \int_{\mathbf{x} } \mathbf{x}\: \frac{p(\mathbf{y} |
    \mathbf{x})p(\mathbf{x})}{p(\mathbf{y})}\: d\mathbf{x} \approx \frac{\sum_i
    p(\mathbf{y}|\mathbf{x}_i) \mathbf{x}_i}{\sum_i
    p(\mathbf{y}|\mathbf{x}_i)}.
\end{align}
The conditional probability $p(\mathbf{y} | \mathbf{x}_i)$ of the input
observation $\mathbf{y}$ given atmospheric state $\mathbf{x}_i$ is taken as the
probability of the deviations of $\mathbf{y}$ from the observations
$\mathbf{y}_i$ to be caused by the random error in the observations, which is
assumed to be unbiased and Gaussian:
  %
\begin{align}\label{eq:gprof_error}
  p(\mathbf{y}|\mathbf{x}_i) =
  \frac{n_i}{\sqrt{\text{det}(2\pi\mathbf{S})}} \exp \left \{ - \frac{1}{2}
  (\mathbf{y} - \mathbf{y}_i)^T \mathbf{S}^{-1} (\mathbf{y} - \mathbf{y}_i)
  \right \}
\end{align}
where $\mathbf{S}$ is a diagonal covariance matrix. The observation error
includes sensor noise as well as other causes of deviations of real
observation from the observations in the database, such as calibration errors or
modeling errors. It should be noted here, that the assumption of Gaussian errors
with state independent, diagonal covariance matrix is made for simplicity but
likely insufficient to accurately describe modeling errors that are state
dependent and correlated between channels. As illustrated in
Fig~\ref{fig:gprof_legacy} (c), Eq.~\ref{eq:gprof_retrieval} corresponds to a
resampling of the states in the database with case-specific weights calculated
using Eq.~\ref{eq:gprof_error}. This approach can be extended to approximate the
probability density function of the posterior distribution or to derive
probabilities of certain characteristics of the a posteriori state such as the
presence of precipitation in a given observation.

\section{Implementation of the GPROF-NN retrievals}    %% Appendix A

\subsection{Training data}     %% Appendix A1, A2, etc.
\label{sec:gprof_nn_training_data}

\subsubsection{Structure}

The training data for the GPROF-NN retrievals is stored in an intermediate format to simplify the loading of the data during the training process. The data is organized into scenes measuring 221 contiguous GMI pixels in both along- and across-track directions. Each scene contains the GMI L1C brightness temperatures and the corresponding values of the retrieval quantities at the center of the GMI swath. For sensors other than GMI, each scene also contains the simulated brightness temperatures of the corresponding sensor.

\subsubsection{Generation}
\label{app:training_data}

An overview of the data flow for the training data generation for the GPROF-NN retrievals is displayed in Fig.~\ref{fig:data_flow}. The training data originates from four primary sources: The GPROF simulator files, which contain surface precipitation, hydrometeor profiles, and simulated brightness temperatures for an orbit of the GPM combined product. Surface precipitation over snow surfaces and sea-ice are derived from MRMS and ERA5 data, respectively. This data is matched with GMI L1C-R brightness temperatures. The data is split into non-overlapping scenes measuring 221 scans and 221 pixels. For sensors other than GMI, the brightness temperature differences between actual and simulated GMI observations are included and added to the simulated observations to provide a first-order correction for the modeling error in the observations.

Simulated brightness temperatures are only available where the hydrometeor profiles and surface precipitation is known, i.e., at the center of the training scenes. Because this is insufficient to train a CNN with 2D convolutions for sensors other than GMI, an intermediate simulator retrieval is trained to retrieve simulated brightness temperatures from GMI observations. This retrieval the applied to the training data to fill in the simulated brightness temperatures across the entire GMI swath. The simulator neural network uses the same architecture as GPROF-NN 3D retrieval.

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{figs_revised/fig16}
  \caption{
    Data flow diagram for the generation and organization of the GPROF-NN training data. Grey rectangles represent datasets, and colored rectangles with rounded corners represent algorithms.
  }
  \label{fig:data_flow}
\end{figure}

\subsection{Training}     %% Appendix A1, A2, etc.
\label{sec:gprof_nn_training}

\begin{table}
  \caption{Sizes of neural network models and the training data.}
  \label{tab:parameters}
  \begin{tabular}{|l||cc|}
    \hline
    & Model parameters (GMI) &  Training samples \\
    \hline
    \hline
    GPROF-NN 1D & $5\,453\,056$ &  $2\,136\,604\,660$ pixels \\
    \hline
    GPROF-NN 3D & $23\,855\,792$ &  $86\,350$ scenes \\
    \hline
  \end{tabular}
\end{table}

Tab.~\ref{tab:parameters} lists the number of parameters of the neural networks
used in the GPROF-NN retrievals together with the number of samples in the
training data. Owing to its more complex network architecture, the neural
network employed by GPROF-NN 3D has a larger number of parameters. The training
data comprises $86\,350$ scenes of $221\times 221$ GMI pixels. From those
scenes, only the pixels with known surface precipitation are used for the
training of the GPROF-NN 1D retrieval. The total number of those amounts to
$2\,136\,604\,660$. The

\subsubsection{GPROF-NN 1D}

The GPROF-NN 1D network is trained by simultaneously minimizing the sum of the
losses of all retrieval variables. The training is performed over 70 epochs
using the Adam optimizer \citep{kingma14} with an initial learning rate of
$5\cdot10^{-4}$ and a cosine annealing learning rate schedule
\citep{loshchilov16}. Warm restarts are performed after 10, 30 and, 50 epochs.

The following pre-processing steps are performed when the training data for the
GPROF-NN 1D retrieval are loaded. The steps are performed anew for each training
epoch. A detailed explanation follows.

\begin{enumerate}
  \item Extract pixels with known surface-precipitation
  \item For cross-track scanners: Sample earth-incidence angle and interpolate in- and outputs
  \item For sensors other than GMI: Application of the brightness temperature correction
  \item Input normalization and encoding
  \item Replacement of zeros
  \item For sensors other than GMI: Addition of thermal noise
  \item Shuffle training samples
\end{enumerate}

 The retrieval outputs in the GPROF-NN training data are known only at a limited
 number of pixels at the center of each scene. Only these pixels are extracted
 from each scene in step (2). For cross-track scanning sensors, the training data
 contains retrieval in- and output for a sequence of discrete earth-incidence
 angles. A random earth-incidence angle is generated for each training sample,
 and the in- and outputs are interpolated to that angle (2). If the sensor
 relies on simulations, the brightness temperatures are corrected using the
 method described in Sec.~\ref{sec:other_sensors}. The retrieval inputs are then
 encoded and normalized (4, see Sec.~\ref{sec:input_normalization}). Zero values
 of non-negative retrieval quantities are replaced by very small, random values
 to avoid degenerate quantiles and problems with the application of the
 log-linear transformation Eq.~(\ref{eq:log_linear}) (5). Finally, thermal noise
 according to sensor specification is added to simulated observations for
 sensors other than GMI (6) and the loaded samples are shuffled (7).

\subsubsection{GPROF-NN 3D}

 The training of the GPROF-NN 3D retrieval is performed over 70 epochs using the
 Adam optimizer \citep{kingma14} with an initial learning rate of
 $5\cdot10^{-4}$ and a cosine annealing learning rate schedule
 \citep{loshchilov16}. Warm restarts are performed after 10, 30 and 50 epochs.

The following pre-processing steps are performed when the training data for the
GPROF-NN 3D retrieval are loaded. These steps are performed anew for each
training epoch. A detailed explanation follows.

\begin{enumerate}
\item Remapping of observations to viewing geometry of sensor
\item For sensors other than GMI: Application of the brightness temperature correction
\item Input normalization and encoding
\item Replacement of zeros
\item For sensors other than GMI: Addition of thermal noise and simulator error
\item Shuffle training samples
\end{enumerate}

Each training scene is randomly remapped from the GMI swath to the viewing
geometry of the sensor for which the training is performed (1, see
Sec.~\ref{sec:remapping}). If the sensor relies on simulations,
the correction described in Sec.~\ref{sec:other_sensors} is applied to
the brightness temperatures.
The
input for each scene is then encoded and normalized (3., see
Sec.~\ref{sec:input_normalization}). Zero values of non-negative retrieval
quantities are replaced by very small, random values to avoid degenerate
quantiles and avoid problems with the application of the log-linear
transformation Eq.~(\ref{eq:log_linear}) (4). Thermal noise and a
simulator error are added to the simulated observations for sensors other than
GMI (5). The simulator error is modeled to be constant across each scene
and determined from the MSE of the simulator network on the training data.
Finally, the samples are shuffled (7).

\subsubsection{Viewing geometry remapping}
\label{sec:remapping}

Because the largest part of the GPROF retrieval database is derived from
collocations of GMI observations with the GPM CMB product, the spatial sampling
of most training scenes corresponds to that of the GMI L1C-R product regardless
of the sensor for which the database was generated. The viewing geometry of the
observations in the database does therefore not match that of the other sensors
of the GPM constellation. In addition to that, the values of the retrieval
targets are only known at the center of the GMI swath. The distortions that
occur towards the sides of the swath of GMI and the other sensors are therefore
not well represented in the training data.

A custom data augmentation scheme is applied  to overcome these limitations, which
consists of a random remapping of the scenes  to the viewing
geometry of the target sensor. The remapping is implemented as follows.
\begin{enumerate}
\item A random center location $c_\text{out}$ in the swath of the target sensor is sampled.
\item The approximate positions $\mathbf{p_\text{out}}$ of $h \times w$ pixels
  in the swath of the target sensor  in a two-dimensional Euclidean coordinate system
  centered on $c_\text{out}$ are calculated.
\item A random center location $c_\text{in}$ in the GMI swath is sampled.
\item The approximate positions $\mathbf{p_\text{in}}$ of all GMI pixels in the
  training scene in a two-dimensional Euclidean coordinate system centered on
  $c_\text{in}$ are calculated.
\item The retrieval in- and outputs are interpolated from the
  positions $\mathbf{p_\text{in}}$ to the positions $\mathbf{p_\text{out}}$.
\item For cross-track scanning sensors, the simulated brightness temperatures and
  retrieval outputs are interpolated to the earth-incidence angles corresponding
  to the positions $\mathbf{p_\text{out}}$ in the output window.
\end{enumerate}

The height $h$ and $w$ of the output window for GMI is 128 in the along-track
direction and 96 in the across-track direction. Since many sensors have
considerably wider swaths than GMI, the size of the output window is adapted to
avoid that too many pixels lie outside the GMI swath. The width of the output
window in across-track direction for MHS was set to 32 pixels.

\subsubsection{Input normalization and encoding}
\label{sec:input_normalization}

The brightness temperatures and scalar ancillary data that constitute the input to the
retrieval are normalized using minimum-maximum
normalization. For each scalar input  $x$, the minimum $x_\text{min}$ and maximum
$x_\text{max}$ values in the training data are calculated. The values are then normalized to
the range $[-1, 1]$ using
\begin{align}
  x_\text{normalized} &= \frac{x - x_\text{min}}{x_\text{max} - x_\text{min}}.
\end{align}
Missing values in the input are set to the value -1.5. Categorical ancillary data, i. e,
the surface type and air-lifting index, are encoded using one-hot encoding.

\subsection{Retrieval processing}     %% Appendix A1, A2, etc.
\label{sec:gprof_nn_processing}

The data flow for the application of the GPROF and GPROF-NN retrievals is
displayed in Fig.~\ref{fig:data_flow_application}. The first step, which is
common for all three retrievals, is the augmentation of the GPM L1C data with
ancillary data. This process is performed by the GPROF preprocessor application.
A detailed description of the ancillary data and its derivation can be found in
the GPROF ATBD \citep{atbd}. The GPROF preprocessor produces a binary file
containing the observations and ancillary data. This file serves as input for
both GPROF and the GPROF-NN retrievals.

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{figs_revised/fig17}
  \caption{
    Data flow diagram for the application of the GPROF and GPROF-NN retrievals.
    The input for all retrieval is a GPROF preprocessor file, which is a binary file that
    contains the brightness temperatures and corresponding ancillary data. From this input
    all retrievals produce the retrieval results, which are stored in a common binary
    format before being converted to HDF5 files.
  }
  \label{fig:data_flow_application}
\end{figure}

\subsubsection{GPROF-NN 1D}

The processing of input observations for the GPROF-NN 1D retrieval involves the
following steps.

\begin{enumerate}
  \item Flattening of retrieval inputs
  \item Input normalization and encoding
  \item Batch-wise evaluation of network and calculation of posterior statistics 
  \item Re-assembly into swath structure
  \item Writing of GPROF binary output file
\end{enumerate}

The observations and corresponding ancillary data are flattened into a list of
inputs (1). All inputs are normalized and the categorical input variables are
one-hot encoded using the same statistics as during training (2). The GPROF-NN
1D network is then used to calculate the posterior distributions of the
retrieval targets from which the relevant posterior statistics are derived
(3). Finally, the results for each pixel are re-assembled into the original
swath structure and written to the GPROF binary output format, which is
 converted to HDF5 format in a separate step.

\subsubsection{GPROF-NN 3D}

The processing of input observations for the GPROF-NN 3D retrieval involves the
following steps.

\begin{enumerate}
\item Input normalization and encoding
\item Input padding
\item Evaluation of network and calculation of the posterior statistics 
\item Removal of padding
\end{enumerate}

The input observations and ancillary data are normalized and encoded using the
same statistics as during the training. The input observations are then padded
using symmetric padding so that the dimension of the input data are a multiple
of 32, which is required to ensure of symmetry requirements of the down- and
up-sampling transformation in the neural network. The GPROF-NN 3D network
is then evaluated and the posterior statistics are calculated. Because the
GPROF-NN 3D network employs a fully-convolutional architecture, the results can
be calculated for a full orbit of observations at once. However, since this may
require excessive amounts of memory, the processing allows for optional tiling
of the processing in along-track direction. After removal of the padding, the
retrieval results are written to the same binary format that is used by GPROF-NN
1D and GPROF.

\section{Error metrics}

\appendixfigures  %% needs to be added in front of appendix figures

\begin{figure}[hbpt]
  \centering
  \includegraphics[width=\textwidth]{figs_revised/fig18}
  \caption{Like Fig.~\ref{fig:results_spatial_gmi} but for MHS}
  \label{fig:results_spatial_mhs}
\end{figure}

\appendixtables   %% needs to be added in front of appendix tables

\begin{table}[hbpt!]
  \centering
  \caption{Like Tab.~\ref{tab:metrics_gmi} but for convective precipitation.}
  \label{tab:metrics_gmi_convective}
\begin{tabular}{|c||p{3.5cm}|p{3.5cm}|p{3.5cm}|}
  \hline
  Metric &
  \multicolumn{1}{|c}{GPROF} &
  \multicolumn{1}{|c}{GPROF-NN 1D} &
  \multicolumn{1}{|c|}{GPROF-NN 3D} \\
  \hline\hline
  Bias \hfill [$\unit{mm\ h^{-1}}$]   & \hfill $ -0.0007 \pm 0.0001$ &\hfill $ -0.0015 \pm 0.0001$ &\hfill $ -0.0011 \pm 0.0001$ \\
  MAE \hfill [$\unit{mm\ h^{-1}}$]    & \hfill $  0.0322 \pm 0.0001$ &\hfill $  0.0239 \pm 0.0001$ &\hfill $  0.0204 \pm 0.0001$ \\
  MSE \hfill [$\unit{mm\ h^{-1}}$]    & \hfill $  0.1927 \pm 0.0001$ &\hfill $  0.1298 \pm 0.0001$ &\hfill $  0.0854 \pm 0.0001$ \\
  MAPE$_{0.01}$ \hfill [$\unit{\%}$]  & \hfill $118.151 \pm 0.0391$ &\hfill $107.1976 \pm 0.0378$ &\hfill $ 92.8343 \pm 0.0542$ \\
  Correlation & \hfill $  0.6380 $ &\hfill $  0.7467 $ &\hfill $  0.8152 $ \\
  \hline
\end{tabular}
\end{table}

\begin{table}[hbpt!]
  \centering
  \caption{Like Tab.~\ref{tab:metrics_gmi} but for RWP.}
  \label{tab:metrics_gmi_rwp}
  \begin{tabular}{|c||p{3.5cm}|p{3.5cm}|p{3.5cm}|}
    \hline
    Metric &
    \multicolumn{1}{|c}{GPROF} &
    \multicolumn{1}{|c}{GPROF-NN 1D} &
    \multicolumn{1}{|c|}{GPROF-NN 3D} \\
    \hline\hline
    Bias \hfill [$\unit{mm\ h^{-1}}$]  & \hfill $  0.0016 \pm 0.0000$ &\hfill $ -0.0005 \pm 0.0000$ &\hfill $ -0.0003 \pm 0.0000$ \\
    MAE \hfill [$\unit{mm\ h^{-1}}$]   & \hfill $  0.0185 \pm 0.0000$ &\hfill $  0.0127 \pm 0.0000$ &\hfill $  0.0094 \pm 0.0000$ \\
    MSE \hfill [$\unit{mm\ h^{-1}}$]   & \hfill $  0.0120 \pm 0.0000$ &\hfill $  0.0086 \pm 0.0000$ &\hfill $  0.0047 \pm 0.0000$ \\
    MAPE$_{0.001}$ \hfill [$\unit{\%}$] & \hfill $ 84.072 \pm 0.0287$ &\hfill $ 69.6918 \pm 0.0284$ &\hfill $ 61.8979 \pm 0.0315$ \\
    Correlation & \hfill $  0.8308$ &\hfill $  0.8777$ &\hfill $ 0.9241$ \\
    \hline
  \end{tabular}
\end{table}
\clearpage

%% Please add \clearpage between each table and/or figure. Further guidelines on figures and tables can be found below.

\begin{table}[hbpt!]
  \centering
\caption{Like Tab.~\ref{tab:metrics_gmi} but for IWP.}
\label{tab:metrics_gmi_iwp}
\begin{tabular}{|c||p{3.5cm}|p{3.5cm}|p{3.5cm}|}
  \hline
  Metric &
  \multicolumn{1}{|c}{GPROF} &
  \multicolumn{1}{|c}{GPROF-NN 1D} &
  \multicolumn{1}{|c|}{GPROF-NN 3D} \\
  \hline\hline
  Bias \hfill [$\unit{mm\ h^{-1}}$]   & \hfill $ -0.0022 \pm 0.0000$ &\hfill $ -0.0006 \pm 0.0000$ &\hfill $ -0.0002 \pm 0.0000$ \\
  MAE \hfill [$\unit{mm\ h^{-1}}$]    & \hfill $  0.0204 \pm 0.0000$ &\hfill $  0.0123 \pm 0.0000$ &\hfill $  0.0085 \pm 0.0000$ \\
  MSE \hfill [$\unit{mm\ h^{-1}}$]    & \hfill $  0.0186 \pm 0.0000$ &\hfill $  0.0123 \pm 0.0000$ &\hfill $  0.0053 \pm 0.0000$ \\
  MAPE$_{0.001}$ \hfill [$\unit{\%}$] & \hfill $ 88.26 \pm 0.0312$ &\hfill $ 67.3705 \pm 0.0305$ &\hfill $ 58.5831 \pm 0.0334$ \\
  Correlation & \hfill $0.7897$ &\hfill $0.8637$ &\hfill $  0.9350$ \\
  \hline
\end{tabular}
\end{table}

\clearpage

\begin{table}[hbpt!]
  \centering
  \caption{Like Tab.~\ref{tab:metrics_gmi} but for CWP.}
  \label{tab:metrics_gmi_cwp}
  \begin{tabular}{|c||p{3.5cm}|p{3.5cm}|p{3.5cm}|}
    \hline
    Metric &
    \multicolumn{1}{|c}{GPROF} &
    \multicolumn{1}{|c}{GPROF-NN 1D} &
    \multicolumn{1}{|c|}{GPROF-NN 3D} \\
    \hline\hline
    Bias \hfill [$\unit{mm\ h^{-1}}$]   & \hfill $ -0.0019 \pm 0.0000$ &\hfill $ -0.0005 \pm 0.0000$ &\hfill $ -0.0005 \pm 0.0000$ \\
    MAE \hfill [$\unit{mm\ h^{-1}}$]    & \hfill $  0.0268 \pm 0.0000$ &\hfill $  0.0157 \pm 0.0000$ &\hfill $  0.0115 \pm 0.0000$ \\
    MSE \hfill [$\unit{mm\ h^{-1}}$]    & \hfill $  0.0027 \pm 0.0000$ &\hfill $  0.0015 \pm 0.0000$ &\hfill $  0.0009 \pm 0.0000$ \\
    MAPE$_{0.001}$ \hfill [$\unit{\%}$]  & \hfill $ 62.2267 \pm 0.0100$ &\hfill $ 36.6584 \pm 0.0078$ &\hfill $ 27.9016 \pm 0.0087$ \\
    Correlation & \hfill $  0.8709$ &\hfill $  0.9265$ &\hfill $  0.9531$ \\
    \hline
  \end{tabular}
\end{table}
\clearpage

\begin{table}[hbpt!]
  \centering
  \caption{Like Tab.~\ref{tab:metrics_mhs} but for convective precipitation.}
  \label{tab:metrics_mhs_convective}
\begin{tabular}{|c||p{3.5cm}|p{3.5cm}|p{3.5cm}|}
  \hline
  Metric &
  \multicolumn{1}{|c}{GPROF} &
  \multicolumn{1}{|c}{GPROF-NN 1D} &
  \multicolumn{1}{|c|}{GPROF-NN 3D} \\
  \hline\hline
  Bias \hfill [$\unit{mm\ h^{-1}}$] & \hfill $ -0.0046 \pm 0.0001$ &\hfill $ -0.0023 \pm 0.0001$ &\hfill $ -0.0012 \pm 0.0001$ \\
  MAE \hfill [$\unit{mm\ h^{-1}}$] & \hfill $  0.0330 \pm 0.0001$ &\hfill $  0.0281 \pm 0.0001$ &\hfill $  0.0210 \pm 0.0001$ \\
  MSE \hfill [$\unit{mm\ h^{-1}}$] & \hfill $  0.1674 \pm 0.0001$ &\hfill $  0.1337 \pm 0.0001$ &\hfill $  0.0824 \pm 0.0001$ \\
  SMAPE$_{0.01}$ \hfill [$\unit{\%}$] & \hfill $108.8755 \pm 0.0480$ &\hfill $104.2921 \pm 0.0507$ &\hfill $ 94.0801 \pm 0.1057$ \\
  Correlation & \hfill $  0.5927 $ &\hfill $  0.6839$ &\hfill $  0.7336$ \\
  \hline
\end{tabular}

\end{table}

\clearpage

\begin{table}[hbpt!]
  \centering
  \caption{Like Tab.~\ref{tab:metrics_mhs} but for RWP.}
  \label{tab:metrics_mhs_rwp}
  \begin{tabular}{|c||p{3.5cm}|p{3.5cm}|p{3.5cm}|}
    \hline
    Metric &
    \multicolumn{1}{|c}{GPROF} &
    \multicolumn{1}{|c}{GPROF-NN 1D} &
    \multicolumn{1}{|c|}{GPROF-NN 3D} \\
    \hline\hline
    Bias \hfill [$\unit{mm\ h^{-1}}$] & \hfill $ -0.0002 \pm 0.0000$ &\hfill $ -0.0015 \pm 0.0000$ &\hfill $ -0.0005 \pm 0.0000$ \\
    MAE \hfill [$\unit{mm\ h^{-1}}$] & \hfill $  0.0210 \pm 0.0000$ &\hfill $  0.0144 \pm 0.0000$ &\hfill $  0.0116 \pm 0.0000$ \\
    MSE \hfill [$\unit{mm\ h^{-1}}$] & \hfill $  0.0143 \pm 0.0000$ &\hfill $  0.0102 \pm 0.0000$ &\hfill $  0.0060 \pm 0.0000$ \\
    SMAPE$_{0.001}$ \hfill [$\unit{\%}$] & \hfill $ 88.1093 \pm 0.0327$ &\hfill $ 75.4804 \pm 0.0335$ &\hfill $ 72.0101 \pm 0.0703$ \\
    Correlation & \hfill $  0.7591 $ &\hfill $  0.8346 $ &\hfill $  0.8785 $ \\
    \hline
  \end{tabular}

\end{table}
\clearpage

%% Please add \clearpage between each table and/or figure. Further guidelines on figures and tables can be found below.

\begin{table}[hbpt!]
  \centering
\caption{Like Tab.~\ref{tab:metrics_mhs} but for IWP.}
\label{tab:metrics_mhs_iwp}
\begin{tabular}{|c||p{3.5cm}|p{3.5cm}|p{3.5cm}|}
  \hline
  Metric &
  \multicolumn{1}{|c}{GPROF} &
  \multicolumn{1}{|c}{GPROF-NN 1D} &
  \multicolumn{1}{|c|}{GPROF-NN 3D} \\
  \hline\hline
  Bias \hfill [$\unit{mm\ h^{-1}}$] & \hfill $ -0.0035 \pm 0.0000$ &\hfill $ -0.0009 \pm 0.0000$ &\hfill $ -0.0008 \pm 0.0000$ \\
  MAE \hfill [$\unit{mm\ h^{-1}}$] & \hfill $  0.0222 \pm 0.0000$ &\hfill $  0.0123 \pm 0.0000$ &\hfill $  0.0100 \pm 0.0000$ \\
  MSE \hfill [$\unit{mm\ h^{-1}}$] & \hfill $  0.0137 \pm 0.0000$ &\hfill $  0.0093 \pm 0.0000$ &\hfill $  0.0060 \pm 0.0000$ \\
  SMAPE$_{0.001}$ \hfill [$\unit{\%}$] & \hfill $ 92.0949 \pm 0.0357$ &\hfill $ 74.1056 \pm 0.0362$ &\hfill $ 69.5782 \pm 0.0762$ \\
  Correlation & \hfill $  0.8372 $ &\hfill $  0.8878 $ &\hfill $  0.9129 $ \\
  \hline
\end{tabular}

\end{table}

\end{table}
\clearpage

\begin{table}[hbpt!]
  \centering
  \caption{Like Tab.~\ref{tab:metrics_mhs} but for CWP.}
  \label{tab:metrics_mhs_cwp}
  \begin{tabular}{|c||p{3.5cm}|p{3.5cm}|p{3.5cm}|}
    \hline
    Metric &
    \multicolumn{1}{|c}{GPROF} &
    \multicolumn{1}{|c}{GPROF-NN 1D} &
    \multicolumn{1}{|c|}{GPROF-NN 3D} \\
    \hline\hline
    Bias \hfill [$\unit{mm\ h^{-1}}$] & \hfill $ -0.0019 \pm 0.0000$ &\hfill $  0.0000 \pm 0.0000$ &\hfill $ -0.0004 \pm 0.0000$ \\
    MAE \hfill [$\unit{mm\ h^{-1}}$] & \hfill $  0.0268 \pm 0.0000$ &\hfill $  0.0195 \pm 0.0000$ &\hfill $  0.0149 \pm 0.0000$ \\
    MSE \hfill [$\unit{mm\ h^{-1}}$] & \hfill $  0.0027 \pm 0.0000$ &\hfill $  0.0016 \pm 0.0000$ &\hfill $  0.0011 \pm 0.0000$ \\
    SMAPE$_{0.001}$ \hfill [$\unit{\%}$] & \hfill $ 62.219 \pm 0.0130$ &\hfill $ 47.2591 \pm 0.0114$ &\hfill $ 38.3892 \pm 0.0237$ \\
    Correlation & \hfill $  0.8701 $ &\hfill $  0.9194 $ &\hfill $  0.9369 $ \\

    \hline
  \end{tabular}
\end{table}
\clearpage

\noappendix       %% use this to mark the end of the appendix section. Otherwise the figures might be numbered incorrectly (e.g. 10 instead of 1).

%% Regarding figures and tables in appendices, the following two options are possible depending on your general handling of figures and tables in the manuscript environment:

%% Option 1: If you sorted all figures and tables into the sections of the text, please also sort the appendix figures and appendix tables into the respective appendix sections.
%% They will be correctly named automatically.

%% Option 2: If you put all figures after the reference list, please insert appendix tables and figures after the normal tables and figures.
%% To rename them correctly to A1, A2, etc., please add the following commands in front of them:

\authorcontribution{
  SP has implemented the GPROF-NN algorithms, performed the data analysis and
  written the manuscript. PB has developed the GPROF 2021 retrieval. CK has
  supervised the project and provided feedback. PE has initiated the project
  and supervised it.
} %% this section is mandatory

\competinginterests{No competing interests are present.} %% this section is mandatory even if you declare that no competing interests are present

\begin{acknowledgements}

  The work of SP, PE and on this study was financially supported by the Swedish
  National Space Agency (SNSA) under grant 154/19.

  The work of PB on this study was financially supported by NASA Precipitation
  Measurement Missions grant 80NSSC19K0680.

  SP's visit to the Colorado State University during which parts of this work were
  conducted was supported by a scholarship from the Hans Werthén Foundation.

  The computations for this study were performed using several freely available
  programming languages and software packages, most prominently the Python
  language \citep{python}, the IPython computing environment \citep{ipython},
  the numpy package for numerical computing \citep{numpy}, xarray \citep{xarray}
  and satpy \citep{satpy} for the processing of satellite data, PyTorch
  \citep{pytorch} for implementing the machine learning models as well as
  matplotlib \citep{matplotlib} and cartopy \citep{cartopy} for generating
  figures.

\end{acknowledgements}

%% REFERENCES

\bibliographystyle{copernicus}
\bibliography{references}
%% Since the Copernicus LaTeX package includes the BibTeX style file copernicus.bst,
%% authors experienced with BibTeX only have to include the following two lines:
%%
%% \bibliographystyle{copernicus}
%% \bibliography{example.bib}
%%
%% URLs and DOIs can be entered in your BibTeX file as:
%%
%% URL = {http://www.xyz.org/~jones/idx_g.htm}
%% DOI = {10.5194/xyz}

%% LITERATURE CITATIONS
%%
%% command                        & example result
%% \citet{jones90}|               & Jones et al. (1990)
%% \citep{jones90}|               & (Jones et al., 1990)
%% \citep{jones90,jones93}|       & (Jones et al., 1990, 1993)
%% \citep[p.~32]{jones90}|        & (Jones et al., 1990, p.~32)
%% \citep[e.g.,][]{jones90}|      & (e.g., Jones et al., 1990)
%% \citep[e.g.,][p.~32]{jones90}| & (e.g., Jones et al., 1990, p.~32)
%% \citeauthor{jones90}|          & Jones et al.
%% \citeyear{jones90}|            & 1990

%% FIGURES

%% When figures and tables are placed at the end of the MS (article in one-column style), please add \clearpage
%% between bibliography and first table and/or figure as well as between each table and/or figure.

% The figure files should be labelled correctly with Arabic numerals (e.g. fig01.jpg, fig02.png).

%% ONE-COLUMN FIGURES

%%f
%
%
%%% TABLES
%%%
%%% The different columns must be seperated with a & command and should
%%% end with \\ to identify the column brake.
%
%%% ONE-COLUMN TABLE
%
%%t
%\begin{table}[t]
%\caption{TEXT}
%\begin{tabular}{column = lcr}
%\tophline
%
%\middlehline
%
%\bottomhline
%\end{tabular}
%\belowtable{} % Table Footnotes
%\end{table}
%
%%% TWO-COLUMN TABLE
%
%%t
%\begin{table*}[t]
%\caption{TEXT}
%\begin{tabular}{column = lcr}
%\tophline
%
%\middlehline
%
%\bottomhline
%\end{tabular}
%\belowtable{} % Table Footnotes
%\end{table*}
%
%%% LANDSCAPE TABLE
%
%%t
%\begin{sidewaystable*}[t]
%\caption{TEXT}
%\begin{tabular}{column = lcr}
%\tophline
%
%\middlehline
%
%\bottomhline
%\end{tabular}
%\belowtable{} % Table Footnotes
%\end{sidewaystable*}
%
%
%%% MATHEMATICAL EXPRESSIONS
%
%%% All papers typeset by Copernicus Publications follow the math typesetting regulations
%%% given by the IUPAC Green Book (IUPAC: Quantities, Units and Symbols in Physical Chemistry,
%%% 2nd Edn., Blackwell Science, available at: http://old.iupac.org/publications/books/gbook/green_book_2ed.pdf, 1993).
%%%
%%% Physical quantities/variables are typeset in italic font (t for time, T for Temperature)
%%% Indices which are not defined are typeset in italic font (x, y, z, a, b, c)
%%% Items/objects which are defined are typeset in roman font (Car A, Car B)
%%% Descriptions/specifications which are defined by itself are typeset in roman font (abs, rel, ref, tot, net, ice)
%%% Abbreviations from 2 letters are typeset in roman font (RH, LAI)
%%% Vectors are identified in bold italic font using \vec{x}
%%% Matrices are identified in bold roman font
%%% Multiplication signs are typeset using the LaTeX commands \times (for vector products, grids, and exponential notations) or \cdot
%%% The character * should not be applied as mutliplication sign
%
%
%%% EQUATIONS
%
%%% Single-row equation
%
%\begin{equation}
%
%\end{equation}
%
%%% Multiline equation
%
%\begin{align}
%& 3 + 5 = 8\\
%& 3 + 5 = 8\\
%& 3 + 5 = 8
%\end{align}
%
%
%%% MATRICES
%
%\begin{matrix}
%x & y & z\\
%x & y & z\\
%x & y & z\\
%\end{matrix}
%
%
%%% ALGORITHM
%
%\begin{algorithm}
%\caption{...}
%\label{a1}
%\begin{algorithmic}
%...
%\end{algorithmic}
%\end{algorithm}
%
%
%%% CHEMICAL FORMULAS AND REACTIONS
%
%%% For formulas embedded in the text, please use \chem{}
%
%%% The reaction environment creates labels including the letter R, i.e. (R1), (R2), etc.
%
%\begin{reaction}
%%% \rightarrow should be used for normal (one-way) chemical reactions
%%% \rightleftharpoons should be used for equilibria
%%% \leftrightarrow should be used for resonance structures
%\end{reaction}
%
%
%%% PHYSICAL UNITS
%%%
%%% Please use \unit{} and apply the exponential notation

\end{document}

### 2022-07-13 gprof_nn/models.py
Fixed hyperresidual network.

@@ -337,8 +337,7 @@ class HyperResidualMLP(ResidualMLP):
             if li > 1:
                 y += (1.0 / (li - 1)) * acc
             acc += x
-            return y, acc
-        return y
+        return y, acc
 
 
 class MultiHeadMLP(nn.Module):

### 2022-07-13 gprof_nn/data/l1c.py
Replace header property by simple attribute.

@@ -215,13 +215,6 @@ class L1CFile:
             )
 
     @property
-    def header(self):
-        import h5py
-        with h5py.File(self.path, "r") as data:
-            header = data.attrs["FileHeader"].decode().splitlines()
-        return header
-
-    @property
     def start_time(self):
         import h5py
         with h5py.File(self.path, "r") as input:

### 2022-07-14 gprof_nn/bin/train.py
Updated number of processes used for training.

@@ -359,6 +359,7 @@ def run_training_1d(
     dataset_factory = GPROF_NN_1D_Dataset
     normalizer_path = GPROF_NN_DATA_PATH / f"normalizer_{sensor.name.lower()}.pckl"
     normalizer = Normalizer.load(normalizer_path)
+    print(sensor.correction)
     kwargs = {
         "sensor": sensor,
         "batch_size": batch_size,
@@ -369,7 +370,7 @@ def run_training_1d(
     }
 
     training_data = DataFolder(
-        training_data, dataset_factory, kwargs=kwargs, queue_size=64, n_workers=8
+        training_data, dataset_factory, kwargs=kwargs, queue_size=1024, n_workers=8
     )
 
     if args.no_validation:
@@ -555,7 +556,7 @@ def run_training_3d(
         "augment": True,
     }
     training_data = DataFolder(
-        training_data, dataset_factory, queue_size=256, kwargs=kwargs, n_workers=4
+        training_data, dataset_factory, queue_size=256, kwargs=kwargs, n_workers=6
     )
 
     if args.no_validation:
@@ -867,7 +868,7 @@ def run_training_hr(
     kwargs = {"batch_size": batch_size, "normalizer": normalizer, "augment": True}
 
     training_data = DataFolder(
-        training_data, dataset_factory, queue_size=256, kwargs=kwargs, n_workers=4
+        training_data, dataset_factory, queue_size=256, kwargs=kwargs, n_workers=6
     )
 
 

### 2022-07-14 gprof_nn/bin/train.py
Include normalizer in package and load it during training.

@@ -13,6 +13,7 @@ from concurrent.futures import ProcessPoolExecutor
 from pathlib import Path
 
 from gprof_nn import sensors
+from gprof_nn.normalizer import get_normalizer
 import gprof_nn.logging
 from gprof_nn.retrieval import RetrievalDriver, RetrievalGradientDriver
 from gprof_nn.definitions import (
@@ -53,7 +54,18 @@ def add_parser(subparsers):
         "train",
         description=(
             """
-            Trains a GPROF-NN 1D or 3D network.
+            Trains a GPROF-NN 1D, 3D, simulator or HR retrieval network.
+
+            By default, the training will proceed as described in the GPROF-NN
+            paper and the network will be trained to retrieve all standard
+            GPROF variables. The options below can be used to modify
+            the training.
+
+            Tensorboard logs are written into the 'runs' directory in the
+            current working directory.
+
+            If the 'model_output' argument points to an exisiting model, the
+            training of this model will be continued.
             """
         ),
     )
@@ -69,28 +81,40 @@ def add_parser(subparsers):
         "sensor",
         metavar="sensor",
         type=str,
-        help="Name of the sensor for which to train the algorithm.",
+        help=(
+            "Name of the sensor for which to train the algorithm. "
+            "The name should correspond to a sensor object defined in"
+            " 'gprof_nn.sensors'."
+        )
     )
     parser.add_argument(
         "configuration",
         metavar="[ERA5/GANAL]",
         type=str,
-        help="The configuration for which the model is trained.",
+        help=(
+            "The type of ancillary data that the model is trained for."
+        ),
     )
     parser.add_argument(
         "training_data",
         metavar="training_data",
         type=str,
-        help="Path to training data.",
+        help="Path containing the training data.",
     )
     parser.add_argument(
         "validation_data",
         metavar="validation_data",
         type=str,
-        help="Path to validation data.",
+        help="Path containing the validation data.",
     )
     parser.add_argument(
-        "output", metavar="output", type=str, nargs=1, help="Where to store the model."
+        "output",
+        metavar="output",
+        type=str,
+        help="""
+        Path or file to which to write the trained model. If 'output'
+        points to an existing model, its training will be continued.
+        """
     )
 
     # Model configuration
[…]

### 2022-07-14 gprof_nn/models.py

@@ -503,7 +503,7 @@ class GPROF_NN_1D_QRNN(MRNN):
         self.sensor = sensor
         residuals = residuals.lower()
         if residuals not in RESIDUALS:
-            raise ValueError(f"'residuals' argument should be one of {RESIDUALS}.")
+            raise ValueError(f"{residuals} 'residuals' argument should be one of {RESIDUALS}.")
 
         if targets is None:
             targets = ALL_TARGETS
@@ -714,7 +714,6 @@ def adapt_normalizer(gmi_normalizer, sensor):
         if key < 15:
             if key in sensor.gmi_channels:
                 new_key = sensor.gmi_channels.index(key)
-                print(key, new_key)
                 new_stats[new_key] = gmi_normalizer.stats[key]
         else:
             d = 15 - sensor.n_chans

### 2022-07-14 gprof_nn/normalizer.py

@@ -1,0 +1,54 @@
+"""
+gprof_nn.normalizer
+===================
+
+This module provides access to normalizer objects, whose task is
+to normalize the retrieval inputs.
+"""
+from copy import deepcopy
+from pathlib import Path
+from quantnn.normalizer import Normalizer
+
+from gprof_nn import sensors
+
+
+NORMALIZER = Normalizer.load(Path(__file__).parent / "files" / "normalizer_gmi.pckl")
+
+
+def get_normalizer(sensor):
+    """
+    Return default normalizer for a given sensor.
+
+    This function derives a normalizer for a given sensor from the
+    normalizer used for GMI by extracting the statistics from the
+    corresponding GMI channels.
+
+    Args:
+        sensor: Sensor object representing the sensor for which
+            return a normalizer.
+
+    Return:
+        A new normalizer object for the given sensor.
+    """
+    if sensor == sensors.GMI:
+        return NORMALIZER
+
+    new_stats = {}
+    for index, gmi_index in enumerate(sensor.gmi_channels):
+        new_stats[index] = NORMALIZER.stats[gmi_index]
+
+    index = sensor.n_chans
+    if isinstance(sensor, sensors.CrossTrackScanner):
+        eias = sensor.viewing_geometry.get_earth_incidence_angles()
+        eia_min = eias.min()
+        eia_max = eias.max()
+        new_stats[index] = (eia_min, eia_max)
+        index += 1
+
+    for i in range(2):
+        new_stats[index + i] = NORMALIZER.stats[15 + i]
+
+    normalizer = deepcopy(NORMALIZER)
+    normalizer.stats = new_stats
+
+    return normalizer

### 2022-07-14 gprof_nn/sensors.py

@@ -18,7 +18,7 @@ GPM constellation. Most sensors exist in two or three variants:
      instance of the sensor plays a role, e.g. training of
      a retrieval model.
   3. A specific variant named using the naming scheme
-     {sensor}_{platform}_C. This sensor also applies a
+     sensor}_{platform}_C. This sensor also applies a
      quantile-matching correction to the training data.
 """
 from abc import ABC, abstractmethod, abstractproperty
@@ -191,6 +191,7 @@ class Sensor(ABC):
         self.platform = platform
         self.viewing_geometry = viewing_geometry
         self.missing_channels = None
+        self.l1c_version = ""
 
         # Bin file types
         self._bin_file_header = types.get_bin_file_header(n_chans, n_angles, kind)
@@ -999,7 +1000,7 @@ class ConstellationScanner(ConicalScanner):
             st = scene.surface_type.data[mask]
 
             if self.correction is not None:
-                tbs = self.correction(self, st, None, tcwv, tbs, augment=augment)
+                tbs = self.correction(rng, self, tbs, st, None, tcwv, augment=augment)
 
             # Add thermal noise to simulated observations.
             if augment and self.nedt is not None:
@@ -1093,7 +1094,7 @@ class ConstellationScanner(ConicalScanner):
         st = scene.surface_type.data
 
         if self.correction is not None:
-            tbs = self.correction(self, st, None, tcwv, tbs, augment=augment)
+            tbs = self.correction(rng, self, tbs, st, None, tcwv, augment=augment)
 
         if augment:
             if self.nedt is not None:
@@ -1551,7 +1552,7 @@ class CrossTrackScanner(Sensor):
             st = scene.surface_type.data[mask]
 
             if self.correction is not None:
[…]

### 2022-07-14 setup.py

@@ -36,6 +36,7 @@ setup(
             "files/mhs_era5_sensitivities.npy",
             "files/preprocessor_template.pp",
             "files/matplotlib_style.rc",
+            "files/normalizer_gmi.pckl"
         ]
     },
     python_requires=">=3.6",

### 2022-07-14 test/test_normalizer.py

@@ -1,0 +1,24 @@
+"""
+Tests for the gprof_nn.normalizer module.
+"""
+from gprof_nn import sensors
+from gprof_nn.normalizer import get_normalizer
+
+
+def test_get_normalizer():
+
+    normalizer_gmi = get_normalizer(sensors.GMI)
+    assert len(normalizer_gmi.stats) == 17
+
+    # For cross-track scanners stats should an entry for each
+    # channel, earth incidence angle, tcwv and t2m.
+    normalizer_mhs = get_normalizer(sensors.MHS)
+    assert len(normalizer_mhs.stats) == sensors.MHS.n_chans + 3
+
+    for index, gmi_index in enumerate(sensors.MHS.gmi_channels):
+        normalizer_mhs.stats[index] == normalizer_gmi.stats[gmi_index]
+
+    # For conical scanners stats should an entry for each
+    # channel, tcwv and t2m.
+    normalizer_mhs = get_normalizer(sensors.SSMI)
+    assert len(normalizer_mhs.stats) == sensors.SSMI.n_chans + 2

### 2022-07-14 gprof_nn/bin/retrieve.py
Refactored CLI to run retrieval.

@@ -39,8 +39,9 @@ def add_parser(subparsers):
 
             The input file may be a preprocessor file or a NetCDF4 file in
             the same format as the training data. This command can also be used
-            to run a simulator model on a training dataset in order to extend
-            the field of view of the simulated observations.
+            to run a GPROF-NN simulator on training files. This
+            will create new training data files with extended simulated TBs
+            and biases.
             """
         ),
     )
@@ -55,7 +56,15 @@ def add_parser(subparsers):
         metavar="input",
         type=str,
         nargs="+",
-        help="Folder or file containing the input data.",
+        help=(
+            """
+        Single input file or a folder containing input files. The input
+        must be a preprocessor file, a L1C file (in this case the
+        preprocessor will be run automatically) or a NetCDF4 training
+        data file. If a folder is given all files with suffixes
+        '.pp', '.HDF5' and '.nc' will be processed.
+        """
+        ),
     )
     parser.add_argument(
         "output",
@@ -63,9 +72,30 @@ def add_parser(subparsers):
         type=str,
         help="Folder or file to which to write the output.",
     )
-    parser.add_argument("--gradients", action="store_true")
-    parser.add_argument("--no_profiles", action="store_true")
-    parser.add_argument("--format", type=str, help="Output file format")
+    parser.add_argument(
+        "--gradients",
+        action="store_true",
+        help=(
+            """
+            If set, the gradients of the surface precipitation with
+            respect to the inputs will be included in the output.
+            """
+        ),
+    )
+    parser.add_argument(
+        "--no_profiles",
+        action="store_true",
+        help="If set, no profiles will be retrieved.",
+    )
+    parser.add_argument(
+        "--format",
+        type=str,
+        help=(
+            """
+            The output file format. Should be 'GPROF_BINARY' or 'NETCDF'.
+            """
+        ),
+    )
     parser.add_argument(
         "--sensor",
         type=str,
@@ -77,8 +107,11 @@ def add_parser(subparsers):
         "--preserve_structure",
         action="store_true",
         help=(
-            "Whether or not to preserve the spatial structure"
-            " of the 1D retrieval on training data."
+            """
+            If set and the input file is a training data file, the retrieval
+            will be performed on a spatially coherent scene such as those
+            used for the training of the GPROF-NN 3D retrieval.
+            """
         ),
     )
     parser.add_argument(
@@ -86,14 +119,25 @@ def add_parser(subparsers):
         metavar="n",
         type=int,
         default=4,
[…]

### 2022-07-14 gprof_nn/bin/__init__.py
Added help to sub-commands of CLI.

@@ -42,11 +42,11 @@ def gprof_nn():
     subparsers = parser.add_subparsers(help="Sub-commands")
 
     extract_data.add_parser(subparsers)
-    retrieve.add_parser(subparsers)
-    statistics.add_parser(subparsers)
     train.add_parser(subparsers)
-    legacy.add_parser(subparsers)
+    retrieve.add_parser(subparsers)
     run_preprocessor.add_parser(subparsers)
+    legacy.add_parser(subparsers)
+    statistics.add_parser(subparsers)
     extract_validation_data.add_parser(subparsers)
     combine_validation_data.add_parser(subparsers)
 

### 2022-07-14 gprof_nn/bin/extract_data.py

@@ -30,6 +30,7 @@ def add_parser(subparsers):
     """
     parser = subparsers.add_parser(
             'extract_data',
+            help="Generate training, validation or test data.",
             description='Extract training data from *.sim files.'
             )
     parser.add_argument(

### 2022-07-14 gprof_nn/bin/retrieve.py

@@ -33,6 +33,7 @@ def add_parser(subparsers):
     """
     parser = subparsers.add_parser(
         "retrieve",
+        help="Run a GPROF-NN retrieval.",
         description=(
             """
             Runs the GPROF-NN algorithm on a given input file.

### 2022-07-14 gprof_nn/bin/train.py

@@ -52,6 +52,7 @@ def add_parser(subparsers):
     """
     parser = subparsers.add_parser(
         "train",
+        help="Train a GPROF-NN model.",
         description=(
             """
             Trains a GPROF-NN 1D, 3D, simulator or HR retrieval network.

### 2022-07-14 gprof_nn/bin/train.py
Fix parsing of model directory.

@@ -337,7 +337,7 @@ def run(args):
         return 1
 
     # Check output path and define model name if necessary.
-    output = Path(args.output[0])
+    output = Path(args.output)
     if output.is_dir() and not output.exists():
         LOGGER.error("The output path '%s' doesn't exist.", output)
         return 1
